---
title: "HIKET --- Bayesian Calibration of Forest SOC Models"
subtitle: "Methods and pipeline documentation"
author: "Lorenzo Menichetti"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
    extra_dependencies: ["booktabs", "amsmath", "longtable", "array", "siunitx", "float"]
header-includes:
  - \usepackage{tikz}
  - \usetikzlibrary{arrows.meta, positioning, shapes.geometric}
geometry: margin=2.5cm
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

# Overview

## Research question

Finnish forest soils showed a positive trend in mineral SOC stocks across
the first two systematic measurement campaigns --- VMI8 (1985--86) and
Biosoil (2006). This accumulation is what the forest-sink assumptions were
built on, and its attribution is politically consequential: Finland reports
the soil as a genuine sink, whereas comparable northern countries do not. The
most recent campaign (Komeetta, 2024) --- still under review during HIKET ---
appears to corroborate a \emph{saturation} of this sink: the increase seen
between the first two campaigns does not continue at the same rate.

Yasso --- the model family underpinning Finland's soil carbon reporting ---
tends, when run with published parameter sets or calibrated under standard
steady-state assumptions, to predict a saturation of the soil sink that is
difficult to reconcile with the accumulation observed between the first two
campaigns. (Other models would plausibly behave similarly, though this has
not been tested here.) Because Finland's climate policy is built on the
forest sink, the prospect that this sink is saturating is consequential ---
and motivates testing whether the predicted saturation is genuine or a
calibration artefact. This raises the central question of HIKET:

\begin{quote}
\emph{Do divergent model projections of Finnish forest SOC --- including
apparent saturation --- reflect genuine structural differences between
models, or artefacts of how the models are calibrated?}
\end{quote}

HIKET addresses this by calibrating six models against the same Finnish
data under identical likelihood, error, and filtering assumptions, the same
litter-input estimates and climate forcing, and --- crucially --- the same
initialisation protocol. (Yasso20 differs slightly in how the climate forcing
is integrated intra-annually, but the functional form is unchanged; it simply
has the freedom to apply separate parameters to each pool group.) Residual
differences among models are then attributable to structure rather than to
parametric choices.

A key innovation is the \emph{transient initialisation} approach
(§\ref{transient}). The analytical steady state is not abandoned but
\emph{pushed back in time}: instead of imposing equilibrium at the start of
the observation period (1985), each model is initialised at its analytical
steady state in 1917 and then run transiently forward to 1985. Over this
68-year pre-run the imprint of the 1917 equilibrium assumption largely
dissipates --- the fast pools fully relax and only the slow humus pool retains
appreciable memory of it --- so the calibration begins from a dynamically
reconstructed, physically grounded 1985 state rather than an equilibrium one.
The strength of the historical anchor is itself a Bayesian-calibrated
parameter ($\sigma_{\mathrm{init}}$), so the steady-state starting point
carries a calibrated uncertainty rather than being fixed. This change
substantially improves each model's ability to track the observed SOC
trajectory.

## Models

Six models span a complexity ladder crossed with two climate-integration
choices:

| Model    | Pools | Free params | Climate $\xi$                     | Notes                          |
|----------|------:|------------:|-----------------------------------|--------------------------------|
| SP1      | 1     | 6           | single triplet, Yasso07 form      | Pure R                         |
| TP2      | 2     | 8           | single triplet, Yasso07 form      | ICBM topology, pure R          |
| TP3      | 3     | 10          | single triplet, Yasso07 form      | ASH cascade, pure R            |
| Yasso07  | 5     | 20          | single triplet                    | Fortran core                   |
| Yasso15  | 5     | 26          | pool-group-specific               | Fortran core (yasso15.so)      |
| Yasso20  | 5     | 26          | pool-group-specific, monthly $T$  | Same Fortran as Yasso15        |

Two comparison axes run through this design. First, a \emph{pool
complexity} axis: SP1 $\to$ TP2 $\to$ TP3 $\to$ Yasso07 increases the
number of pools from 1 to 5 while maintaining the same Yasso07 climate
response, isolating the effect of structural complexity on the
calibrated SOC trajectory. Second, a \emph{climate-integration} axis:
Yasso07 $\to$ Yasso15 $\to$ Yasso20 adds pool-group-specific climate
responses and finer intra-annual temperature integration, isolating
the effect of Yasso-specific climate-coupling refinements.

All models share the same AWEN litter fractionation, observation error
model, likelihood form, and transient initialisation protocol. Residual
differences in calibrated trajectories are therefore structural.

## Pipeline

The pipeline runs in four sequential stages. Data preparation is
documented separately in `Data_work.R`.

| Script                                  | Stage       | Purpose                                          |
|-----------------------------------------|-------------|--------------------------------------------------|
| `run_{Model}_baseline.R`                | 0 — Baseline | Forward run at published defaults; no calibration |
| `run_{Model}_transient_calibration.R`  | 1 — Calibration | MCMC posterior sampling                      |
| `run_{Model}_transient_predictive.R`   | 2 — Predictive  | Posterior predictive trajectories, residuals |
| `run_residual_analysis.R`              | 3 — Residuals   | Random-forest residual analysis              |
| `run_multimodel_comparison.R`          | 4 — Comparison  | Cross-model metrics and overlay figures      |

Shared MCMC settings are defined in `calib_config.R`. Prior
specifications live in `Prior_specs/{Model}_priors.R`, sourced at the
top of each calibration script.

---

# Data

A brief data description is given here; the complete treatment is
documented in `Data_work.R`.

## Plots and SOC observations

The calibration dataset covers approximately 350--400 permanent Finnish
NFI plots classified as calibration-ready
(`calib_ready = TRUE` in `site_raw.csv`). Plots are excluded if they
have zero or missing litter inputs, absent organic-layer measurements
(OFH), or implied mean residence times exceeding 100 years.

Three SOC measurement campaigns provide the calibration targets:

| Campaign  | Year(s)   | Notes                                        |
|-----------|-----------|----------------------------------------------|
| VMI8      | 1985--86  | First national campaign; mineral SOC depth   |
| Biosoil   | 2006      | European network; same depth protocol        |
| Komeetta  | 2024      | Third Finnish campaign; confirms saturation  |

With three temporally separated observations per plot, the calibration
constrains both the 1985--2006 accumulation phase and the subsequent
2006--2024 near-stability. This separation is important: models that
happen to fit the first interval but diverge over the second are
penalised.

## Litter inputs

Annual AWEN-fractionated litter inputs (tC ha$^{-1}$ yr$^{-1}$) come
from the plot-level dataset of Tupek et al.
(Zenodo, DOI: 10.5281/zenodo.19736499). Across the plot population the mean
input is not static: it rises by roughly 40\% from the mid-1980s to a peak in
the mid-2000s and then plateaus (Figure \ref{fig:drivers}c), with substantial
between-plot heterogeneity --- growing stands rise as they accumulate biomass,
while harvested stands fall. This input trajectory is an important constraint
on the initialisation problem, discussed in §\ref{transient}.

## Climate

Gridded daily climate data (temperature, precipitation) span 1961--2025
(`nfi_plot_weather_data_1961_2025.nc`). Annual or monthly summaries
are computed per plot according to each model's climate-integration
requirements.

---

# Models

All six models follow first-order kinetics: each pool $C_i$ decomposes
at a rate proportional to its size, modulated by a dimensionless climate
modifier $\xi \geq 0$. Litter input $J$ is AWEN-fractionated (acid-,
water-, ethanol-, non-soluble) and the fractionation is fixed from the
input data.

## SP1 (single pool)\label{models-sp1}

All litter inputs --- NWL, FWL, and CWL summed across AWEN fractions
--- enter a single pool $C$:

$$\frac{dC}{dt} = J(t) - \alpha\, \xi(t)\, C(t).$$

The annual step has a closed-form solution (piecewise-constant $J$ and
$\xi$ within each year):

$$C(t+1) = C_{\mathrm{ss}}(t) + \bigl(C(t) - C_{\mathrm{ss}}(t)\bigr)
  \,e^{-\alpha \xi(t)}, \qquad C_{\mathrm{ss}}(t) = \frac{J(t)}{\alpha\, \xi(t)}.$$

\begin{center}
\begin{tikzpicture}[
  box/.style = {rectangle, rounded corners=3pt, draw=black!80, thick,
                minimum width=1.05cm, minimum height=0.85cm,
                font=\sffamily\bfseries, fill=white},
  inp/.style = {-{Stealth[length=5pt]}, thick, blue!65!black},
  rsp/.style = {-{Stealth[length=4pt]}, thick, red!55!black, dotted},
  lbl/.style = {font=\scriptsize\sffamily, fill=white, inner sep=1pt}]
  \node[box] (C) at (0,0) {C};
  \draw[inp] (-1.9,0) -- (C) node[midway, above, lbl] {$J$};
  \draw[rsp] (C.south) -- +(0,-0.85)
        node[below, font=\scriptsize\sffamily] {CO$_2$\,($\alpha\,\xi$)};
\end{tikzpicture}\\[4pt]
{\footnotesize Litter input (blue); respiration (dotted). A single pool
decaying at rate $\alpha\,\xi$.}
\end{center}

Pure R; no Fortran.

## TP2 (two pools)

Two pools, $A$ (active) and $H$ (humus). All litter enters $A$; $H$
receives transferred mass from $A$ only:

$$\frac{dA}{dt} = J(t) - \alpha_A \xi(t) A(t),
  \qquad \frac{dH}{dt} = p_H \alpha_A \xi(t) A(t) - \alpha_H \xi(t) H(t).$$

\begin{center}
\begin{tikzpicture}[
  box/.style = {rectangle, rounded corners=3pt, draw=black!80, thick,
                minimum width=1.05cm, minimum height=0.85cm,
                font=\sffamily\bfseries, fill=white},
  hbox/.style= {box, fill=gray!15},
  inp/.style = {-{Stealth[length=5pt]}, thick, blue!65!black},
  tr/.style  = {-{Stealth[length=4pt]}, thick, black!75},
  rsp/.style = {-{Stealth[length=4pt]}, thick, red!55!black, dotted},
  lbl/.style = {font=\scriptsize\sffamily, fill=white, inner sep=1pt}]
  \node[box]  (A) at (0,0)   {A};
  \node[hbox] (H) at (3.2,0) {H};
  \draw[inp] (-1.9,0) -- (A) node[midway, above, lbl] {$J$};
  \draw[tr]  (A) -- (H) node[midway, above, lbl] {$p_H$};
  \draw[rsp] (A.south) -- +(0,-0.85)
        node[below, font=\scriptsize\sffamily] {CO$_2$\,($\alpha_A\,\xi$)};
  \draw[rsp] (H.south) -- +(0,-0.85)
        node[below, font=\scriptsize\sffamily] {CO$_2$\,($\alpha_H\,\xi$)};
\end{tikzpicture}\\[4pt]
{\footnotesize Litter input (blue); humifying transfer $A\to H$ (solid,
fraction $p_H$); respiration (dotted). $\xi$ modulates both decay rates.}
\end{center}

This is the ICBM topology (Andrén \& Kätterer 1997) with $\xi$
multiplying both pool decay rates. Parameter naming follows Yasso07,
making the TP2 parameter set a strict subset of Yasso07. The annual step
is solved exactly (see §\ref{models-sp1}), not Euler-approximated. Pure R.

## TP3 (three pools)\label{models-tp3}

Three pools in a strict sequential cascade: Active ($A$) $\to$ Slow
($S$) $\to$ Humus ($H$). All litter enters $A$; a fraction $p_S$ of $A$
decomposition transfers to $S$; a fraction $p_H$ of $S$ decomposition
transfers to $H$. $H$ has no climate modifier (consistent with TP2 and
Yasso conventions). The dynamics are a first-order linear cascade, with $J$ and
$\xi$ held constant within each annual step:

$$\frac{dA}{dt} = J - \alpha_A\,\xi\,A, \qquad
  \frac{dS}{dt} = p_S\,\alpha_A\,\xi\,A - \alpha_S\,\xi\,S, \qquad
  \frac{dH}{dt} = p_H\,\alpha_S\,\xi\,S - \alpha_H\,H.$$

\begin{center}
\begin{tikzpicture}[
  box/.style = {rectangle, rounded corners=3pt, draw=black!80, thick,
                minimum width=1.05cm, minimum height=0.85cm,
                font=\sffamily\bfseries, fill=white},
  hbox/.style= {box, fill=gray!15},
  inp/.style = {-{Stealth[length=5pt]}, thick, blue!65!black},
  tr/.style  = {-{Stealth[length=4pt]}, thick, black!75},
  rsp/.style = {-{Stealth[length=4pt]}, thick, red!55!black, dotted},
  lbl/.style = {font=\scriptsize\sffamily, fill=white, inner sep=1pt}]
  \node[box]  (A) at (0,0)   {A};
  \node[box]  (S) at (3.0,0) {S};
  \node[hbox] (H) at (6.0,0) {H};
  \draw[inp] (-1.9,0) -- (A) node[midway, above, lbl] {$J$};
  \draw[tr]  (A) -- (S) node[midway, above, lbl] {$p_S$};
  \draw[tr]  (S) -- (H) node[midway, above, lbl] {$p_H$};
  \draw[rsp] (A.south) -- +(0,-0.85)
        node[below, font=\scriptsize\sffamily] {CO$_2$\,($\alpha_A\,\xi$)};
  \draw[rsp] (S.south) -- +(0,-0.85)
        node[below, font=\scriptsize\sffamily] {CO$_2$\,($\alpha_S\,\xi$)};
  \draw[rsp] (H.south) -- +(0,-0.85)
        node[below, font=\scriptsize\sffamily] {CO$_2$\,($\alpha_H$)};
\end{tikzpicture}\\[4pt]
{\footnotesize Sequential cascade $A\to S\to H$ (solid, fractions $p_S,p_H$);
litter input (blue); respiration (dotted). $\xi$ modulates $A$ and $S$ decay;
$H$ has no climate modifier.}
\end{center}

Each annual step is integrated \emph{exactly}, via the matrix exponential of
the cascade generator evaluated in closed form (divided differences of its
eigenvalues $-\alpha_A\xi,\,-\alpha_S\xi,\,-\alpha_H$). This is the same
analytical treatment SP1 and TP2 apply to their pools, and that the Yasso family
applies through its matrix-exponential solver: every model in the ensemble
integrates its linear system exactly, so the differences between them are purely
structural. (An earlier implementation used an explicit annual Euler step.
Being the only approximate integrator in the ensemble, it \emph{inflated} TP3's
interannual oscillation with numerical overshoot whenever the calibrated active
pool turned over faster than the one-year step; the exact step above removes that
artefact, though a smaller physical oscillation genuinely remains --- see
§\ref{sec:tp3osc}.) Analytical steady states exist and start the transient
pre-run:

$$A^* = \frac{J}{\alpha_A\,\xi}, \qquad
  S^* = \frac{p_S J}{\alpha_S\,\xi}, \qquad
  H^* = \frac{p_H p_S J}{\alpha_H}.$$

TP3 sits between TP2 and Yasso07 on the complexity ladder, adding a
slow intermediate pool but retaining a sequential (no back-transfer)
topology. Pure R.

## The Yasso model family: shared structure

Yasso07, Yasso15, and Yasso20 share the same five-pool AWEN+H topology
and the same Fortran matrix-exponential solver. They differ only in how
the climate modifier $\xi$ is computed (§\ref{models-yasso07}--\ref{models-yasso20}).
This subsection describes the shared architecture for reference.

\subsubsection*{Pool structure and notation}

The five pools are: \textbf{A} (acid-hydrolysable), \textbf{W}
(water-soluble), \textbf{E} (ethanol-soluble), \textbf{N} (non-soluble,
largely lignin), and \textbf{H} (humus). Litter enters the four AWEN
pools in proportions fixed by the chemical fractionation of the input
data. Carbon flows between AWEN pools at rates $p_{ij} \cdot \alpha_j \cdot \xi \cdot C_j$,
where $p_{ij}$ is the fraction of pool $j$'s decomposition flux
routed to pool $i$. A fixed fraction $p_H$ of each AWEN pool's
decomposition flux forms humus; the remainder is respired as CO$_2$.
Humus decomposes irreversibly at rate $\alpha_H$ (no climate modifier
in Yasso07; pool-group-specific $\xi_H$ in Yasso15/20).

\subsubsection*{Pool diagram}

The diagram below shows all pools, litter inputs, inter-pool transfers
(labelled with HIKET parameter names), humification, and respiration
pathways. Litter inputs $J_A, J_W, J_E, J_N$ are AWEN fractions of the
annual plot-level litter; for FWL and CWL cohorts a woody-size modifier
$h_S(d)$ additionally scales $\alpha_j$.

\begin{center}
\begin{tikzpicture}[
  box/.style  = {rectangle, rounded corners=4pt, draw=black!80, thick,
                 minimum width=1.15cm, minimum height=0.85cm,
                 font=\sffamily\bfseries\small, fill=white},
  hbox/.style = {box, fill=gray!15},
  inp/.style  = {-{Stealth[length=5pt]}, thick, blue!65!black},
  tr/.style   = {-{Stealth[length=4pt]}, thick, black!75},
  hum/.style  = {-{Stealth[length=4pt]}, thick, gray!70, dashed},
  rsp/.style  = {-{Stealth[length=4pt]}, thick, red!55!black, dotted},
  lbl/.style  = {font=\tiny\sffamily, fill=white, inner sep=0.8pt,
                 text=black!80}
]

%% --- Pool nodes ---
\node[box]  (A) at (0.0, 3.6) {A};
\node[box]  (W) at (4.2, 3.6) {W};
\node[box]  (E) at (0.0, 0.0) {E};
\node[box]  (N) at (4.2, 0.0) {N};
\node[hbox] (H) at (2.1,-2.4) {H};

%% --- Litter inputs ---
\draw[inp] (-1.5, 3.6) -- (A) node[midway, above, lbl] {$J_A$};
\draw[inp] (2.1,  5.3) -- (W) node[midway, right, lbl] {$J_W$};
\draw[inp] (-1.5, 0.0) -- (E) node[midway, above, lbl] {$J_E$};
\draw[inp] (5.7,  0.0) -- (N) node[midway, above, lbl] {$J_N$};

%% --- A <-> W (top horizontal) ---
\draw[tr, bend left=13]  (A) to node[lbl, above] {$p_{AW}$} (W);
\draw[tr, bend left=13]  (W) to node[lbl, below] {$p_{WA}$} (A);

%% --- E <-> N (bottom horizontal) ---
\draw[tr, bend right=13] (E) to node[lbl, below] {$p_{EN}$} (N);
\draw[tr, bend right=13] (N) to node[lbl, above] {$p_{NE}$} (E);

%% --- A <-> E (left vertical) ---
\draw[tr, bend right=13] (A) to node[lbl, left]  {$p_{AE}$} (E);
\draw[tr, bend right=13] (E) to node[lbl, right] {$p_{EA}$} (A);

%% --- W <-> N (right vertical) ---
\draw[tr, bend left=13]  (W) to node[lbl, right] {$p_{WN}$} (N);
\draw[tr, bend left=13]  (N) to node[lbl, left]  {$p_{NW}$} (W);

%% --- A <-> N (diagonal) ---
\draw[tr, bend left=7]  (A) to node[lbl, above right, yshift=2pt] {$p_{AN}$} (N);
\draw[tr, bend left=7]  (N) to node[lbl, below left,  yshift=-2pt]{$p_{NA}$} (A);

%% --- W <-> E (diagonal) ---
\draw[tr, bend right=7] (W) to node[lbl, above left,  yshift=2pt] {$p_{WE}$} (E);
\draw[tr, bend right=7] (E) to node[lbl, below right, yshift=-2pt]{$p_{EW}$} (W);

%% --- Humification: AWEN -> H ---
\draw[hum] (A.south) to[out=-80, in=145]
           node[lbl, left, pos=0.35] {$p_H$} (H.north west);
\draw[hum] (W.south) to[out=-100,in=35]
           node[lbl, right,pos=0.35] {$p_H$} (H.north east);
\draw[hum] (E.east)  to[out=-20, in=175]
           node[lbl, below,pos=0.5] {$p_H$} (H.west);
\draw[hum] (N.west)  to[out=200, in=5]
           node[lbl, below,pos=0.5] {$p_H$} (H.east);

%% --- H respiration ---
\draw[rsp] (H.south) -- +(0,-0.9)
           node[below, font=\tiny\sffamily] {CO$_2$\;($\alpha_H$)};

%% --- AWEN respiration ---
\draw[rsp] (A.north west) -- +(-0.7, 0.7)
           node[above left, font=\tiny\sffamily] {CO$_2$};
\draw[rsp] (W.north east) -- +( 0.7, 0.7)
           node[above right,font=\tiny\sffamily] {CO$_2$};
\draw[rsp] (E.south west) -- +(-0.7,-0.7)
           node[below left, font=\tiny\sffamily] {CO$_2$};
\draw[rsp] (N.south east) -- +( 0.7,-0.7)
           node[below right,font=\tiny\sffamily] {CO$_2$};

%% --- xi annotation ---
\node[font=\scriptsize\sffamily, align=center, text=black!60]
  at (2.1, 4.8)
  {$\xi$: climate modifier on $\alpha_{A,W,E,N}$
   (and $\alpha_H$ in Yasso15/20)};

\end{tikzpicture}
\end{center}

\vspace{0.3em}
\noindent
\textbf{Arrow key:}
\tikz[baseline=-0.5ex]{\draw[-{Stealth[length=4pt]},thick,black!75](0,0)--(0.6,0);}~inter-pool transfer ($p_{ij}$)\quad
\tikz[baseline=-0.5ex]{\draw[-{Stealth[length=4pt]},thick,gray!70,dashed](0,0)--(0.6,0);}~humification ($p_H$)\quad
\tikz[baseline=-0.5ex]{\draw[-{Stealth[length=4pt]},thick,red!55!black,dotted](0,0)--(0.6,0);}~respiration (CO$_2$)\quad
\tikz[baseline=-0.5ex]{\draw[-{Stealth[length=4pt]},thick,blue!65!black](0,0)--(0.6,0);}~litter input ($J$)

\subsubsection*{Parameter summary}

\begin{small}
\begin{longtable}{l l l l}
\toprule
Parameter & Symbol & Role & Models \\
\midrule
Decomp.\ rates (AWEN) & $\alpha_A, \alpha_W, \alpha_E, \alpha_N$ &
  Base first-order rate of each AWEN pool (yr$^{-1}$); fixed at published values & Y07/15/20 \\
Decomp.\ rate (H) & $\alpha_H$ &
  Humus decomposition rate (yr$^{-1}$); fixed & Y07/15/20 \\
Transfer fractions & $p_{ij}$, $i \neq j \in \{A,W,E,N\}$ &
  Fraction of pool $j$ decomp.\ flux routed to pool $i$; 12 per model, all free & Y07/15/20 \\
Humification & $p_H$ &
  Fraction of each AWEN decomp.\ flux routed to H; same for all four pools; fixed & Y07/15/20 \\
AWE climate, linear & $\beta_1$ &
  Linear temperature coefficient in $\xi_\mathrm{AWE}$; free, log-transformed & Y07/15/20 \\
AWE climate, quadratic & $\beta_2$ &
  Quadratic temperature coefficient; free, unconstrained & Y07/15/20 \\
AWE precipitation & $\gamma$ &
  Precipitation saturation exponent in $\xi_\mathrm{AWE}$; free, unconstrained & Y07/15/20 \\
N-pool climate & $\beta_{N1}, \beta_{N2}, \gamma_N$ &
  As above, for the N pool separately & Y15/20 \\
H-pool climate & $\beta_{H1}, \beta_{H2}, \gamma_H$ &
  As above, for the H pool separately & Y15/20 \\
Size modifier & $\delta_1, \delta_2, r$ &
  Reduce $\alpha$ for woody cohorts: $h_S(d) = \min(1,(1+\delta_1 d + \delta_2 d^2)^{-|r|})$; free & Y07/15/20 \\
Leaching & $w_1,\ldots,w_5$ &
  Leaching loss weights; inactive (leac\,=\,0) throughout; fixed & Y15/20 \\
Historical litter & $\sigma_\mathrm{init}$ &
  Scales 1917 litter anchor in the pre-run; encodes pre-1985 productivity history; free & all \\
Contemporary litter & $\sigma_\mathrm{input}$ &
  Scales observed litter inputs (1985 onward); absorbs litter model uncertainty; free & all \\
\bottomrule
\end{longtable}
\end{small}
\vspace{-0.5em}
{\footnotesize Y07 = Yasso07; Y15 = Yasso15; Y20 = Yasso20; Y15/20 = both.}

## Yasso07\label{models-yasso07}

Five pools (Tuomi et al. 2009, 2011): A (acid-soluble), W
(water-soluble), E (ethanol-soluble), N (non-soluble / lignin), H
(humus). Litter inputs enter A, W, E, N proportionally to AWEN
fractions; transfers between AWEN pools are controlled by 12 fractions
$p_{ij}$; all four AWEN pools contribute to H at a humification rate
$p_H$; H decomposes irreversibly at rate $\alpha_H$.

The dynamics are $dC/dt = AC + b(t)$, with the matrix $A$ assembled
from decomposition rates and transfer fractions. The Fortran core
solves each annual step analytically via the matrix exponential
$C(t+1) = A^{-1}(e^{A\Delta t}(AC(t) + b) - b)$, with a Euler fallback
for near-singular matrices. NWL, FWL, and CWL litter types are tracked
as three parallel cohorts; the per-cohort state vector has 15 elements
(5 pools $\times$ 3 cohorts).

The climate modifier uses a four-point Gaussian approximation of the
intra-annual temperature cycle:

$$\xi(\bar T, T_{\mathrm{amp}}, P) =
  \frac{1}{4}\sum_{k=1}^{4} \exp\!\bigl(\beta_1 T_k + \beta_2 T_k^2\bigr)
  \times \bigl(1 - \exp(\gamma P/1000)\bigr),$$

where $T_k$ are four quarterly temperatures derived from annual mean
$\bar T$ and seasonal amplitude $T_{\mathrm{amp}}$. A woody-size modifier
scales decomposition for FWL/CWL cohorts:

$$h_S(d) = \min\!\Bigl\{1,\; (1 + \delta_1 d + \delta_2 d^2)^{-|r|}\Bigr\},$$

where $d$ is the cohort diameter (cm; fixed at 2 cm for FWL,
15 cm for CWL).

## Yasso15

Identical pool topology, AWEN partitioning, and matrix-exponential
machinery as Yasso07. The structural change is \emph{pool-group-specific}
climate response: three triplets $(\beta_1, \beta_2, \gamma)$,
$(\beta_{N1}, \beta_{N2}, \gamma_N)$, $(\beta_{H1}, \beta_{H2},
\gamma_H)$ act on the AWE, N, and H pool groups respectively. The
functional form matches Yasso07 (four-point Gaussian temperature
approximation, exponential-with-saturation precipitation response).
Leaching parameters $w_1, \ldots, w_5$ are present in the 35-element
parameter vector but are inactive throughout (leaching scalar held at
zero).

## Yasso20\label{models-yasso20}

Same Fortran core as Yasso15 (compiled as `yasso15.so`, reused). The
single structural change is the intra-annual temperature integration:
instead of the four-point Gaussian approximation, the modifier is
averaged directly over all 12 monthly temperatures:

$$\xi_T^{(g)} = \frac{1}{12}\sum_{m=1}^{12}
  \exp\!\bigl(\beta_1^{(g)} T_m + \beta_2^{(g)} T_m^2\bigr),
  \qquad g \in \{\mathrm{AWE}, N, H\}.$$

This is the only computational difference from Yasso15. Monthly climate
data (12 rows per year per plot) are required, compared to annual
summaries for all other models.

Yasso20 inherits the same 35-parameter structure as Yasso15 in this
calibration (all 12 inter-pool flow fractions free; see §\ref{param-yasso20}).

\medskip\noindent\textbf{Note on the published Yasso20 parameterisation.}
The MAP parameter values from the original global calibration (Viskari
et al. 2022) include six transfer fractions constrained to zero. When
these published values are applied directly, the per-pool mass budget
for pools A and N is marginally exceeded (transfer fractions plus
humification fraction sum slightly above 1.0), producing a numerically
unstable forward simulation in the HIKET Fortran implementation, which
enforces strict diagonal dominance. The HIKET calibration avoids this
region of parameter space entirely through the stick-breaking budget
constraint (§\ref{stick-breaking}). The Yasso20 uncalibrated baseline
uses FMI prior-centre values, which satisfy mass conservation.

---

# Fortran implementation: differences from the FMI reference code\label{sec:fortran-diffs}

The HIKET Fortran cores (`yasso07.f90`, `yasso15.f90`) derive from the
FMI Yasso implementations (Järvenpää 2015 for Yasso15/20; Tuomi et al.
for Yasso07) but incorporate several modifications introduced in 2025.
These are documented here for transparency and to assist cross-comparison
with the original FMI code.

## Diagonal dominance check and Euler fallback

The annual timestep in each model is solved via the matrix exponential
$\exp(A \Delta t)$. The original FMI code applies the matrix exponential
unconditionally. The HIKET implementation adds a pre-check for diagonal
dominance: for each column $j$ of $A$, if the sum of off-diagonal absolute
values (outflows to other pools and to humus) exceeds 99.99\% of the diagonal
absolute value (total decomposition rate), the matrix is classified as
near-singular and a first-order Euler step is used instead:

$$C(t+\Delta t) \approx \max\bigl(0,\; C(t) + (A\,C(t) + b)\,\Delta t\bigr).$$

The condition being detected is mathematically equivalent to the per-pool
mass budget $\sum_i p_{ij} + p_H \gtrsim 1$: the sum of all transfer
fractions and the humification fraction approaches or exceeds 1, leaving
zero or negative respiration from pool $j$. The $\max(0, \cdot)$ clamp
prevents negative pool values in the Euler path but does not conserve mass
exactly; repeated Euler steps with clamping can cause slow carbon drift
(see §\ref{models-yasso20} for the Yasso20 context). The HIKET calibration
avoids this regime entirely via the stick-breaking budget constraint.

## Matrix exponential: scaling cap and Taylor terms

The matrix exponential is computed by scaling-and-squaring with a Taylor
series expansion. Two changes were made to the original implementation:

* **Scaling-step cap.** The original code had no upper bound on the number
  of scaling steps $j$ (chosen so that $\|A\| < 2^j$). For ill-conditioned
  matrices with very large norms, $j$ grew without bound and the squaring
  loop ran for thousands of iterations, effectively hanging. The cap
  $j \leq 64$ (covering $\|A\| \leq 2^{64} \approx 1.8 \times 10^{19}$)
  terminates the scaling phase for any physically meaningful matrix.

* **Taylor terms.** The number of Taylor series terms was increased from 10
  to 20. When heavy scaling is applied (large $j$), the scaled matrix
  $A / 2^j$ is close to zero and many terms are needed for the expansion to
  accumulate the correction accurately.

## Terminal per-cohort state output (\texttt{C\_final})

The Fortran run routines (`yasso07_run_r`, `yasso15_run_r`) return a new
output argument `C_final(15)`: the terminal per-cohort pool state
$[C_\mathrm{nwl} \mid C_\mathrm{fwl} \mid C_\mathrm{cwl}]$ at the end of the
last simulated year. This argument is absent from the original FMI code,
which returns only the five summed pool totals.

`C_final` is required for two purposes in HIKET. First, the transient
pre-run chains directly into the calibration period: the pre-run's
terminal state becomes $C_\mathrm{init}$ for 1985 (§\ref{transient}).
Without the full 15-element per-cohort breakdown, re-initialising the
forward simulation would require an analytical steady-state approximation
that introduces additional error. Second, the 60-year forward projection
(predictive stage) chains from the end of the historical run in the same
way. The `C_final` field is attached to the R-side output data frame via
`attr(run_out, "C_final")`.

## $\xi$ precomputation in R

The original FMI code computes the climate modifier $\xi$ internally
within the Fortran timestep loop, from raw temperature and precipitation
inputs. HIKET separates this computation: $\xi$ is computed as an annual
(or monthly-aggregated) array in R \emph{before} the Fortran call, and
passed to the Fortran as a pre-computed vector argument. This separation
has two benefits. First, it allows the Jensen's inequality correction
(§\ref{transient}) to be applied consistently and transparently at the R
level. Second, it makes the $\xi$ functional form easily modifiable
without recompiling Fortran; the different forms for Yasso07 (four-point
Gaussian), Yasso15 (same form, pool-group-specific), and Yasso20
(12-monthly average) are handled in dedicated R functions
(`compute_xi_yasso07`, `compute_xi_yasso15`, `compute_xi_yasso20`).

## Floating-point precision (Yasso15 and Yasso20)

In `yasso15_mod`, the precision kind is defined as
`dp = SELECTED_REAL_KIND(6, 37)`, which evaluates to `REAL(4)`
(single precision, $\sim$7 significant digits) on standard hardware.
Despite the `REAL(dp)` declarations throughout the source, all Fortran
arguments for Yasso15 and Yasso20 are therefore 4-byte. The R wrappers
use `as.single()` for all `.Fortran()` calls to Yasso15/20.

Yasso07 uses `dp = SELECTED_REAL_KIND(15, 307)` = genuine double
precision. Its R wrapper correctly uses `as.double()`.

An earlier version of the Yasso15 wrapper incorrectly used `as.double()`
for the Yasso15/20 calls. This caused segfaults in the Fortran (type
mismatch in the `.Fortran()` interface) that were resolved by reverting
to `as.single()`.

## Parameter ordering

The FMI `.dat` posterior files use a different parameter ordering from
HIKET. In the FMI files, leaching weights $w_1$--$w_5$ occupy positions
17--21, followed by the climate parameters; in HIKET, climate parameters
occupy positions 17--25 and leaching weights are at 31--35. The remapping
is applied in `Priors_model_matching.R` via the index vector
`ORIG_TO_LORENZO` and verified by cross-checking MAP values against
named CSV files (`y15par.csv`, `y07par_gui.csv`).

## Yasso07: sign convention for $r$

The original Yasso07 parameter files store $r$ as a negative value
(e.g., $-0.307$). The Fortran applies $-|r|$ in the size modifier
exponent, so the sign is irrelevant at runtime. HIKET stores $r$ as
positive and applies a log transform in the calibration (which requires
positivity). The `abs()` conversion is applied when reading reference
values from `y07par_gui.csv` or `YASSO07_DEFAULT_PARAMS`.

---

# Parameter specification

## Free, fixed, and derived parameters

For each model, parameters split into:

* **Free**: sampled by MCMC. Includes the two auxiliary uncertainty
  parameters $\sigma_{\mathrm{init}}$ and $\sigma_{\mathrm{input}}$
  in every model.
* **Fixed**: held at published or structurally motivated values. For
  the Yasso family, decomposition rates and humus parameters are fixed
  at published defaults, reducing free-parameter count and keeping the
  intercomparison clean.
* **Derived**: none in the current pipeline.

Transforms map constrained physical parameters to an unconstrained
sampling space, with log-Jacobian contributions added to the prior
density:

* `log`: positivity; $x = \log\theta$.
* `logit`: $(0,1)$ bounds; $x = \log(\theta/(1-\theta))$.
* `unconstrained`: identity; no constraint beyond prior.
* `stick_break`: $K$-fraction simplex constrained to sum $\leq B$;
  described in §\ref{stick-breaking}.

## SP1 (6 free, 0 fixed)

| Parameter | Meaning | Transform |
|-----------|---------|-----------|
| $\alpha$ | Decomposition rate (yr$^{-1}$) | log |
| $\beta_1$ | Temperature response, linear | log |
| $\beta_2$ | Temperature response, quadratic | unconstrained |
| $\gamma$ | Precipitation response | unconstrained |
| $\sigma_{\mathrm{init}}$ | Historical litter scaling (1917 anchor) | log |
| $\sigma_{\mathrm{input}}$ | Contemporary litter scaling | log |

## TP2 (8 free, 0 fixed)

| Parameter | Meaning | Transform |
|-----------|---------|-----------|
| $\alpha_A$ | A-pool decomposition rate (yr$^{-1}$) | log |
| $\alpha_H$ | H-pool decomposition rate (yr$^{-1}$) | log |
| $p_H$ | Humification fraction (A $\to$ H) | logit |
| $\beta_1, \beta_2, \gamma$ | Climate response | log, unc., unc. |
| $\sigma_{\mathrm{init}}, \sigma_{\mathrm{input}}$ | Litter scalings | log, log |

## TP3 (10 free, 0 fixed)

| Parameter | Meaning | Transform |
|-----------|---------|-----------|
| $\alpha_A$ | A-pool decomposition rate (yr$^{-1}$) | log |
| $\alpha_S$ | S-pool decomposition rate (yr$^{-1}$) | log |
| $\alpha_H$ | H-pool decomposition rate (yr$^{-1}$) | log |
| $p_S$ | Transfer fraction A $\to$ S | logit |
| $p_H$ | Transfer fraction S $\to$ H | logit |
| $\beta_1, \beta_2, \gamma$ | Climate response | log, unc., unc. |
| $\sigma_{\mathrm{init}}, \sigma_{\mathrm{input}}$ | Litter scalings | log, log |

No fixed parameters. Like SP1 and TP2, TP3 has no published global
calibration, so all parameters are learned from the Finnish data.

## Yasso07 (20 free, 6 fixed)

**Fixed (6):** $\alpha_A, \alpha_W, \alpha_E, \alpha_N$ (AWEN
decomposition rates), $p_H$, $\alpha_H$. Published Tuomi et al. defaults.

**Free (18):**

| Parameter | Count | Transform |
|-----------|------:|-----------|
| 12 transfer fractions $p_{ij}$, 4 columns of 3 | 12 | stick\_break |
| $\beta_1$, $\beta_2$, $\gamma$ | 3 | log, unc., unc. |
| $\delta_1$, $\delta_2$, $r$ | 3 | unc., log, log |
| $\sigma_{\mathrm{init}}, \sigma_{\mathrm{input}}$ | 2 | log, log |

$\delta_2 > 0$ enforced (Fortran NaN for $\delta_2 < 0$ at $d > 0.4$
cm). $r$ stored as positive in HIKET; reference defaults store it
negative.

## Yasso15 (26 free, 11 fixed)

**Fixed (11):** $\alpha_A, \alpha_W, \alpha_E, \alpha_N$, $p_H$,
$\alpha_H$, $w_1, \ldots, w_5$ (leaching weights; inactive at leac = 0).
Published Yasso15 defaults.

**Free (28):**

| Parameter | Count | Transform |
|-----------|------:|-----------|
| 12 transfer fractions $p_{ij}$ | 12 | stick\_break |
| AWE climate: $\beta_1, \beta_2, \gamma$ | 3 | log, unc., unc. |
| N climate: $\beta_{N1}, \beta_{N2}, \gamma_N$ | 3 | log, unc., unc. |
| H climate: $\beta_{H1}, \beta_{H2}, \gamma_H$ | 3 | log, unc., unc. |
| Size modifier: $\delta_1, \delta_2, r$ | 3 | unc., log, log |
| $\sigma_{\mathrm{init}}, \sigma_{\mathrm{input}}$ | 2 | log, log |

## Yasso20 (26 free, 11 fixed)\label{param-yasso20}

Identical fixed/free structure to Yasso15. The same 35-parameter vector
is used; only the $\xi$ computation method differs at runtime.

\medskip\noindent\textbf{Uniform 12-fraction structure across the Yasso
family.} Earlier Yasso20 calibrations (Viskari 2022) fixed six transfer
fractions at zero (structural zeros from a global-dataset MCMC) and
derived three algebraically. HIKET does not impose these constraints.
All 12 inter-pool fractions are sampled in all three Yasso models,
for two reasons. First, the Viskari zeros reflect a warm-site global
calibration and are not general biological laws; Finnish boreal data
should be free to inform pool-flow structure. Second, fewer free
fractions in Yasso20 than in Yasso07 or Yasso15 would confound
structural (climate-integration) differences with parametric (degrees
of freedom) differences, making the intercomparison uninterpretable.
Prior centres for fractions use the published Yasso15 defaults (see §4).

## Stick-breaking parameterisation for transfer fractions\label{stick-breaking}

Each column $j$ of the AWEN transfer matrix has three outflow fractions
satisfying $\sum_i p_{ij} + p_H \leq 1$ (mass conservation; the unallocated
remainder $1 - \sum_i p_{ij} - p_H$ is respired as CO$_2$). The constraint
is enforced via stick-breaking with budget $B = 1 - p_H$. Three
unconstrained parameters $z_1, z_2, z_3$ map to the three fractions by
sequentially breaking pieces off the remaining budget:

$$p_1 = B \cdot \sigma(z_1), \quad
  p_2 = (B - p_1) \cdot \sigma(z_2), \quad
  p_3 = (B - p_1 - p_2) \cdot \sigma(z_3),$$

where $\sigma$ is the logistic function. Because each $\sigma(z_i) < 1$,
the sum $p_1 + p_2 + p_3$ is strictly less than $B$, guaranteeing a
strictly positive respired fraction. All three fractions in a column are
therefore free parameters --- none is algebraically determined --- and
the log-Jacobian of the transform is added to the prior. This guarantees
the budget constraint exactly and produces a well-conditioned Jacobian.
Each of the four AWEN columns uses an independent stick-breaking transform
with the same budget $B$, contributing $4 \times 3 = 12$ free parameters
for the 12 transfer fractions.

## Cross-model parameter overlap

\begin{small}
\begin{longtable}{l c c c c c c}
\toprule
Parameter & SP1 & TP2 & TP3 & Yasso07 & Yasso15 & Yasso20 \\
\midrule
$\alpha$ (single rate)             & free  & ---   & ---   & ---   & ---   & ---   \\
$\alpha_A, \alpha_H$               & ---   & free  & free  & fixed & fixed & fixed \\
$\alpha_S$                         & ---   & ---   & free  & ---   & ---   & ---   \\
$\alpha_W, \alpha_E, \alpha_N$     & ---   & ---   & ---   & fixed & fixed & fixed \\
$p_H$                              & ---   & free  & free  & fixed & fixed & fixed \\
$p_S$                              & ---   & ---   & free  & ---   & ---   & ---   \\
12 transfer fractions $p_{ij}$     & ---   & ---   & ---   & free  & free  & free  \\
$\beta_1, \beta_2, \gamma$         & free  & free  & free  & free  & free (AWE) & free (AWE) \\
$\beta_{N1}, \beta_{N2}, \gamma_N$ & ---   & ---   & ---   & ---   & free  & free  \\
$\beta_{H1}, \beta_{H2}, \gamma_H$ & ---   & ---   & ---   & ---   & free  & free  \\
$\delta_1, \delta_2, r$            & ---   & ---   & ---   & free  & free  & free  \\
$w_1, \ldots, w_5$                 & ---   & ---   & ---   & ---   & fixed & fixed \\
$\sigma_{\mathrm{init}}, \sigma_{\mathrm{input}}$ & free & free & free & free & free & free \\
\midrule
\textbf{Total free} & \textbf{6} & \textbf{8} & \textbf{10} & \textbf{20} & \textbf{26} & \textbf{26} \\
\bottomrule
\end{longtable}
\end{small}

---

# Priors

All priors are normal in unconstrained (transformed) space, centred at
published model defaults. The joint prior factorises across parameters:

$$\log p(x) = \sum_{i} \log\mathcal{N}\!\bigl(x_i;\, x_i^{\mathrm{ctr}},
  \sigma_{\mathrm{ppm},i}\bigr) + \log J(x),$$

where $x^{\mathrm{ctr}}$ is the unconstrained image of the physical-space
prior centre, $\sigma_{\mathrm{ppm}}$ is the prior standard deviation in
unconstrained space, and $\log J$ is the log-Jacobian of the transforms.
Prior specifications are stored in `Prior_specs/{Model}_priors.R` and
sourced at the top of each calibration script.

## Homogenization across models

The scientific question --- whether divergent SOC projections reflect genuine
structural differences or calibration artefacts --- requires that the
calibration setup itself inject no asymmetry that confounds structure with
calibration choice. Priors are therefore homogenized in their
\emph{construction method, scale, constraints, and degrees of freedom} ---
\textbf{not} in their raw numbers. Identical numbers would in fact be wrong,
because the parameters live on model-specific scales; forcing the same number
is precisely what produced the Yasso07 $\beta_2$ pathology. Each parameter
class is assigned to a tier by a single criterion: \emph{was it consistently
and freely calibrated across the original models?}

\begin{small}
\begin{longtable}{p{2.6cm} p{2.4cm} p{2.4cm} p{2.0cm} p{2.4cm}}
\toprule
Parameter class & Centre source & Width source & Transform & Consistently calibrated across originals? \\
\midrule
\textbf{Tier 1} climate \& size ($\beta$'s, $\gamma$'s, $\delta$'s, $r$) &
each model's own published MAP &
each model's own published posterior/limit (1$\sigma$) &
log or unconstrained &
yes --- free in every original $\Rightarrow$ per-model widths kept \\
\textbf{Tier 2} 12 transfer fractions &
each model's own published structure &
\emph{common} logit SD = 0.4 (identical) &
logit (stick-break) &
no --- Viskari 2022 fixed/derived 9 of 12 $\Rightarrow$ common prior restores symmetry \\
\textbf{Tier 3} auxiliary ($\sigma_{\mathrm{init}}$, $\sigma_{\mathrm{input}}$) &
1.0 &
\emph{common} log SD = 0.50 (identical) &
log &
HIKET-only $\Rightarrow$ identical everywhere \\
\bottomrule
\end{longtable}
\end{small}

Tier 1 keeps each model's own published uncertainty because coverage is
consistent across the original calibrations. Tier 2 overrides the published
fractions with a single common prior precisely because they were \emph{not}
consistently calibrated --- applying the Tier-1 recipe would freeze Yasso20's
nine fixed fractions (zero posterior spread) while Yasso15's stay free,
re-introducing the very asymmetry the design exists to remove. What
legitimately differs between models --- Tier-1 centres and widths, and the
parameter set itself (Yasso07 has no separate N/H climate response) --- is
genuine structure, not a calibration choice.

## Sources

\begin{small}
\begin{longtable}{l l l}
\toprule
Model & Source for prior centres & Source for prior widths \\
\midrule
SP1   & Hand-set (no published global calib.) & Climate on Yasso07 scale; rest weak \\
TP2   & Hand-set ICBM-style defaults          & Climate on Yasso07 scale; rest weak \\
TP3   & Hand-set; extends TP2 with Slow pool  & Climate on Yasso07 scale; rest weak \\
Yasso07 & \texttt{y07par\_gui.csv} (Tuomi et al. 2011) & Tuomi 2009 T3 (climate), 2011 T4 (woody) \\
Yasso15 & \texttt{Yasso15.dat} posterior mean (FMI) & \texttt{Yasso15.dat} posterior SD \\
Yasso20 & \texttt{Yasso20\_sample\_parameters.rda} (FMI) & \texttt{Yasso20.dat} posterior SD \\
\bottomrule
\end{longtable}
\end{small}

The FMI repository is at
\url{https://github.com/YASSOmodel/Ryassofortran/tree/master/data}.
The `.dat` files contain 10\,000 MCMC samples from the original
calibrations (row 1 = MAP, rows 2--10\,001 = posterior draws). Column
ordering in the FMI files differs from HIKET's parameter ordering;
remapping is performed and verified in `Priors_model_matching.R`.

For Yasso07, no posterior file is publicly available; $\sigma_{\mathrm{ppm}}$
is derived from the published posterior confidence limits --- climate
parameters from Tuomi et al. (2009) Table 3, woody-size parameters from
Tuomi et al. (2011) Table 4. These ``$\pm$'' limits are read conservatively
as $1\sigma$ (not divided by 1.96), giving broader, less-informative priors;
for log-transformed parameters the relative SD is the limit divided by the
paper MAP (delta method). For SP1, TP2, and TP3 the climate widths are placed
on this same Yasso07 empirical scale --- their climate centres were already
borrowed from the Yasso07 MAP --- while their remaining structural parameters
stay weakly informative so the Finnish data dominates.

This replaces an earlier hand-set $\sigma_{\mathrm{ppm}}(\beta_2) = 0.05$
shared by Yasso07/SP1/TP2/TP3, which was $\sim$100--360$\times$ looser than
the empirical Yasso15/20 widths. Because $\beta_2$ multiplies $T^2$
($\sim$600 at boreal seasonal extremes), that width let the climate modifier
$\xi$ span tens of orders of magnitude, producing 25--46\,\% non-finite MCMC
proposals under the all-or-nothing likelihood. The $1\sigma$ value
($6.5\times10^{-4}$) keeps the prior-predictive temperature factor bounded
(99th percentile $\approx 9$ rather than $\sim$$10^{31}$).

## Transfer-fraction priors

For all three Yasso models the prior standard deviation on every transfer
fraction is $\sigma_{\mathrm{ppm}} = 0.4$ in unconstrained (stick-breaking
logit) space, identical across models. This is the only prior class given a
literally identical width everywhere: pool-flow structure is the primary
structural-comparison axis, and no single model's published structure should
be allowed to dominate --- six of Yasso20's fractions were structural zeros in
Viskari et al. (2022), and the homogenization deliberately lets the Finnish
data, not the original warm-site constraints, decide the flow architecture.
The logit scale gives every fraction the same \emph{relative} exploration
window regardless of whether its centre is near zero or near one. The chosen
width of 0.4 keeps a small fraction (centre 0.01) within $\sim$2.2$\times$ of
its centre at $+2\sigma$, versus $\sim$7$\times$ under the previous
$\sigma_{\mathrm{ppm}} = 1.0$; this stops the chains from wandering into
forward-model-exploding regions without manufacturing convergence. The
fractions remain genuinely non-identifiable (only two SOC observations per
plot constrain $\sim$12 flows), so their $R$-hat is expected to stay poor ---
that poor identifiability is a scientific finding, not a defect to be tuned
away. Prior centres for the fractions use the published Yasso15 defaults for
all three Yasso models (Yasso20 included, since the previously-fixed zeros
are now free parameters and Yasso15 defaults are the closest available
nonzero prior centres from the same chemical-fractionation framework).

## Climate-response and size-modifier priors

Climate and size parameters receive \emph{more informative} priors than the
other parameter classes (though still relatively broad in absolute terms),
derived from the FMI posterior (Yasso15/20) or from the Tuomi 2009/2011
published posterior limits (Yasso07). This prevents two pathologies:

1. **The $\gamma$\,/\,$\sigma_{\mathrm{input}}$ compensation loop.** Both
   parameters scale total decomposition; without an informative prior on
   $\gamma$ they trade off freely, producing a ridge in the posterior that
   destabilises MCMC mixing.
2. **The $\delta_2 < 0$ NaN pathology.** The Fortran size modifier returns
   NaN for $\delta_2 < 0$ at $d > 0.4$ cm; a log transform with
   informative prior keeps $\delta_2$ in the well-behaved region.

$\gamma_H$ for Yasso15 is capped at $\sigma_{\mathrm{ppm}} = 1.5$: the
H-pool precipitation response is near-unidentifiable at Finnish
precipitation levels ($\sim 600$ mm). The large raw posterior SD ($\sim 3$)
reflects a flat likelihood surface, not genuine uncertainty.

## Numerical values

\textbf{Yasso07}

| Parameter | $x^{\mathrm{ctr}}$ (physical) | $\sigma_{\mathrm{ppm}}$ | Space |
|-----------|------------------------------:|------------------------:|-------|
| $p_{ij}$ (12) | published defaults | 0.4 | logit (stick-break) |
| $\beta_1$ | 0.0987 | 0.26 | log |
| $\beta_2$ | $-0.00157$ | $6.5 \times 10^{-4}$ | physical |
| $\gamma$ | $-1.272$ | 0.20 | physical |
| $\delta_1$ | $-1.708$ | 0.16 | physical |
| $\delta_2$ | 0.859 | 0.12 | log |
| $r$ | 0.307 | 0.042 | log |
| $\sigma_{\mathrm{init}}, \sigma_{\mathrm{input}}$ | 1.00 | 0.50 | log |

\textbf{Yasso15}

| Parameter | $x^{\mathrm{ctr}}$ | $\sigma_{\mathrm{ppm}}$ | Space |
|-----------|-------------------:|------------------------:|-------|
| $p_{ij}$ (12) | published defaults | 0.4 | logit (stick-break) |
| $\beta_1$ | 0.0906 | 0.0459 | log |
| $\beta_2$ | $-2.15 \times 10^{-4}$ | $1.4 \times 10^{-4}$ | physical |
| $\gamma$ | $-1.809$ | 0.0702 | physical |
| $\beta_{N1}$ | 0.0488 | 0.105 | log |
| $\beta_{N2}$ | $-7.92 \times 10^{-5}$ | $8 \times 10^{-5}$ | physical |
| $\gamma_N$ | $-1.173$ | 0.155 | physical |
| $\beta_{H1}$ | 0.0352 | 0.135 | log |
| $\beta_{H2}$ | $-2.08 \times 10^{-4}$ | $1.6 \times 10^{-4}$ | physical |
| $\gamma_H$ | $-12.54$ | 1.50 (capped) | physical |
| $\delta_1$ | $-0.439$ | 0.328 | physical |
| $\delta_2$ | 1.267 | 0.286 | log |
| $r$ | 0.257 | 0.055 | log |
| $\sigma_{\mathrm{init}}, \sigma_{\mathrm{input}}$ | 1.00 | 0.50 | log |

\textbf{Yasso20}

| Parameter | $x^{\mathrm{ctr}}$ | $\sigma_{\mathrm{ppm}}$ | Space |
|-----------|-------------------:|------------------------:|-------|
| $p_{ij}$ (12) | Yasso15 defaults | 0.4 | logit (stick-break) |
| $\beta_1$ | 0.158 | 0.104 | log |
| $\beta_2$ | $-0.002$ | $5.4 \times 10^{-4}$ | physical |
| $\gamma$ | $-1.440$ | 0.146 | physical |
| $\beta_{N1}$ | 0.170 | 0.0625 | log |
| $\beta_{N2}$ | $-0.005$ | $4.1 \times 10^{-4}$ | physical |
| $\gamma_N$ | $-2.000$ | 0.0353 | physical |
| $\beta_{H1}$ | 0.067 | 0.0912 | log |
| $\beta_{H2}$ | 0.000 | $9 \times 10^{-5}$ | physical |
| $\gamma_H$ | $-6.900$ | 1.391 | physical |
| $\delta_1$ | $-2.550$ | 0.435 | physical |
| $\delta_2$ | 1.240 | 0.211 | log |
| $r$ | 0.250 | 0.0545 | log |
| $\sigma_{\mathrm{init}}, \sigma_{\mathrm{input}}$ | 1.00 | 0.50 | log |

\textbf{SP1, TP2, TP3}

| Parameter | $x^{\mathrm{ctr}}$ | $\sigma_{\mathrm{ppm}}$ | Space |
|-----------|-------------------:|------------------------:|-------|
| \textit{SP1} | | | |
| $\alpha$ | 0.09 | 0.50 | log |
| $\beta_1, \beta_2, \gamma$ | 0.095, $-0.00014$, $-1.21$ | 0.26, $6.5\times10^{-4}$, 0.20 | log, phys., phys. |
| \textit{TP2} | | | |
| $\alpha_A, \alpha_H$ | 0.73, 0.0015 | 0.50, 0.50 | log |
| $p_H$ | 0.028 | 0.40 | logit |
| $\beta_1, \beta_2, \gamma$ | 0.095, $-0.00014$, $-1.21$ | 0.26, $6.5\times10^{-4}$, 0.20 | log, phys., phys. |
| \textit{TP3} | | | |
| $\alpha_A, \alpha_S, \alpha_H$ | 0.73, 0.10, 0.0015 | 0.50, 0.50, 0.50 | log |
| $p_S, p_H$ | 0.028, 0.50 | 0.40, 0.40 | logit |
| $\beta_1, \beta_2, \gamma$ | 0.095, $-0.00014$, $-1.21$ | 0.26, $6.5\times10^{-4}$, 0.20 | log, phys., phys. |
| \textit{All three} | | | |
| $\sigma_{\mathrm{init}}, \sigma_{\mathrm{input}}$ | 1.00 | 0.50 | log |

---

# Uncalibrated baselines

Before examining what calibration against the new data tells the models,
it is useful to know what the models say \emph{without} any such information
--- that is, how well published global parameter sets predict the observed
SOC stocks. These uncalibrated baseline runs serve as both a diagnostic and
a narrative anchor: they show where calibration starts and how far it
moves the predictions.

## Purpose and scope

Baseline scripts (`run_yasso07_baseline.R`, `run_yasso15_baseline.R`,
`run_yasso20_baseline.R`) run each Yasso model forward at its published
default parameters across all calibration-ready plots. No MCMC, no
posterior --- a single deterministic forward run per plot. Because SP1,
TP2, and TP3 have no published global calibration, no baselines are
produced for them; the natural reference for the simple models would be
their prior centres, but such runs are not currently generated.

Each baseline uses a steady-state initial condition rather than the
68-year pre-run. At published defaults $\sigma_{\mathrm{init}} = 1$ and
$\sigma_{\mathrm{input}} = 1$, so the steady state and the pre-run
endpoint coincide; the simpler steady-state formulation is used for
transparency. The two auxiliary uncertainty parameters are therefore not
exercised in the baseline --- their effect is, by construction, the
identity.

## Parameter choices

The baseline parameter vector for each model combines the published
decomposition rates, transfer fractions, climate-response parameters, and
size-modifier parameters from the source indicated below. Auxiliary
parameters are fixed at $\sigma_{\mathrm{init}} = \sigma_{\mathrm{input}} = 1$.

| Model   | Transfer fractions | Climate / size parameters |
|---------|--------------------|---------------------------|
| Yasso07 | Tuomi et al.\ published defaults (\texttt{y07par\_gui.csv}) | Same source |
| Yasso15 | Published Yasso15 defaults (\texttt{y15par.csv}) | Same source |
| Yasso20 | Yasso15 published defaults | FMI MAP (\texttt{Yasso20\_sample\_parameters.rda}) |

For Yasso07 and Yasso15 the baseline is the full published parameter set.
For Yasso20, the climate and size parameters are the FMI MAP values, but
the 12 transfer fractions use the Yasso15 published defaults rather than
the Viskari (2022) Yasso20 fraction values. At the published Yasso20
fractions, the per-pool outflow for two pools marginally exceeds the
available budget; the HIKET solver's diagonal-dominance guard (see the
Fortran-implementation section) classifies this as non-conservative and
the run is numerically unstable. The Yasso15 fractions satisfy the budget
and run stably. This substitution applies only to the Yasso20 baseline;
the Yasso20 calibration (§\ref{param-yasso20}) samples all 12 fractions
freely under the budget constraint and is unaffected.

\medskip\noindent\textbf{Status.} The choice of Yasso20 baseline
parameterisation is provisional. An alternative would be to reproduce the
original Yasso20 baseline using the unmodified FMI reference code, which
tolerates the published fractions without the diagonal-dominance guard.
Which option to adopt is left open pending the manuscript narrative.

## Diagnostic outputs

Each baseline produces the following outputs in
`diagnostics/{Model}_baseline/`:

| File | Content |
|------|---------|
| `*_mean_soc_trajectory.png` | Mean predicted SOC $\pm$ 1 SE across plots vs observed campaign means |
| `*_delta_soc_trajectory.png` | Mean annual $\Delta$SOC rate (tC ha$^{-1}$ yr$^{-1}$) vs observed |
| `*_obs_vs_pred.png` | Scatter coloured by vegetation type (`kasvup_tyyppi`) |
| `*_residuals_diagnostic.png` | 6-panel residual diagnostics |
| `*_rf_importance.png` | Random-forest permutation importance of site covariates |
| `*_worst25_timeseries.png` | Time series for the 25 worst-fitting plots |
| `*_residuals_histogram.png` | Distribution of log-scale residuals |
| `*_residuals.csv` | Per-observation residuals |
| `*_plot_ranking.csv` | Per-plot mean absolute log-residual |
| `*_report.txt` | Summary metrics (R$^2$, RMSE, bias, MAPE) and RF results |

The mean absolute SOC and the annual $\Delta$SOC trajectory plots use
the same visual form as the calibrated predictive outputs, allowing
direct side-by-side comparison of baseline and posterior trajectories
in the manuscript figures.

---

# Transient initialisation\label{transient}

## Why initialisation matters for the Finnish data

The observed Finnish SOC trajectory --- accumulating from 1985 through
Biosoil (2006), then approximately stabilising toward Komeetta (2024) ---
is not easily reproduced by a model pinned to an analytically-imposed steady
state with respect to its \emph{1985} litter inputs. Two features of the data
shape the initialisation problem. First, litter inputs are not static: the
cross-plot mean rises by roughly 40\% from the mid-1980s to a peak in the
mid-2000s and then plateaus (Figure \ref{fig:drivers}c), though individual
plots are heterogeneous --- some rise steeply as stands accumulate biomass,
others stay flat or fall after harvest. Second, this input rise is of
comparable magnitude to the observed SOC accumulation (mean stocks also rise
by roughly 40\% between 1985 and 2006), and the stocks level off at about the
same time the inputs peak.

These two facts have an important consequence. A rising litter supply is
itself a candidate driver of the accumulation: even a soil initialised at
equilibrium gains carbon while its inputs climb. But a model pinned to
$C(1985) = C_{\mathrm{ss}}(\text{1985 litter})$ can express only this one
mechanism --- the transient response to rising inputs --- and must fit the
entire accumulation through it, with no freedom to represent a 1985 soil that
was already \emph{below} the level its contemporary inputs would sustain. If
the true 1985 state was sub-equilibrium --- as the management history of
Finnish forests suggests, the post-war decades of intensive clearcutting,
drainage and fertilisation having left many stands still rebuilding carbon by
1985 --- that disequilibrium contribution is instead forced into implausible
kinetic parameters, or the model is rejected outright.

The transient initialisation approach removes this restriction. By
reconstructing the 1985 state from a calibrated pre-run (§\ref{transient})
rather than fixing it at the contemporary equilibrium, it lets the data
apportion the observed accumulation between the two mechanisms: the transient
response to rising inputs, and recovery from a sub-equilibrium initial state.
The free parameter $\sigma_{\mathrm{init}}$ controls this apportionment: it
sets the 1985 stock relative to the contemporary-input equilibrium, so a
posterior favouring $\sigma_{\mathrm{init}} < 1$ is direct evidence that the
1985 soil was below equilibrium and that disequilibrium recovery contributes
to the observed trend --- over and above the accumulation that the rising
contemporary inputs alone would produce.

## The 68-year pre-run

Instead of computing $C(1985)$ analytically from contemporary inputs,
each MCMC draw runs the model forward 68 years from 1917 to 1985, ending
at a calibrated initial state.

The choice of 1917 (Finland's independence) as the pre-run anchor is
partly practical (a natural reference point for Finnish forestry history)
and partly statistical (68 years is long enough that the fast AWEN pools
are approximately at steady state by 1985 regardless of the start
condition; only H retains meaningful memory of the initial state).

During the pre-run, two inputs must be specified:

\textbf{Climate:} No gridded climate data exist before 1961. The pre-run
therefore uses a constant climate modifier equal to $\xi(\bar{\mathrm{climate}})$
--- the modifier evaluated at the mean of the observed climate window
(1985--2004). This is an approximation, but an appropriate one: the
Yasso climate response is smooth and the mean is a reasonable
representative value for this multi-decadal period.

\textbf{Litter:} Litter inputs are linearly interpolated between a
historical anchor at 1917 and the start of the observed series at 1985:

$$J(i) = J_{1917} + \frac{i-1}{67}\,(J_{1985} - J_{1917}), \qquad i = 1, \ldots, 68.$$

The anchors are parameterised as:

$$J_{1917} = J_{\mathrm{full\,mean}} \cdot \sigma_{\mathrm{init}} \cdot \sigma_{\mathrm{input}},
\qquad
J_{1985}  = J_{t_0\,\mathrm{mean}}  \cdot \sigma_{\mathrm{input}},$$

where $J_{\mathrm{full\,mean}}$ is the mean litter input over the full
observed series and $J_{t_0\,\mathrm{mean}}$ is the mean over the first
five observed years (1985--1989). The pre-run begins at the analytical
steady state under $J_{1917}$ and constant climate.

## The two auxiliary uncertainty parameters

$\sigma_{\mathrm{init}}$ and $\sigma_{\mathrm{input}}$ are free
parameters in every model, both log-transformed with a prior centred at
1.0 (log-normal, unconstrained-space SD = 0.5, giving a physical 95\% prior
interval of approximately [0.37, 2.72] --- i.e. each multiplier lies between
about $0.37\times$ and $2.72\times$ its nominal value with 95\% prior
probability).

\textbf{$\sigma_{\mathrm{input}}$} scales the entire contemporary litter
time series (1985 onward). It absorbs uncertainty in the litter input
model --- plot-level litter estimates carry uncertainty from both the
litter model itself and the representation of understorey vegetation,
which is not individually measured. $\sigma_{\mathrm{input}}$ also
scales $J_{1985}$, the endpoint of the pre-run interpolation.

\textbf{$\sigma_{\mathrm{init}}$} scales only the 1917 litter anchor,
$J_{1917}$. It encodes the ratio of historical (pre-1985) to contemporary
mean productivity. The two parameters have strictly non-overlapping roles
in the pre-run formula:

\begin{quote}
$\sigma_{\mathrm{init}} > 1$: forests were historically \emph{more}
productive (old-growth or pre-intensification conditions). The 1985
initial SOC is above what contemporary litter inputs would sustain at
steady state. The model predicts declining SOC from 1985.

$\sigma_{\mathrm{init}} < 1$: forests were historically \emph{less}
productive (post-disturbance recovery). The 1985 initial SOC is below
the contemporary equilibrium. The model predicts accumulating SOC from
1985, consistent with the Finnish observations.

$\sigma_{\mathrm{init}} = 1$: the 1917 litter was equal to the
contemporary series mean; the pre-run approaches the same initial state
as the analytic steady state.
\end{quote}

The posterior on $\sigma_{\mathrm{init}}$ is therefore a direct inference
about the pre-1985 productivity history of Finnish forest plots. A
posterior that favours $\sigma_{\mathrm{init}} < 1$ across models
constitutes model-based evidence for the post-disturbance recovery
narrative.

## Jensen's inequality correction

The steady-state climate modifier used to start the pre-run is computed
as $\xi(\bar{\mathrm{climate}})$, not as $\overline{\xi(\mathrm{climate}_t)}$.
Because $\xi$ is convex in temperature and concave in precipitation,
Jensen's inequality gives $\overline{\xi_t} \geq \xi(\bar{\mathrm{climate}})$;
the bias is 5--15\% under typical Finnish inter-annual variability. Each
model wrapper provides a `compute_xi_mean_*` function that averages the
climate scalars over the steady-state window (first 20 observed years)
and evaluates $\xi$ once at the means, correctly avoiding the average-of-$\xi$
bias.

## Implementation across models

The pre-run protocol is uniform but adapted to each model's solver:

| Model | Pre-run implementation |
|-------|----------------------|
| SP1, TP2 | Analytical 68-step R loop (exact per-step solution) |
| TP3 | Analytical 68-step R loop (exact per-step matrix exponential) |
| Yasso07/15/20 | 68-row litter input fed to the Fortran solver; terminal per-cohort state extracted via `attr(run, "C_final")` |

For the Yasso models, the 1917--1985 propagation is advanced by the
compiled Fortran solver itself --- the same solver used in the forward
run, so the pre-run and the forward simulation share identical numerics.
This is a consistency choice, not a question of accuracy: Yasso's annual
step is analytical (matrix-based), exactly as the exact per-step solution
used by the SP1/TP2 R loops. The reason for calling the real solver is
faithful per-cohort bookkeeping: the 15-element terminal state (5 pools
$\times$ 3 litter cohorts) is returned via `C_final`, an output argument
added specifically to support this chaining. Without it, only the 5
summed pool totals would be available, which is insufficient to correctly
initialise the per-cohort bookkeeping for the forward simulation.

## Likelihood adjustment under transient initialisation

The engine retains a non-transient code path (the original steady-state
initialisation), kept only as a fallback; all six production calibrations
reported here run with `transient_init = TRUE`. The two paths differ in how
$\sigma_{\mathrm{init}}$ enters the likelihood. In the non-transient path,
$\sigma_{\mathrm{init}}$ inflated the observation standard deviation at the
first observation per plot --- a likelihood patch absorbing initial-state
uncertainty. Under transient initialisation $\sigma_{\mathrm{init}}$ has
already entered the model dynamics (it scales the 1917 litter anchor of the
pre-run, §\ref{transient}), so retaining the patch would double-count its
effect; the `transient_init = TRUE` flag therefore removes it. The
non-transient path is documented here only for completeness --- it is legacy
code, not part of the current results.

---

# Likelihood

## Form

Conditional on parameters, observations are modelled as

$$y_{p,t} \sim \mathcal{N}\!\bigl(\hat{C}_{p,t},\;
  \hat{C}_{p,t} \cdot \sigma_{\mathrm{obs}}\bigr),$$

where $y_{p,t}$ is observed SOC (tC ha$^{-1}$) at plot $p$, year $t$,
and $\hat{C}_{p,t}$ is predicted total SOC. This is a multiplicative-normal
model: the error standard deviation scales with the prediction, so
relative observation error is constant across the SOC range.

## Fixed $\sigma_{\mathrm{obs}}$

$\sigma_{\mathrm{obs}}$ is held at the empirical coefficient of variation
of the SOC observations across all calibration plots and years
($\approx 0.45$--$0.50$). It is not sampled.

This is a deliberate design choice. If $\sigma_{\mathrm{obs}}$ were
sampled, a structurally inadequate model could inflate it to absorb its
residual variance, hiding structural inadequacy. By fixing
$\sigma_{\mathrm{obs}}$ at the same empirical CV across all six models,
structural differences are forced into the residuals --- which is exactly
where the intercomparison wants them.

## Per-plot log-likelihood

For plot $p$ with observations at $\{t_1, \ldots, t_{n_p}\}$:

$$\log L_p = \sum_{k=1}^{n_p} \log \mathcal{N}\!\bigl(y_{p,t_k};\,
  \hat{C}_{p,t_k},\; \hat{C}_{p,t_k} \cdot \sigma_{\mathrm{obs}}\bigr).$$

Total log-likelihood: $\sum_p \log L_p$ plus the log-Jacobian of parameter
transforms. Plots are evaluated in parallel via `mclapply`.

## Rejection logic

Some parameter draws push a model into numerically invalid territory --- a
non-finite or non-positive predicted stock, or an outright solver failure ---
where the likelihood cannot be evaluated (the multiplicative-normal density is
undefined for $\hat{C} \leq 0$). For these the log-likelihood returns $-\infty$
and the proposal is rejected. This is a robustness guard, not an efficiency
device or a tuning knob: it stops invalid proposals (and any NaN/Inf they would
propagate) from contaminating the chains, while leaving the posterior shape in
the valid region untouched. Three hard rejections (return $-\infty$) apply:

1. $\hat{C}_{p,t}$ is non-finite at any observation year.
2. $\hat{C}_{p,t} \leq 0$ at any observation year (undefined in the
   multiplicative-error model).
3. The model run returned `NULL` (Fortran crash or wrapper exception).

No physical-ceiling guard is applied. Implausible predictions yield
large negative log-likelihoods and are rejected by MCMC naturally.
The likelihood evaluates $\hat{C}_{p,t}$ only at observation years;
trajectory values between campaigns are computed but not entered into
the likelihood.

## Holdout split for independent validation

Plots flagged `is_holdout` in `site_raw.csv` are excluded from the
likelihood sum: the calibration is fit only to the non-holdout plots.
The held-out plots are carried forward (via the inputs RDS) to the
predictive stage, where the prediction metrics --- $R^2$, RMSE, median bias,
and the parameter-CI coverage --- are computed \emph{separately} for the
calibration plots and the held-out plots. The held-out figures are a genuine
out-of-sample test: those plots never entered the likelihood, so they check
that the calibrated models generalise beyond the plots used to fit them rather
than only reproducing in-sample fit. Throughout the results, these
independent-validation numbers are kept in their own holdout table
(Table \ref{tab:holdout}), distinct from the calibration-sample table
(Table \ref{tab:calib}), and the predictive figures are labelled with the
sample they show, so it is always explicit whether a number is in-sample or
independent. The split is applied uniformly across all six models, using the
same `is_holdout` plot assignment so that every model is validated against the
same held-out plots.

---

# MCMC sampling

The Differential Evolution with snooker update sampler (DEzs) is used,
via the `BayesianTools` package (Hartig et al. 2023). DEzs proposes new
states using weighted differences between current chain states, then
accepts/rejects via the standard Metropolis ratio. It mixes well for
moderate-dimensional problems (10--50 parameters) and handles correlated
posteriors better than random-walk Metropolis.

## Settings

Shared across all six models, defined in `calib_config.R`:

| Setting | Value | Note |
|---------|-------|------|
| Chains | 5 | Sufficient for Gelman--Rubin $\hat{R}$ |
| Iterations | 50\,000 | Per chain |
| Burn-in | 5\,000 | 10\% of iterations |
| Thinning | None | Full chain retained |
| Initial state | prior centre $x^{\mathrm{ctr}}$ | All chains seeded at same point |

## Parallelisation

Two levels of parallelism are available:

* **Per-plot, within-chain:** each likelihood evaluation runs the per-plot
  inner loop in parallel via `mclapply` with `CORES_PER_CHAIN` workers.
* **Across-chain:** chains can be launched in parallel on a multi-core
  machine, each running its own per-plot parallelism.

The within-chain mode is mandatory on macOS (nested forking is unsafe).

---

# Sanity checks and diagnostics

## Pre-MCMC sanity check

Before launching MCMC, each calibration script runs a forward simulation
at the prior centre for four randomly selected plots. Three conditions
must hold:

1. All predicted SOC values are finite and positive.
2. Predicted-to-observed ratio lies in $[0.1, 10]$ at every observation year.
3. The likelihood `ll_fn(best_x)` is finite across all calibration plots.

A diagnostic PNG is written (`{Model}_forward_at_defaults_{RUN\_ID}.png`).
Condition 1 or 3 failure aborts the calibration; condition 2 produces a
warning only.

Input diagnostics (`diagnostics_inputs.R`, run separately) confirm that
implied mean residence times (MRT = mean SOC / mean litter) do not exceed
100 years. Plots exceeding this threshold are excluded upstream
(`calib_ready = FALSE`).

## Convergence diagnostics

Computed on post-burnin samples in unconstrained (sampling) space:

| Diagnostic | Threshold | Interpretation |
|------------|-----------|----------------|
| $\hat{R}$ (univariate) | $< 1.05$ | Good per-parameter convergence |
| $\hat{R}$ (multivariate) | $< 1.05$ | Good joint convergence |
| Effective sample size | $> 200$ | Adequate independent draws |
| Acceptance rate | 0.2--0.4 | Reasonable DEzs mixing |

With 20--26 free parameters in the Yasso models and only two to three SOC
observations per plot, most transfer fractions are weakly identified by the
aggregate SOC likelihood. It is tempting to expect this to show up as poor
$\hat{R}$, but that conflates non-identifiability with poor \emph{mixing}: once
the sampler mixes cleanly, it traverses the non-identified ridge without
difficulty and $\hat{R}$ can be excellent even for parameters the data does not
constrain. The genuine signature of non-identifiability is therefore not
$\hat{R}$ but (i) marginals pinned to the prior, (ii) near-perfect
anti-correlation ridges between fractions leaving the same pool, and (iii) low
KL divergence from prior to posterior. This distinction is borne out by the
production run (§\ref{sec:results}) and is developed there.

## Diagnostic outputs

Written to `Calibration_real_data_transient/diagnostics/{Model}/`:

| File | Content |
|------|---------|
| `*_traces_*.png` | Trace plots, one panel per parameter |
| `*_marginals_*.png` | Prior vs posterior KDE (shared x-grid) |
| `*_rhat_barplot_*.png` | $\hat{R}$ per parameter |
| `*_kl_divergence_*.png` | KL divergence posterior $\|$ prior |
| `*_report_*.txt` | Tabular $\hat{R}$, ESS, quantile summaries |
| `*_forward_at_defaults_*.png` | Pre-MCMC sanity check (4 plots) |

Prior and posterior KDE are evaluated on the same x-grid derived from
the prior's 99th-percentile range, so tight priors are not artificially
compressed.

---

# Output files

Written to `Calibration_real_data_transient/runs/`:

| File | Content | Used by |
|------|---------|---------|
| `{Model}_chains_{RUN\_ID}.rds` | BayesianTools chain objects (unconstrained space) | Re-diagnostics |
| `{Model}_posterior_{RUN\_ID}.rds` | Post-burnin samples in physical space | Predictive scripts |
| `{Model}_metadata_{RUN\_ID}.rds` | Run configuration, wallclock, $\hat{R}$, ESS | Reproducibility |
| `{Model}_inputs_{RUN\_ID}.rds` | Per-plot input structures used in calibration | Predictive scripts |

The inputs RDS contains `climate_by_plot`, `inputs_by_plot`,
`litter_means`, `obs_meta`, `plot_info`, `plots_real`, `holdout_plots`,
`sigma_obs_fixed`, `sigma_ppm`, `STEADY_STATE_YEARS`, and the transient
pre-run parameters (`PREINIT_YEAR`, `N_PREINIT_SMOOTH`). It provides
exact provenance: predictive scripts use the same per-plot structures the
calibration used, regardless of any subsequent updates to `Data_work.R`.

---

# Results\label{sec:results}

```{r results-helpers, include=FALSE}
# Locate the most recent diagnostic figures by pattern. Filenames embed the
# per-model RUN_IDs, so we glob and pick the newest match rather than hard-code.
diag_dir <- "../diagnostics"
mm_dir   <- file.path(diag_dir, "multimodel")
latest <- function(dir, pattern) {
  f <- list.files(dir, pattern = pattern, full.names = TRUE)
  if (length(f) == 0) return(NA_character_)
  f[which.max(file.mtime(f))]
}
show_fig <- function(path) if (!is.na(path) && file.exists(path)) knitr::include_graphics(path)
```

This section reports the outcome of the homogenised, transient-initialised
calibration. All six models were calibrated against the same Finnish dataset
(\num{358} plots, \num{1121} SOC observations), with an independent holdout
of \num{224} observations (\num{897} retained for the random-forest residual
analysis) excluded from the likelihood and scored separately. The narrative
proceeds from the figures to the numbers and closes with a short
interpretation; the emphasis is on \emph{predictive skill of the ensemble},
with convergence and information-gain treated as supporting evidence.

## Uncalibrated baselines

Before calibration, the three Yasso models were run at their published
parameter sets with no input scaling ($\sigma_{\mathrm{input}}=1$), to
establish what the literature defaults predict on Finnish data
(Table \ref{tab:baseline}). All three explain almost none of the
between-plot variance ($R^2 \approx 0.02$) and, more tellingly, all three
\emph{under-predict} the observed stocks. The bias worsens monotonically with
model complexity: Yasso07 is nearly unbiased ($-1.1$ tC ha$^{-1}$), Yasso15
under-predicts by $13$ tC ha$^{-1}$, and Yasso20 by $41$ tC ha$^{-1}$ --- a
gross under-prediction consistent with the per-pool mass-balance issue
documented for the published Yasso20 parameter set (§\ref{sec:fortran-diffs}
and the accompanying mass-balance note).

Table: Uncalibrated baseline performance at published defaults (447 plots,
1121 matched observations). Bias is mean predicted $-$ observed; negative =
under-prediction. \label{tab:baseline}

| Model    | $R^2$ | RMSE (tC ha$^{-1}$) | Bias (tC ha$^{-1}$) | MAPE |
|----------|------:|--------------------:|--------------------:|-----:|
| Yasso07  | 0.021 | 54.1 | $-1.1$  | 50.2\% |
| Yasso15  | 0.026 | 52.5 | $-13.0$ | 44.9\% |
| Yasso20  | 0.023 | 60.9 | $-40.8$ | 45.9\% |

```{r fig-baseline-traj, fig.cap="Uncalibrated Yasso baselines at published defaults: cross-plot mean predicted SOC trajectories for Yasso07, Yasso15 and Yasso20 (lines) against the observed campaign means (points, $\\pm$1 SE). All three under-predict the 1985--2006 accumulation, and the under-prediction deepens with model complexity --- most severely for Yasso20, consistent with the published-parameter mass-balance issue (\\S\\ref{sec:fortran-diffs}). Yasso20 is shown alongside the other two, not interpreted alone. \\label{fig:baseline}", out.width="82%", fig.align="center", fig.pos="H"}
ybase <- c("Yasso07", "Yasso15", "Yasso20")
ycol  <- c(Yasso07 = "#1b9e77", Yasso15 = "#7570b3", Yasso20 = "#d95f02")
traj  <- lapply(ybase, function(m) {
  f <- file.path(diag_dir, paste0(m, "_baseline"), paste0(m, "_baseline_predictions.rds"))
  aggregate(total_soc ~ year, readRDS(f), mean)
})
names(traj) <- ybase
# observed campaign means (same for every model; read once)
rb   <- read.csv(file.path(diag_dir, "Yasso07_baseline", "Yasso07_baseline_residuals.csv"))
om   <- tapply(rb$soc_obs_tCha, rb$year, mean, na.rm = TRUE)
ose  <- tapply(rb$soc_obs_tCha, rb$year, function(x) sd(x, na.rm = TRUE) / sqrt(sum(is.finite(x))))
obs  <- data.frame(year = as.integer(names(om)), mean = as.numeric(om), se = as.numeric(ose))
xr   <- range(unlist(lapply(traj, function(d) d$year)))
yr   <- range(c(unlist(lapply(traj, function(d) d$total_soc)), obs$mean - obs$se, obs$mean + obs$se))
par(mar = c(4, 4.4, 1, 1), mgp = c(2.5, 0.7, 0), las = 1)
plot(NA, xlim = xr, ylim = yr, xlab = "Year",
     ylab = expression("Cross-plot mean SOC (tC ha"^-1*")"))
for (m in ybase) lines(traj[[m]]$year, traj[[m]]$total_soc, col = ycol[m], lwd = 2.2)
arrows(obs$year, obs$mean - obs$se, obs$year, obs$mean + obs$se,
       angle = 90, code = 3, length = 0.04, col = "black")
points(obs$year, obs$mean, pch = 21, bg = "black", cex = 1.3)
legend("topleft", bty = "n",
       legend = c(ybase, "Observed campaign mean (±1 SE)"),
       col = c(ycol[ybase], "black"),
       lwd = c(2.2, 2.2, 2.2, NA),
       pch = c(NA, NA, NA, 21), pt.bg = c(NA, NA, NA, "black"))
```

## Transient initialisation: evidence for a sub-equilibrium initial state

The transient initialisation (§\ref{transient}) is the keystone methodological
choice of the work described here, and the calibration provides a direct test
of it. Rather than
assuming the 1985 soil sits at steady state, each model reconstructs its 1985
pool state from a 68-year pre-run whose historical litter anchor is scaled by
the free parameter $\sigma_{\mathrm{init}}$ --- the ratio of 1917 to
contemporary litter input. A value below 1 means the soil entered the
observation period below equilibrium and still accumulating; a value of 1
recovers a flat, steady-state-like history. Because $\sigma_{\mathrm{init}}$ is
estimated rather than fixed, its posterior is itself an answer to the question
\emph{does the data want a transient start?}

The posteriors are unambiguous (Figure \ref{fig:siginit},
Table \ref{tab:siginit-post}): \textbf{every model places the entire 95\%
credible interval of $\sigma_{\mathrm{init}}$ below 1.} Medians range from
$0.19$ (TP2) to $0.72$ (Yasso15), and no model's upper 97.5\% bound reaches
unity (the highest, Yasso15, stops at $0.91$). This is a structure-independent
result: regardless of pool count or climate form, the Finnish SOC data reject a
steady-state 1985 start and require a sub-equilibrium, accumulating initial
condition. The gradient across models is itself informative --- the single- and
two-pool models infer the most extreme historical deficit (medians near
$0.2$--$0.3$), leaning harder on the initial state because they have fewer
internal degrees of freedom, whereas the Yasso family, with its slow humified
pool, reaches the same observed stocks from a milder deficit (medians near
$0.5$--$0.7$). This gradient is the mirror image of the litter-multiplier result
(§\ref{sec:identifiability}): the models that infer the largest historical
deficit also rescale contemporary inputs the hardest (TP2/TP3
$\sigma_{\mathrm{input}}\approx 13$), because their fast pools store little
carbon per unit flux and must be fed harder to reach the observed stocks. So
$\sigma_{\mathrm{init}}$ and $\sigma_{\mathrm{input}}$ are not independent
findings but two coordinates of a single structural compensation: each model
reaches the same SOC by its own trade-off between how far below equilibrium it
starts and how much it rescales the flux.

Table: Posterior of the historical litter ratio $\sigma_{\mathrm{init}}$
(median and 95\% CI). Every interval lies entirely below 1.
\label{tab:siginit-post}

| Model   | Median | 95\% CI         |
|---------|-------:|:---------------:|
| SP1     | 0.26   | [0.16, 0.37]    |
| TP2     | 0.19   | [0.11, 0.31]    |
| TP3     | 0.33   | [0.19, 0.57]    |
| Yasso07 | 0.51   | [0.40, 0.64]    |
| Yasso15 | 0.72   | [0.59, 0.91]    |
| Yasso20 | 0.63   | [0.48, 0.85]    |

```{r fig-siginit, fig.cap="Posterior of the historical litter ratio sigma-init across the six models (open circle = median, thick bar = 50\\%, thin bar = 95\\% interval). The dashed line marks the steady-state value of 1; every posterior lies entirely to its left, so all models independently infer a sub-equilibrium, accumulating 1985 state. \\label{fig:siginit}", out.width="80%", fig.align="center"}
show_fig(latest(mm_dir, "multimodel_sigma_init_[0-9].*\\.png"))
```

This bears directly on HIKET's motivating question. Studies that initialise SOC
models at steady state start every model on a flat trajectory, which
predisposes them to project saturation or decline; an apparent saturation may
then be partly an artefact of the initial condition rather than a genuine
structural property of the model. By estimating $\sigma_{\mathrm{init}}$ instead
of fixing it, HIKET lets the data decide, and the data decide for accumulation:
the same models that would relax to a flat trajectory from a steady-state start
instead produce a rising, gradually saturating one once the initial state is
freed (Figures \ref{fig:delta}, \ref{fig:meansoc}) --- the very behaviour HIKET
sets out to study. The models match the 2006 and 2024 campaigns well; the
residual is that they begin above the observed 1985 baseline and accumulate
somewhat less steeply, which we treat as an open question tied to the coarse,
data-free pre-1985 phase (see Discussion) rather than a model deficiency.
Ecologically, $\sigma_{\mathrm{init}}<1$ is consistent
with Finnish forest history --- lower mid-century litter inputs under younger,
less productive, post-war and afforested stands building toward present-day
levels --- though the parameter conflates that productivity history with any
slow disequilibrium and should be read as a magnitude rather than a literal
1917 flux.

## Convergence and sampler health

Sampling was healthy across all six models. Univariate $\hat{R}$ stayed below
\num{1.05} for every parameter in every model with a single exception
(SP1 `alpha`, $\hat{R}=1.092$, discussed below); multivariate $\hat{R}$ ranged
\numrange{1.007}{1.067}, and effective sample sizes were comfortably above the
\num{200} threshold (ESS \numrange{208}{3663}). Invalid-proposal rates were
negligible for the well-behaved models ($<0.1\%$) and elevated only where the
posterior has a genuinely heavy tail (SP1 chain 5 $12\%$; TP3 $8$--$17\%$;
both traced to long right tails in `alpha`/kinetic parameters, not to displaced
chains).

The single notable $\hat{R}$ signal --- SP1 `alpha` --- is \emph{not} a
non-identifiability symptom but its opposite: `alpha` is the most strongly
\emph{identified} parameter in the entire study, with a heavy right tail
(its \num{97.5}th percentile is $\sim 6\times$ the median) that is simply hard
to sample. Crucially, every Yasso transfer fraction converged in $\hat{R}$
despite being weakly identified by the data. Good $\hat{R}$ is thus the
\emph{expected} result of a cleanly mixing sampler and says nothing about
whether a parameter is informed; the identifiability question is taken up in
§\ref{sec:identifiability}.

## Predictive skill of the calibrated ensemble

The calibrated models are compared in Table \ref{tab:calib} (calibration
sample) and Table \ref{tab:holdout} (independent holdout), and visually in
Figures \ref{fig:obspred}--\ref{fig:resid}. Three features stand out.

\textbf{Skill at the level of individual plots is low, and similar across
models.} Calibration $R^2$ ranges \numrange{0.05}{0.11} and holdout $R^2$
\numrange{0.02}{0.05}; RMSE clusters near \num{50} tC ha$^{-1}$ (calibration)
and \num{60} tC ha$^{-1}$ (holdout) for all six. This refers specifically to
predicting the SOC stock of a \emph{single} plot from its litter and climate
--- a genuinely hard target: the random-forest residual analysis
(§\ref{sec:identifiability}) recovers an out-of-bag $R^2$ of only
\numrange{0.34}{0.52} even with \num{54} site covariates, so the low ceiling is
a property of the data, not of any one model. That none of the models comes
close to that ceiling is itself informative --- it points to drivers of
plot-level SOC that the litter--climate--kinetics structure does not capture,
and the residual analysis (§\ref{sec:identifiability}) is where we ask what
those drivers might be.

\textbf{The ranking is consistent, but rests on small absolute differences.}
TP3 is nominally the best model on both calibration ($R^2=0.109$) and holdout
($R^2=0.052$) and Yasso20 the weakest ($R^2=0.049/0.016$), with the two- and
three-pool transient models (TP2, TP3) edging out the full Yasso family. The
ordering is stable across the calibration/holdout split --- so it is not simply
in-sample overfitting --- but it should be read with care: at $R^2$ this low,
the models are separated by a few percent of an already small explained-variance
budget, and we would not claim any one of them is meaningfully ``better'' at
predicting an individual plot. The ranking is useful as a \emph{structural}
signal --- the simple transient models are at least as good as the full Yasso
family on these data --- not as a performance verdict.

\textbf{All calibrated models over-predict, but by structurally different
amounts.} Median bias is positive for every model, and its size tracks the
skill ranking: the simple transient models are closest to unbiased (TP2
\num{+6.2}, TP3 \num{+6.2} tC ha$^{-1}$ in calibration; \num{+2.3} and
\num{+1.3} on the holdout), whereas the Yasso family over-predicts more,
increasing through Yasso07 (\num{+9.3}) and Yasso15 (\num{+8.8}) to Yasso20,
the most biased (\num{+11.7} tC ha$^{-1}$ calibration, \num{+7.8} holdout).
Across all six the calibration bias of \numrange{+6}{+12} tC ha$^{-1}$ shrinks
to \numrange{+1}{+8} on the holdout --- a reversal of the baseline
under-prediction. The `Param cov` column (fraction of observations
inside the 95\% credible interval of the predicted \emph{mean}) is small by
construction: this interval reflects parameter uncertainty in the model
expectation only and deliberately excludes the multiplicative observation
error, so it is a measure of how tightly the mean is pinned down, not a
posterior-predictive coverage.

\textbf{The absolute trajectory reproduces the rising--saturating shape, with a
caveat at the 1985 baseline.} Plotting mean SOC over time
(Figure \ref{fig:meansoc}) shows that all six calibrated models reproduce the
behaviour at the heart of HIKET: a SOC stock that increases after 1985 and
begins to level off toward the recent campaigns --- the incipient sink
saturation that the Yasso family in particular projects --- and they match the
2006 and 2024 campaign means closely. The one clear discrepancy is at the start:
the observed mean rises from $63$ tC ha$^{-1}$ in 1985 to $102$ in 2006 and
$105$ in 2024 (a $+42$ increase), whereas every model begins its 1985 trajectory
above the observed baseline (predicted 1985 means $76$--$95$) and accumulates
somewhat less steeply ($+11$ to $+31$). Within this common pattern the ranking
is mechanistic: the simpler TP2/TP3 start lowest and accumulate most, tracking
the \emph{shape} of the trend best, while the stiffer Yasso family starts
highest and accumulates least. We do not read the 1985 offset as a model failure
--- it is entangled with the deliberately coarse, data-free transient phase and
with an unresolved tension in the litter inputs, taken up in the Discussion.

\textbf{The starting increment encodes the same structural split.} The
annual-increment view (Figure \ref{fig:delta}) shows this from a different
angle: TP2 and TP3 begin with the steepest accumulation rate, SP1 with the
shallowest, and the Yasso family in between. This is the dynamical fingerprint
of the input/initial-state compensation above (§\ref{sec:identifiability}). TP2
and TP3 hold carbon in fast pools, so reaching the observed stocks forces an
order-of-magnitude input inflation ($\sigma_{\mathrm{input}}\approx 13$) and a
deep initial deficit ($\sigma_{\mathrm{init}}\approx 0.2$--$0.3$); starting far
below a high, fast-turning equilibrium produces a large early increment that
decays as the deficit closes. SP1, with a single slow pool, instead reaches the
stocks by \emph{lowering} inputs ($\sigma_{\mathrm{input}}\approx 0.3$) and
turning over slowly: its target equilibrium is low and its approach gradual, so
the increment is shallow and nearly flat. The spread in starting increments is
therefore not noise but the same structural disagreement --- about \emph{how}
each model reaches the stocks --- seen in the time derivative.

\textbf{The forcing is strongly oscillatory in climate but smooth in litter.}
Figure \ref{fig:drivers} summarises the three model drivers as cross-plot means
over the observed period. Mean annual temperature and precipitation swing sharply
from year to year ($\pm 1$--$2^\circ$C and $\pm 100$--$150$ mm), whereas the
litter input rises smoothly to a mid-2000s peak and then plateaus, with no
high-frequency component, so any interannual variation in SOC could only come
from the climate modifier $\xi(T,P)$, not the inputs. TP3 shows a visible
year-to-year oscillation in mean SOC (Figure \ref{fig:meansoc}) that the other
models do not. An earlier TP3 run inflated it with a numerical artefact --- an
explicit Euler integrator overshooting a sub-annual active pool --- which we
removed by integrating TP3 exactly like the rest of the ensemble; but a bounded,
\emph{physical} oscillation survives exact integration, because TP3's calibrated
active pool genuinely turns over faster than a year and so tracks the climate
modifier. The distinction, and why TP2 stays smooth despite an equally fast pool,
is developed in §\ref{sec:tp3osc}.

Table: Calibrated predictive performance (full calibration sample, 1121 obs,
median-prediction metrics). \label{tab:calib}

| Model    | $R^2$ | RMSE | Bias  | Param cov |
|----------|------:|-----:|------:|----------:|
| SP1      | 0.065 | 51.3 | $+8.2$  | 0.056 |
| TP2      | 0.089 | 49.6 | $+6.2$  | 0.086 |
| TP3      | **0.109** | 50.0 | $+6.2$  | 0.089 |
| Yasso07  | 0.067 | 52.3 | $+9.3$  | 0.062 |
| Yasso15  | 0.060 | 52.2 | $+8.8$  | 0.062 |
| Yasso20  | 0.049 | 53.6 | $+11.7$ | 0.054 |

Table: Calibrated predictive performance on the independent holdout (224 obs).
\label{tab:holdout}

| Model    | $R^2$ | RMSE | Bias  | Param cov |
|----------|------:|-----:|------:|----------:|
| SP1      | 0.027 | 60.4 | $+3.8$ | 0.071 |
| TP2      | 0.036 | 59.1 | $+2.3$ | 0.071 |
| TP3      | **0.052** | 58.4 | $+1.3$ | 0.098 |
| Yasso07  | 0.027 | 61.9 | $+5.5$ | 0.062 |
| Yasso15  | 0.026 | 60.5 | $+4.3$ | 0.080 |
| Yasso20  | 0.016 | 62.3 | $+7.8$ | 0.062 |

```{r fig-obspred, fig.cap="Observed vs predicted SOC across the six calibrated models (calibration sample). Points are plot-level posterior medians; the spread of predictive skill between models is small. \\label{fig:obspred}", out.width="100%", fig.align="center"}
show_fig(latest(mm_dir, "multimodel_obs_vs_pred_[0-9].*\\.png"))
```

```{r fig-resid, fig.cap="Residual structure across models (calibration sample). Residuals are broadly comparable in scale and share the same site-driven structure (Section \\ref{sec:identifiability}). \\label{fig:resid}", out.width="100%", fig.align="center"}
show_fig(latest(mm_dir, "multimodel_residuals_[0-9].*\\.png"))
```

```{r fig-drivers, fig.cap="The three model drivers over the observed period (cross-plot mean line, band = 95\\% CI of the mean; the 1985 initialisation year is omitted). Mean annual temperature (a) and precipitation (b) are strongly oscillatory year-to-year, whereas total litter input (c) rises smoothly and then plateaus. Interannual oscillations in a model's SOC trajectory (Figure \\ref{fig:meansoc}) therefore reflect the climate modifier, not the inputs. \\label{fig:drivers}", out.width="80%", fig.align="center"}
show_fig(latest(mm_dir, "multimodel_drivers_[0-9].*\\.png"))
```

```{r fig-delta, fig.cap="Mean annual SOC increment across models over the projection window. Transient initialisation yields a rising, gradually saturating trajectory; the modelled increment is somewhat lower than the observed accumulation rate (see the absolute levels in Figure \\ref{fig:meansoc}). \\label{fig:delta}", out.width="100%", fig.align="center"}
show_fig(latest(mm_dir, "multimodel_delta_soc_[0-9].*\\.png"))
```

```{r fig-meansoc, fig.cap="Mean absolute SOC across plots over time for each calibrated model (line = predicted cross-plot mean, band = +/-1 SE; red = observed campaign means +/-95\\% CI); the calibrated analogue of the published-defaults baseline trajectories. All models reproduce the rising-then-levelling shape and match the 2006 and 2024 campaigns closely, but begin above the observed 1985 baseline and accumulate somewhat less steeply; TP2/TP3 track the shape best, the Yasso family least. \\label{fig:meansoc}", out.width="100%", fig.align="center"}
show_fig(latest(mm_dir, "multimodel_mean_soc_[0-9].*\\.png"))
```

## Baseline versus calibrated

Calibration changes the picture in three ways. First, it removes the
structural under-prediction: the strongly negative baseline biases (Yasso15
$-13$, Yasso20 $-41$ tC ha$^{-1}$) become small positive biases
($+8.8$, $+11.7$). Second, it lifts $R^2$ from $\sim 0.02$ to
\numrange{0.05}{0.11} --- a real but modest gain, reinforcing that the
explainable between-plot variance is limited. Third, and most informative, the
correction is achieved almost entirely through the litter-input multiplier
$\sigma_{\mathrm{input}}$ rather than through the decomposition kinetics, which
sets up the identifiability discussion below.

## Residual structure\label{sec:residuals}

Beyond the headline metrics, the residuals carry two further signals.

\textbf{The over-prediction is a uniform multiplicative offset, not a reporting
artefact.} Re-scoring the calibrated predictions on the log scale --- the natural
scale of the multiplicative error model --- gives a mean log-residual
($\log\mathrm{obs}-\log\widehat{\mathrm{SOC}}$) of $-0.10$ to $-0.16$ across the
six models, i.e. predictions sit $\approx 10$--$16\%$ high in geometric-mean
terms. The value is essentially identical whether the predictive mean or median
is used (the predictive spread is narrow), so the bias is not an artefact of
summarising a skewed predictive distribution by its mean. It also tracks
predictive skill: the offset is smallest for the highest-$R^2$ models (TP2, TP3,
$\approx -0.10$) and largest for the lowest (Yasso20, $-0.16$). We argue in the
Discussion (§\ref{sec:biasdisc}) that this is a property of the proportional
likelihood rather than a physical statement about Finnish soils.

\textbf{Residuals are site-structured, and the structure is shared across
models.} A random forest regressing the log-residuals on \num{54} site
covariates recovers an out-of-bag $R^2$ of \numrange{0.34}{0.52}
(Table \ref{tab:rfoob}), so a substantial and partly reducible share of the
unexplained between-plot variance is site-driven rather than random. The
importance ranking is strikingly consistent across all six models
(Figure \ref{fig:rfheat}): \textbf{stand basal area at 1985}
(\texttt{basal\_area\_85}) is the single most important predictor in five of the
six models --- and second in TP3 --- followed by soil chemistry (total nitrogen,
cation-exchange capacity, exchangeable aluminium, clay) and then the climate
variables (mean and warmest-month temperature). No single covariate exceeds
$\sim 5$--$9\%$ of total importance, so the residual signal is diffuse rather
than concentrated in one driver. Notably, the covariate set \emph{does} include
the categorical site classifiers one might expect to carry a regular bias ---
NFI development class, Cajander fertility class, growing-site type
(\texttt{kasvup\_tyyppi}), and dominant species --- yet none of them ranks among
the leading predictors (development class, for instance, is 17th--38th). There
is thus no clean class-wise residual bias, e.g.\ by site or soil type, of the
kind that might have been anticipated; the residual structure is carried by
continuous gradients (biomass, soil chemistry, climate), not by discrete
site categories. That structurally different models have their residuals
ordered by the \emph{same} covariates confirms the unexplained variance is a
property of the data and the site, not of any one model's kinetics.

Table: Out-of-bag $R^2$ of the random-forest residual model, per calibrated
model (calibration sample, \num{54} site covariates, log-residual target).
\label{tab:rfoob}

| Model    | SP1  | TP2  | TP3  | Yasso07 | Yasso15 | Yasso20 |
|----------|-----:|-----:|-----:|--------:|--------:|--------:|
| OOB $R^2$ | 0.42 | 0.50 | 0.52 | 0.42    | 0.38    | 0.34    |

```{r fig-rfheat, fig.cap="Random-forest residual-predictor importance (relative, within-model) for the calibration sample; the most important covariates are shown and column labels give the out-of-bag $R^2$. Stand basal area at 1985 (\\texttt{basal\\_area\\_85}) tops every model, followed by soil chemistry and climate, and the ordering is shared across structurally different models. \\label{fig:rfheat}", out.width="85%", fig.align="center"}
show_fig(latest(mm_dir, "multimodel_rf_importance_heatmap_[0-9].*\\.png"))
```

### [TO DISCUSS --- stand structure/age as the leading residual driver]

\emph{For joint interpretation.} The leading residual predictor is stand basal
area at 1985, a proxy for stand development stage and standing biomass; the
explicit NFI development class (\texttt{dev\_class\_85}) is by contrast a weak
predictor (ranked 17th--38th across models). No direct stand-age variable is
currently in the covariate set. Whether to read \texttt{basal\_area\_85} as a
stand-age/development signal --- and whether to add an explicit age covariate to
sharpen it --- is the open interpretive question to settle before finalising
this paragraph.

## Identifiability and information gain\label{sec:identifiability}

The convergence diagnostics confirm that good $\hat{R}$ coexists with weak
identification. Three independent signatures locate where the data actually
informs the parameters.

\textbf{Prior-pinning.} The fraction of free parameters sitting within
$0.5\sigma$ of their prior centre rises sharply with model dimension: SP1
1/6, TP2 2/8, TP3 6/10, Yasso07 13/20, Yasso15 16/26, Yasso20 17/26. The added
degrees of freedom in the larger models are absorbed by the prior, not by the
data. The per-parameter prior-vs-posterior marginals (Appendix
§\ref{sec:marginals}) show this directly.

\textbf{Ridge correlations.} In all three Yasso models the transfer fractions
leaving a common pool are almost perfectly anti-correlated --- e.g.
$\rho(p_{EN},p_{EA}) \approx -0.98$ and $\rho(p_{NE},p_{NA}) \approx -0.89$ to
$-0.98$. The likelihood constrains the \emph{net} flux out of each pool, not how
that flux splits between destinations; the split rides freely along an
anti-correlation ridge. The per-parameter KL spectra for all three Yasso
models (Figures \ref{fig:kl07}--\ref{fig:kl20}), with bars coloured by
parameter class, tell the same story in each: the twelve transfer fractions
(orange) each gain $<1$ nat, while the auxiliary turnover parameters (purple)
and the climate-and-size modifiers (green) absorb almost all of the information
the data carry. The pattern is consistent across Yasso07, Yasso15 and Yasso20,
which is what makes the diagnosis structural rather than model-specific.

\textbf{Where the information goes: $\sigma_{\mathrm{input}}$.} The litter
multiplier is the parameter the data speaks to most loudly, and it does
strikingly different work in each structure (Table \ref{tab:siginput}). The
single-pool SP1 scales inputs \emph{down} by a factor of three (median
\num{0.31}); the two- and three-pool transient models scale them \emph{up}
by an order of magnitude (TP2 $\sim 13\times$, TP3 $\sim 13\times$); the
Yasso family sits in between, with Yasso20 needing a threefold increase to
overcome its baseline under-prediction. This is itself an intercomparison
result: each structure reaches the same observed stocks by a different
implied litter flux, and $\sigma_{\mathrm{input}}$ is where that structural
disagreement is recorded. For the Yasso development programme specifically,
the reading is that --- on this dataset --- the two SOC campaigns inform the
\emph{aggregate turnover} (via $\sigma_{\mathrm{input}}$ and the climate
modifiers) but carry almost no information about the inter-pool partitioning;
collecting pool-resolved or fractionated measurements would be the way to
identify the transfer fractions.

Table: Posterior of the litter-input multiplier $\sigma_{\mathrm{input}}$
(median and 95\% CI). A value of 1 means no rescaling.
\label{tab:siginput}

| Model   | Median | 95\% CI        | Implied input change |
|---------|-------:|:--------------:|:--------------------:|
| SP1     | 0.31   | [0.25, 0.37]   | $\sim 3\times$ lower  |
| TP2     | 13.4   | [8.0, 21.1]    | $\sim 13\times$ higher |
| TP3     | 12.7   | [10.3, 15.2]   | $\sim 13\times$ higher |
| Yasso07 | 1.37   | [1.04, 1.74]   | near baseline          |
| Yasso15 | 1.91   | [1.58, 2.28]   | $\sim 2\times$ higher  |
| Yasso20 | 2.98   | [2.26, 3.91]   | $\sim 3\times$ higher  |

```{r fig-kl07, fig.cap="Information gain (KL divergence, posterior $\\|$ prior) per parameter for Yasso07, bars coloured by parameter class. The 12 transfer fractions (orange) each gain well under 1 nat, while the auxiliary turnover parameters (purple) and climate-and-size modifiers (green) gain the most --- the data inform aggregate turnover, not the inter-pool splits. \\label{fig:kl07}", out.width="85%", fig.align="center"}
show_fig(latest(file.path(diag_dir, "Yasso07"), "Yasso07_kl_divergence_[0-9].*\\.png"))
```

```{r fig-kl15, fig.cap="KL divergence per parameter for Yasso15 (same colour scheme). The transfer fractions again gain almost no information relative to the prior. \\label{fig:kl15}", out.width="85%", fig.align="center"}
show_fig(latest(file.path(diag_dir, "Yasso15"), "Yasso15_kl_divergence_[0-9].*\\.png"))
```

```{r fig-kl20, fig.cap="KL divergence per parameter for Yasso20 (same colour scheme). Despite all twelve fractions being free, none is appreciably informed by the data. \\label{fig:kl20}", out.width="85%", fig.align="center"}
show_fig(latest(file.path(diag_dir, "Yasso20"), "Yasso20_kl_divergence_[0-9].*\\.png"))
```

## Discussion

The results support a coherent reading. Once initialisation, likelihood, error
model, and priors are homogenised, the six models converge to a narrow band of
modest predictive skill, and the small differences that remain track structure
coherently across the calibration/holdout split. Crucially, the transient
initialisation lets the calibrated models reproduce the phenomenon HIKET set out
to examine: a post-1985 SOC increase that begins to saturate toward the present
(Figures \ref{fig:delta}, \ref{fig:meansoc}), in place of the flat or declining
trajectory a steady-state start imposes. The match at the 2006 and 2024
campaigns is good; the main residual is an over-predicted 1985 baseline and a
somewhat shallower accumulation slope, captured best by the simpler models.

That residual is better read as an open question than as a model deficiency,
because the pre-1985 phase of the approach is deliberately coarse. No historical
litter or climate record exists for 1917--1985: the pre-run interpolates litter
linearly from a single calibrated 1917 anchor ($\sigma_{\mathrm{init}}$) under a
constant climate, so the early trajectory rests entirely on calibration and
carries large uncertainty. A connected and still-unresolved tension concerns the
litter inputs. The plot-level Tupek inputs used here are approximately flat over
the observed period, whereas national-scale input reconstructions are lower in
the past and rise toward the present --- a shape that mirrors the SOC trend
itself. If the true historical inputs were lower and increasing, both the
over-predicted 1985 baseline and the too-shallow slope would shrink; if the flat
plot-level inputs are correct, then the coincidence of a rising SOC trend with
flat inputs is itself informative, and may be what masked the initialisation
problem in earlier work. The present data cannot settle this. The transient
calibration is therefore best framed as a proof of concept --- a first
demonstration that the approach can represent the observed accumulation and
incipient saturation --- whose natural successor is a dedicated historical
reconstruction supplying genuine priors for the pre-1985 phase rather than
leaving it to calibration.

Two structural messages emerge. First, the three simple transient models are
not just a low-complexity baseline for the Yasso family to beat: in plot-level
predictive skill TP2 and TP3 are competitive with, and on this dataset
marginally better than, the full Yasso family, while remaining far easier to
identify. We are careful not to over-read this --- ``competitive on $R^2$'' is
not ``structurally adequate''. The aggregate mean-SOC trajectories of the
simple models (Figure \ref{fig:meansoc}) are less convincing than their
plot-level metrics suggest, so the claim is the narrow one that added pool
structure buys little \emph{predictive} skill here, not that a one- to
three-pool model is a sufficient description of the system. Second, the added structural detail of the Yasso models is largely
\emph{unrewarded} by these data: their extra degrees of freedom flow into
prior-pinned, ridge-correlated transfer fractions that the two SOC campaigns
cannot resolve, and Yasso20 --- the most complex member, with all twelve
fractions free --- is the weakest predictor and the one most dependent on
input rescaling to escape its baseline bias. We read this as a light caution
rather than a verdict: the complexity may simply exceed what two SOC
observations per plot can constrain, and the conclusions here are drawn from
the \emph{ensemble} rather than from any single model. Pool-resolved
measurements would be required to test whether the Yasso partitioning carries
predictive value that the present data cannot expose.

### The over-prediction is a property of the likelihood\label{sec:biasdisc}

All six calibrated models over-predict slightly (§\ref{sec:residuals}), reversing
the baseline under-prediction. This is not evidence that Finnish soils hold less
carbon than the models imply --- it is a direct consequence of the
multiplicative-normal error model. With the observation standard deviation
scaling on the prediction, under-predicting a high-SOC plot incurs a far larger
\emph{relative} penalty than over-predicting it by the same amount (the penalty
on a fractional error $\delta$ behaves as $\delta^2/(1+\delta)^2$: unbounded as
$\delta\to-1$, bounded as $\delta\to\infty$). The calibration therefore hedges
upward, and the more so the less between-plot variance the model can explain ---
which is why the bias tracks $R^2$ inversely and is largest for the lowest-skill
models. Re-scoring confirms the offset is a near-uniform $\approx 10$--$16\%$
multiplicative shift; a symmetric log-error model, which carries no such
asymmetry, would re-centre the log-residuals at zero and remove essentially all
of it. Crucially, the bias is the \emph{only} headline sensitive to the
likelihood choice: the skill ranking, the fraction non-identifiability, and the
input-dominance result are governed by the information content of two SOC
observations per plot, not by the error-model shape, and are invariant to it. We
therefore report the bias as a modelling-choice consequence and read the
structural conclusions as robust to it.

### Forcing sensitivity and the resolved TP3 oscillation\label{sec:tp3osc}

Earlier TP3 runs (and, mildly, TP2) showed an interannual ringing in mean SOC,
absent in SP1 and the Yasso family. Two distinct effects are superimposed here,
and separating them matters. The first was a genuine \emph{numerical} artefact.
TP3 was the \emph{only} model integrated by an explicit annual Euler step, and
when a pool turns over faster than the step the Euler update factor
$(1-\alpha_A\xi)$ turns negative, so the pool overshoots its equilibrium each
year. At the posterior median this inflated the trajectory roughness by roughly
one to three orders of magnitude (mean $|\Delta^2\mathrm{SOC}|$ up to
$\sim\!10^{3}$--$10^{4}$ tC\,ha$^{-1}$ on individual plots). Replacing TP3's
integrator with the exact matrix-exponential step now used throughout the
ensemble (§\ref{models-tp3}) removes that artefact and cuts trajectory roughness
by roughly $5$--$700\times$; under constant forcing the exact step relaxes
monotonically to the equilibrium the Euler step overshoots.

The exact integrator does \emph{not}, however, make the TP3 trajectory smooth,
and it should not. A bounded interannual oscillation remains, and it is
\emph{physical}. TP3 reaches the observed stocks not by slow turnover but by
holding carbon in a fast active pool fed by heavily inflated inputs
($\sigma_{\mathrm{input}}\approx 13$--$20$; §\ref{sec:identifiability}), which
pushes the active pool's residence time below one year: at the posterior median
$\alpha_A\xi$ has median $\approx 1.4$--$2.0$ and exceeds~1 in
$\sim\!95$--$100\%$ of years, more in warm years. A pool that turns over faster
than the annual step re-equilibrates to each year's climate
($A_{\mathrm{ss}}=J/\alpha_A\xi$, small in warm years, large in cold), so total
SOC genuinely tracks interannual climate. This residual oscillation is
\emph{not} removable by finer time resolution: re-integrating the same posterior
with twelve monthly sub-steps --- seasonal temperature reconstructed from
$\bar T$ and its amplitude --- reduces roughness by only $\sim\!10$--$20\%$ and
shifts mean SOC by $<2$ tC\,ha$^{-1}$, because the seasonal cycle is already
inside $\xi$ (the four-node quadrature of §\ref{models-yasso07}) and the residual
signal is year-to-year, not sub-annual.

The contrast with the other models is therefore \emph{structural}, not numerical
--- all six now integrate exactly. It is governed by two quantities: how fast the
active pool turns over ($\alpha_A\xi$ relative to~1) and what fraction of the
total stock that pool holds. TP2's active pool is comparably fast
($\alpha_A\xi\approx 1.6$) but holds only $\sim\!23\%$ of total SOC, the rest
buffered in humus, so its total barely ripples (roughness $\sim\!3.7$ vs TP3's
$\sim\!9.3$); TP3's fast pool holds a comparable share ($\sim\!26\%$) but its
three-pool cascade lets the fluctuation propagate through the intermediate pool
as well, and its higher stocks amplify the absolute swing. SP1 sits at the
opposite extreme: a single pool turning over on century timescales
($\alpha\xi\approx 0.003$) that simply accumulates litter --- perfectly smooth
and still rising through the window. The Yasso family parks most carbon in the
slow humified pool, so despite fast AWEN pools it never enters the sub-annual
stock regime and stays smooth. The genuine structural point is thus unchanged:
TP3's fast-pool, inflated-input solution is the same under-identification seen
statically as input-dominated information gain and as the elevated
invalid-proposal rate ($8$--$17\%$); the interannual oscillation is a
\emph{symptom} of that solution, correctly rendered, not a numerical defect.
The mean-SOC and increment figures below are the exact-integrator run; the TP3
oscillation they show is the physical residual, not the former Euler ringing.

### [PLACEHOLDER --- residual Yasso-specific over-accumulation]

\emph{To resolve.} After removing the likelihood-induced offset
(§\ref{sec:biasdisc}), test via a symmetric-lognormal refit whether any
Yasso-specific over-prediction survives --- e.g. transient over-filling of the
slow humified pool during the 1917--1985 pre-run. Current evidence bounds it to
small.

---

# Key design decisions

1. **Transient initialisation via a 68-year pre-run.** The 1985 SOC
   trajectory is inconsistent with a steady-state initial condition under
   flat litter inputs. The pre-run from 1917 gives all models a calibrated,
   physically-grounded initial state. $\sigma_{\mathrm{init}}$ is the key
   parameter encoding pre-1985 productivity history; its posterior is
   directly interpretable as evidence for or against a post-disturbance
   recovery narrative.

2. **Uniform pre-run protocol and exact integration across all six models.**
   SP1/TP2/TP3 use exact analytical per-step solutions (closed form for one and
   two pools; closed-form matrix exponential for the three-pool cascade) and
   Yasso07/15/20 use the Fortran matrix-exponential solver with `C_final`
   chaining. Every model thus integrates its linear pool system exactly, so the
   inter-model differences are structural, not numerical; and the pre-run
   produces the same conceptual quantity (initial state at 1985) for all.

3. **Six models spanning two design axes.** Pool complexity (SP1 $\to$
   TP2 $\to$ TP3 $\to$ Yasso07) and climate integration (Yasso07 $\to$
   Yasso15 $\to$ Yasso20) vary independently. Conclusions about the
   saturation signal are drawn from the ensemble, not from individual models.

4. **Multiplicative-normal likelihood.** The standard deviation of the
   observation error scales with the prediction, representing proportional
   rather than absolute measurement uncertainty. This is appropriate for
   SOC stocks, where measurement noise and spatial representativeness
   both scale with stock magnitude.

5. **Fixed $\sigma_{\mathrm{obs}}$** at the empirical CV of the SOC
   observations, identical across all six models. Prevents structurally
   inadequate models from absorbing their residual variance into the error
   term; forces differences into the residuals where the intercomparison
   can examine them.

6. **$\xi(\bar{\mathrm{climate}})$ rather than $\overline{\xi(\mathrm{climate}_t)}$**
   for the pre-run climate. Jensen's inequality correction; 5--15\% bias
   in boreal conditions otherwise. Implemented uniformly via
   `compute_xi_mean_*` functions.

7. **Steady-state climate window = first 20 observed years.** The 1985--2004
   climate is used to compute the mean $\xi$ for the pre-run. This
   pre-dates the accelerated boreal warming of the 2010s and is
   representative of the pre-run period.

8. **Decomposition rates fixed at published defaults** for the Yasso
   family ($\alpha_A, \alpha_W, \alpha_E, \alpha_N, p_H, \alpha_H$).
   Reduces free-parameter count and keeps the intercomparison clean.
   SP1, TP2, and TP3 have no published global calibrations; all their
   parameters are free.

9. **All 12 transfer fractions free in all three Yasso models.** The
   Viskari (2022) structural zeros and derived fractions in the original
   Yasso20 are not imposed, because they would give Yasso20 fewer degrees
   of freedom than Yasso07/15 and make the structural intercomparison
   uninterpretable.

10. **Stick-breaking parameterisation with budget $B = 1 - p_H$.**
    Guarantees the per-column mass-budget constraint exactly; the remainder
    of each pool's outflow is respired as CO$_2$.

11. **Common weakly informative prior on transfer fractions**
    ($\sigma_{\mathrm{ppm}} = 0.4$, identical across models, in unconstrained
    logit space). Finnish data drives pool-flow structure --- the primary
    structural-comparison axis --- and an identical width removes any
    model-to-model asymmetry in how freely the flows are explored.

12. **Informative priors on climate and size parameters** (Yasso15/20
    from FMI posterior SDs; Yasso07 from the Tuomi 2009/2011 published
    posterior limits, read as $1\sigma$). Prevents the
    $\gamma$\,/\,$\sigma_{\mathrm{input}}$ compensation loop and the
    $\delta_2$ NaN pathology, and removes the $\beta_2$ climate-modifier
    detonation that the earlier hand-set width caused.

13. **Three SOC observation years (1985, 2006, 2024).** The Komeetta
    campaign (2024) provides a third calibration target. The
    1985--2006 interval constrains the accumulation dynamics; the
    2006--2024 near-stability constrains long-term decomposition rates
    and the saturation behaviour. Full Komeetta data description in the
    data documentation.

14. **SP1, TP2, and TP3 share the Yasso07 $\xi$ formula.** Climate
    response comparison across the pool-complexity axis is not confounded
    by approximation-scheme differences.

15. **ICBM topology for TP2.** All litter enters $A$; $H$ receives only
    transferred mass. Eliminates a weakly-constrained input-partitioning
    parameter and is standard in Nordic forestry SOC models.

16. **Sequential cascade for TP3.** No back-transfers between pools;
    A $\to$ S $\to$ H only. This is the minimal three-pool generalisation
    of TP2, extending the complexity ladder by one step before the full
    AWEN topology of Yasso07.

17. **Global $\sigma_{\mathrm{input}}$.** A single multiplicative litter
    scaling per model run. Per-species or per-region stratification is
    a possible extension but is not used in the current pipeline.

18. **No physical-ceiling rejection** in the likelihood. Implausible
    predictions are penalised via the log-likelihood, not a hard cap.
    Consistent with Viskari (2022).

---

# Appendix: prior and posterior marginals\label{sec:marginals}

For every free parameter in every model, the panels below overlay the
posterior (solid) on its prior (dashed), both in physical (back-transformed)
units. They are the visual counterpart of the identifiability discussion
(§\ref{sec:identifiability}): where the data inform a parameter, the posterior
contracts and shifts away from the prior; where they do not, the posterior sits
on top of the prior. The transfer fractions of the Yasso models are the clearest
example of the latter --- posterior $\approx$ prior, the signature of
non-identifiability --- while the auxiliary turnover and climate parameters show
the strongest contraction. Each panel's $x$-axis is clipped to the union of the
posterior and prior central ranges so that heavy prior tails do not compress the
posterior to an unreadable spike.

## SP1

```{r fig-marg-sp1, fig.cap="SP1: prior (dashed) vs posterior (solid) for all six free parameters. \\label{fig:margsp1}", out.width="92%", fig.align="center", fig.pos="H"}
show_fig(latest(file.path(diag_dir, "SP1"), "SP1_marginals_highlight_[0-9].*\\.png"))
```

## TP2

```{r fig-marg-tp2a, fig.cap="TP2 transfer/humification fractions: prior vs posterior. \\label{fig:margtp2a}", out.width="78%", fig.align="center", fig.pos="H"}
show_fig(latest(file.path(diag_dir, "TP2"), "TP2_marginals_fractions_[0-9].*\\.png"))
```

```{r fig-marg-tp2b, fig.cap="TP2 climate and auxiliary parameters: prior vs posterior. \\label{fig:margtp2b}", out.width="92%", fig.align="center", fig.pos="H"}
show_fig(latest(file.path(diag_dir, "TP2"), "TP2_marginals_climate_nuisance_[0-9].*\\.png"))
```

## TP3

```{r fig-marg-tp3a, fig.cap="TP3 transfer fractions ($p_S$, $p_H$): prior vs posterior. \\label{fig:margtp3a}", out.width="78%", fig.align="center", fig.pos="H"}
show_fig(latest(file.path(diag_dir, "TP3"), "TP3_marginals_fractions_[0-9].*\\.png"))
```

```{r fig-marg-tp3b, fig.cap="TP3 climate and auxiliary parameters: prior vs posterior. \\label{fig:margtp3b}", out.width="92%", fig.align="center", fig.pos="H"}
show_fig(latest(file.path(diag_dir, "TP3"), "TP3_marginals_climate_nuisance_[0-9].*\\.png"))
```

## Yasso07

```{r fig-marg-y07a, fig.cap="Yasso07 twelve transfer fractions: posterior $\\approx$ prior throughout --- the non-identifiability signature. \\label{fig:margy07a}", out.width="100%", fig.align="center", fig.pos="H"}
show_fig(latest(file.path(diag_dir, "Yasso07"), "Yasso07_marginals_fractions_[0-9].*\\.png"))
```

```{r fig-marg-y07b, fig.cap="Yasso07 climate, size and auxiliary parameters: prior vs posterior. \\label{fig:margy07b}", out.width="100%", fig.align="center", fig.pos="H"}
show_fig(latest(file.path(diag_dir, "Yasso07"), "Yasso07_marginals_climate_nuisance_[0-9].*\\.png"))
```

## Yasso15

```{r fig-marg-y15a, fig.cap="Yasso15 twelve transfer fractions: prior vs posterior. \\label{fig:margy15a}", out.width="100%", fig.align="center", fig.pos="H"}
show_fig(latest(file.path(diag_dir, "Yasso15"), "Yasso15_marginals_fractions_[0-9].*\\.png"))
```

```{r fig-marg-y15b, fig.cap="Yasso15 climate, size and auxiliary parameters: prior vs posterior. \\label{fig:margy15b}", out.width="100%", fig.align="center", fig.pos="H"}
show_fig(latest(file.path(diag_dir, "Yasso15"), "Yasso15_marginals_climate_nuisance_[0-9].*\\.png"))
```

## Yasso20

```{r fig-marg-y20a, fig.cap="Yasso20 twelve transfer fractions: prior vs posterior. Despite all twelve being free, none is appreciably informed by the data. \\label{fig:margy20a}", out.width="100%", fig.align="center", fig.pos="H"}
show_fig(latest(file.path(diag_dir, "Yasso20"), "Yasso20_marginals_fractions_[0-9].*\\.png"))
```

```{r fig-marg-y20b, fig.cap="Yasso20 climate, size and auxiliary parameters: prior vs posterior. \\label{fig:margy20b}", out.width="100%", fig.align="center", fig.pos="H"}
show_fig(latest(file.path(diag_dir, "Yasso20"), "Yasso20_marginals_climate_nuisance_[0-9].*\\.png"))
```

---

# Software

| Component | Details |
|-----------|---------|
| R | $\geq$ 4.3 |
| `BayesianTools` | Hartig et al. (2023); DEzs sampler |
| `coda` | Gelman--Rubin $\hat{R}$, ESS |
| `compiler` | `cmpfun()` byte-compilation of likelihood |
| `parallel` | `mclapply()` per-plot parallelism |
| `ranger` | Random-forest residual analysis |
| Fortran 90 cores | `R CMD SHLIB`; `.Fortran()` interface (Yasso family) |
| SP1, TP2, TP3 | Pure R; no Fortran |

Random seed: `set.seed(2025)` before chain initialisation, consistent
across all models. Test plot subsampling (when `N_PLOTS_TEST` is set):
`set.seed(42)`.

---

# References

* Andrén, O. \& Kätterer, T. (1997). ICBM: The introductory carbon
  balance model for exploration of soil carbon balances.
  \textit{Ecological Applications} 7, 1226--1236.
* Hartig, F. et al. (2023). BayesianTools: General-purpose MCMC and SMC
  samplers and tools for Bayesian statistics. R package version 0.1.8.
* Tuomi, M. et al. (2008). Leaf litter decomposition -- estimates of
  global variability based on Yasso07 model. \textit{Ecological Modelling}
  220, 3362--3371.
* Tuomi, M. et al. (2009). Heterotrophic soil respiration --- comparison
  of different models on data from litterbag microenvironments.
  \textit{Agriculture, Ecosystems and Environment} 132, 209--217.
* Tuomi, M. et al. (2011). Soil carbon model Yasso07 graphical user
  interface. \textit{Environmental Modelling \& Software} 26, 1358--1362.
* Tupek, B. et al. Litter input dataset for Finnish NFI plots.
  Zenodo. \url{https://doi.org/10.5281/zenodo.19736499}.
* Viskari, T. et al. (2022). Calibrating the soil organic carbon model
  Yasso20 with multiple Finnish datasets. \textit{Geoscientific Model
  Development} 15, 1735--1752.
