---
title: "A pool-level mass-balance inconsistency in the published Yasso20 parameter set"
subtitle: "Working note for discussion with the Yasso development team (FMI)"
author: "Lorenzo Menichetti (HIKET project)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: false
    number_sections: true
    latex_engine: xelatex
    extra_dependencies: ["booktabs", "amsmath", "array"]
geometry: margin=2.5cm
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# All paths relative to this file's directory (documentation/)
FMI_RDA  <- "../../Model_functions_real_data_transient/Decomposition_functions/Yasso_original/Yasso20_sample_parameters.rda"
FMI_DAT  <- "../../Model_functions_real_data_transient/Decomposition_functions/Yasso_original/ParY20.dat"
```

# Summary

Using the published Yasso20 model code and the published Yasso20 calibration
parameters --- both taken directly from the FMI/Ryassofortran distribution --- the
inter-pool transfer matrix violates carbon conservation at the level of two individual
donor pools. For the acid-soluble pool **A** and the non-soluble pool **N**, the fractions
of decomposed carbon redistributed to other pools plus the humification fraction sum
to **1.0042**, implying a respired (CO$_2$) fraction of **$-$0.0042** for each.

Cross-checking with Viskari et al. (2022) confirms that this arises from two deliberate
calibration decisions made independently, without enforcing a joint per-column budget
constraint. Every numerical result in this note is computed live from the FMI source
files at document-render time.

---

# Sources

All inputs trace directly to FMI sources. No HIKET code or re-derivations are used.

| Role | Source |
|------|--------|
| Calibration methodology, model equations | Viskari et al. (2022), *Geosci.\ Model Dev.* **15**, 1735--1752, doi:10.5194/gmd-15-1735-2022 |
| Parameter values (named) | `Yasso20_sample_parameters.rda`, FMI/Ryassofortran repository |
| Parameter values (raw index order) | `ParY20.dat`, same repository |
| Model code (R) | `Yasso20.r`, copyright 2020 FMI / Janne Pusa, GPL-3.0, same repository |
| Model code (Fortran) | `yassofortran.f90`, subroutine `mod5c20`, same copyright |

The two parameter files are numerically identical (verified below).

---

# The published parameters

```{r load-params}
# Load from the FMI .rda file
e <- new.env()
load(FMI_RDA, envir = e)
theta <- e$sample_parameters   # named numeric vector, length 35

# Cross-check against the raw ParY20.dat
raw_dat <- scan(FMI_DAT, quiet = TRUE)
stopifnot(length(raw_dat) == 35)
stopifnot(all(raw_dat == as.numeric(theta)))
message("ParY20.dat and Yasso20_sample_parameters.rda are identical: verified.")
```

```{r param-table, echo=FALSE}
# Parameters relevant to mass balance (transfer fractions + p_H)
# Indices and names follow the rda file; meaning from Viskari et al. Table 1
frac_idx  <- c(5:16, 31)
frac_name <- c("pWA","pEA","pNA","pAW","pEW","pNW",
                "pAE","pWE","pNE","pAN","pWN","pEN","pH")
frac_mean <- c("W→A","E→A","N→A","A→W","E→W","N→W",
               "A→E","W→E","N→E","A→N","W→N","E→N","AWEN→H")
frac_val  <- as.numeric(theta[frac_idx])

df_params <- data.frame(
  Index   = frac_idx,
  Name    = frac_name,
  Meaning = frac_mean,
  Value   = frac_val
)

knitr::kable(df_params, booktabs = TRUE, linesep = "",
             caption = "Transfer fractions and humification fraction from Yasso20\\_sample\\_parameters.rda. Highlighted rows are the source of the budget violation.",
             col.names = c("idx", "rda name", "Meaning", "Value"))
```

---

# The model structure

The transfer matrix is built in `Yasso20.r` exactly as follows (copied verbatim;
indices refer to the FMI parameter vector `theta`):

```{r build-Ap}
# Verbatim from FMI Yasso20.r (Janne Pusa, FMI, GPL-3.0)
row1 <- c(-1,         theta[5],  theta[6],  theta[7],  0)
row2 <- c(theta[8],  -1,         theta[9],  theta[10], 0)
row3 <- c(theta[11], theta[12], -1,         theta[13], 0)
row4 <- c(theta[14], theta[15], theta[16], -1,         0)
row5 <- c(theta[31], theta[31], theta[31], theta[31], -1)
A_p  <- matrix(c(row1, row2, row3, row4, row5), 5, 5, byrow = TRUE)
colnames(A_p) <- rownames(A_p) <- c("A","W","E","N","H")
```

The Fortran subroutine `mod5c20` in `yassofortran.f90` builds the same matrix
by equivalent index-based assignments (e.g.\ `A(2,1) = theta(8)*ABS(A(1,1))`).
Both implementations agree exactly.

# The conservation criterion

For a donor-controlled linear compartment model $d\mathbf{x}/dt = M(\theta,c)\,\mathbf{x} +
b(t)$ (Viskari et al. 2022, Eq.\ 1), carbon is conserved at every pool if and only if each
column of the structural matrix $F(\theta)$ sums to $\leq 0$, equivalently:

$$\sum_{i \neq j} p_{ij} + p_H \;\leq\; 1 \qquad \forall\, j \in \{A,W,E,N\}$$

where the sum is over all fractions leaving donor pool $j$ to the other AWEN pools,
and $p_H$ is the humification fraction (same for all donors, row 5 of $F$). The
remainder $1 - \sum p_{ij} - p_H$ is the fraction respired as CO$_2$. A negative value
implies carbon creation at that pool.

---

# Result

```{r colsums}
cs <- colSums(A_p)
round(cs, 6)
```

```{r budget-table, echo=FALSE}
pools <- c("A","W","E","N")
transfers <- c(
  theta[8]  + theta[11] + theta[14],   # out of A: pAW + pAE + pAN
  theta[5]  + theta[12] + theta[15],   # out of W: pWA + pWE + pWN
  theta[6]  + theta[9]  + theta[16],   # out of E: pEA + pEW + pEN
  theta[7]  + theta[10] + theta[13]    # out of N: pNA + pNW + pNE
)
pH_val   <- as.numeric(theta[31])
redis    <- transfers + pH_val
respired <- 1 - redis

df_budget <- data.frame(
  Pool       = pools,
  Transfers  = round(transfers, 4),
  pH         = pH_val,
  Total      = round(redis, 4),
  Respired   = round(respired, 4)
)

knitr::kable(df_budget, booktabs = TRUE, linesep = "",
             caption = "Per-donor-pool carbon budget (FMI parameters). Redistributed total = transfers + p\\_H. A negative Respired fraction indicates carbon creation.",
             col.names = c("Donor pool", "Σ transfers out",
                           "+ p_H", "Redistributed total", "Respired fraction"))
```

Columns **A** and **N** have a redistributed total of
`r round(redis[1], 4)`,
implying a respired fraction of
`r round(respired[1], 4)`:
a pool-level violation of carbon conservation.

---

# Cause: two independent calibration decisions without a joint constraint

The origin is precisely documented in Viskari et al. (2022). Two passages, read
together, explain the inconsistency.

**Decision 1 --- p_AW and p_NA fixed at 1 (Sect. 2.3, p. 1741):**

> *"if the estimated parameter values for the p terms in Eq. (3) were within
> three decimals from either 0 or 1, they were set to the nearest limit value
> of 0 or 1, after which the calibration was redone \ldots Parameters p_AW and
> p_NA were set to 1 and the other AWEN-related p values were set to 0."*

During the iterative calibration protocol, p_AW and p_NA converged to within 0.001 of
1.0 and were deliberately fixed at exactly **1** for the final calibration run.

**Decision 2 --- p_H calibrated independently from a separate dataset (Sect. 2.2, p. 1739):**

> *"p_H (transfer fraction from AWEN pools to pool H) could have non-zero values
> only with the Oak Ridge global SOC dataset."*

The humification fraction p_H was estimated exclusively from the Oak Ridge global SOC
measurements --- a dataset entirely separate from the litter-bag data that drove the
AWEN transfer fractions. The MAP value is p_H = `r theta["pH"]`.

**The combination:** once p_AW and p_NA were fixed at 1, the subsequent independent
calibration of p_H from different data had no mechanism to enforce the joint constraint
p_AW + p_H $\leq$ 1. The excess is exactly p_H:

```{r excess}
cat(sprintf("p_AW + p_H = %.4f + %.4f = %.4f  (excess = %.4f)\n",
            as.numeric(theta["pAW"]), as.numeric(theta["pH"]),
            as.numeric(theta["pAW"]) + as.numeric(theta["pH"]),
            as.numeric(theta["pAW"]) + as.numeric(theta["pH"]) - 1))
```

# The paper's stated design intent

The paper makes the design intent explicit (Sect. 2.3, p. 1741):

> *"since we assumed that only decomposition in the W pool results in CO$_2$,
> we estimated only p_EW and set p_EA to be the E pool remnant from 1 with
> p_EN set to 0."*

Under this assumption pools A, E, and N pass all decomposed carbon onward; only W
respires. This is a biologically defensible modelling choice. The algebraically
consistent implementation of this assumption requires, for any donor pool $j$ where
all outflow fractions except one are zero:

$$p_{jk} = 1 - p_H$$

so that the column sum equals exactly 1. With p_H = `r theta["pH"]`, the
conserving values would be:

```{r conserving-values}
pH <- as.numeric(theta["pH"])
cat(sprintf("p_AW* = 1 - p_H = 1 - %.4f = %.4f  (stored: %.4f)\n",
            pH, 1-pH, as.numeric(theta["pAW"])))
cat(sprintf("p_NA* = 1 - p_H = 1 - %.4f = %.4f  (stored: %.4f)\n",
            pH, 1-pH, as.numeric(theta["pNA"])))
```

The difference between the stored value (1.0000) and the conserving value
(`r round(1-pH, 4)`) is exactly p_H = `r pH` --- the amount by which each
column exceeds the budget.

---

# Why it was not detected

The violation is **local** (confined to pools A and N) and small
(`r round(respired[1]*-1*100, 2)` % excess). In a full simulation the carbon
erroneously retained in A and N flows onward to W, which strongly respires
(`r round(respired[2]*100, 1)` % of its outflow). The system as a whole
remains numerically stable: a total-stock run with zero litter input and unit
carbon in any AWEN pool shows declining total carbon throughout, because
downstream respiration from W masks the local source. The inconsistency is
only apparent from the column-level budget check shown above.

---

# Possible resolution

The joint constraint can be restored without altering the model's structural
intent by replacing the fixed limit values with the algebraically consistent
expressions:

$$p_{AW}^* = 1 - p_H \qquad p_{NA}^* = 1 - p_H$$

This sets the respired fraction of columns A and N to exactly zero, preserving
the design assumption that "only W respires." The numerical change is
`r round(pH*100, 2)` % --- negligible for simulation results. The same logic
applies to any future calibration where a transfer fraction converges to 1
while p_H is calibrated independently.

We would welcome the development team's view on whether this is the intended
resolution, and whether the distributed parameter files should be updated.

---

# Appendix: full parameter table

```{r full-param-table, echo=FALSE}
all_names <- c(
  "aA","aW","aE","aN",
  "pWA","pEA","pNA","pAW","pEW","pNW","pAE","pWE","pNE","pAN","pWN","pEN",
  "w1","w2","w3","w4","w5",
  "b1","b2","bN1","bN2","bH1","bH2","g","gN","gH",
  "pH","aH","th1","th2","r"
)
df_all <- data.frame(
  Index = seq_along(all_names),
  Name  = all_names,
  Value = round(as.numeric(theta), 6)
)
knitr::kable(df_all, booktabs = TRUE, linesep = "",
             caption = "Complete Yasso20 MAP parameter vector from Yasso20\\_sample\\_parameters.rda (identical to ParY20.dat).",
             col.names = c("Index", "rda name", "Value"))
```

---

# Reference

Viskari, T., Pusa, J., Fer, I., Repo, A., Vira, J., and Liski, J. (2022).
Calibrating the soil organic carbon model Yasso20 with multiple datasets.
*Geoscientific Model Development*, **15**, 1735--1752.
https://doi.org/10.5194/gmd-15-1735-2022
