---
title: "Modular SOC Decomposition Framework"
subtitle: "Architecture Documentation for Multi-Model Intercomparison"
author: "Lorenzo — HIKET Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{tikz}
  - \usetikzlibrary{arrows.meta, positioning, calc, fit, backgrounds, shapes.geometric, decorations.pathreplacing}
  - \definecolor{inputblue}{HTML}{3B82F6}
  - \definecolor{compatgreen}{HTML}{10B981}
  - \definecolor{decompred}{HTML}{EF4444}
  - \definecolor{modifiergold}{HTML}{F59E0B}
  - \definecolor{outputpurple}{HTML}{8B5CF6}
  - \definecolor{bglight}{HTML}{F8FAFC}
  - \definecolor{bordergray}{HTML}{CBD5E1}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

\vspace{1em}

# 1. Overview

This document defines a modular architecture for soil organic carbon (SOC) decomposition modelling within the HIKET multi-model intercomparison framework. The goal is to define all models — Yasso07, Yasso15, Yasso20, RothC, and Q-model — through common functional interfaces that cleanly separates **litter input processing**, **environmental modification of kinetics**, and **decomposition dynamics**.

This separation enables modularity.

\newpage

# 2. General Framework

## 2.1 Core Equation

The simulated soil organic carbon over time is expressed as:

$$
\mathbf{C}(t, k) = f_{d_k}\!\Big(\, f_{i_k}\!\big(\mathbf{I}(t)\big),\;\boldsymbol{\xi_k}(t) \Big)
$$

where:

| Symbol | Description | Dimensions |
|:-------|:-----------|:-----------|
| $\mathbf{C}(t)$ | SOC pool vector at time $t$ | $[n_{\text{pools}} \times 1]$ |
| $\mathbf{I}(t)$ | Raw litter input data at time $t$ | model-agnostic format |
| $f_i(\cdot)$ | **Input compatibility layer** — transforms raw inputs into model-specific pool inputs | $\mathbf{I}(t) \to \mathbf{J}(t)$ |
| $f_d(\cdot)$ | **Decomposition function** — the structural core of each model | $(\mathbf{J}(t), \boldsymbol{\xi}(t)) \to \mathbf{C}(t)$ |
| $\boldsymbol{\xi}(t)$ | **Kinetic modifier** — composite environmental scalar(s) | $\in \mathbb{R}^{n_{\xi}}$ |

The $k$ denotes model-specific functions.

The kinetic modifier is itself composed of independent environmental response functions:

$$
{\xi}(t) = \xi_T\!\big(T(t)\big) \;\cdot\; \xi_M\!\big(M(t)\big) \;\cdot\;  \xi_e
$$

where $\xi_T$ is the temperature modifier, $\xi_M$ is the moisture/precipitation modifier, and $\xi_e$ represents any additional static or time-varying modifiers, likely edaphic (e.g., clay content, soil cover).

## 2.2 Architecture Schema

\begin{figure}[H]
\centering
\begin{tikzpicture}[
    >=Stealth,
    node distance=1.2cm and 2.0cm,
    box/.style={
        rectangle, rounded corners=4pt, draw=#1, fill=#1!8,
        line width=1pt, minimum height=1.1cm, minimum width=3.2cm,
        text centered, font=\small\sffamily
    },
    smallbox/.style={
        rectangle, rounded corners=3pt, draw=#1, fill=#1!8,
        line width=0.8pt, minimum height=0.8cm, minimum width=2.2cm,
        text centered, font=\footnotesize\sffamily
    },
    tinybox/.style={
        rectangle, rounded corners=2pt, draw=#1, fill=#1!8,
        line width=0.6pt, minimum height=0.6cm, minimum width=1.6cm,
        text centered, font=\scriptsize\sffamily
    },
    data/.style={
        rectangle, rounded corners=2pt, draw=bordergray, fill=bglight,
        line width=0.6pt, minimum height=0.7cm, minimum width=2.0cm,
        text centered, font=\footnotesize\ttfamily
    },
    arr/.style={->, line width=1pt, color=#1},
    darr/.style={->, line width=0.8pt, dashed, color=#1},
    brace/.style={decorate, decoration={brace, amplitude=6pt, raise=2pt}},
    note/.style={font=\scriptsize\sffamily\itshape, text=black!60}
]

% =============================================
% RAW INPUTS (left)
% =============================================
\node[data] (I_litter) {Litter masses by component};
\node[data, below=0.35cm of I_litter] (I_species) {Species composition};

% Brace for inputs
\draw[brace, bordergray, line width=0.8pt]
    (I_litter.north east) -- (I_species.south east)
    node[midway, right=12pt, font=\small\sffamily, inputblue] {$\mathbf{I}(t)$};

% =============================================
% INPUT COMPATIBILITY LAYER
% =============================================
\node[box=compatgreen, right=1.8cm of I_litter, yshift=-0.35cm] (fi) {%
    \begin{tabular}{c}
    I compatibility \\  Layer \\ $f_{i_k}\!\big(\mathbf{I}(t)\big)$
    \end{tabular}};

% =============================================
% DECOMPOSITION FUNCTION
% =============================================
\node[box=decompred, right=2.5cm of fi] (fd) {%
    \begin{tabular}{c}
    Decomposition \\ $f_{d_k}\!\big(\mathbf{J}(t),\, \boldsymbol{\xi}(t)\big)$
    \end{tabular}};

% Hard-coded note inside fd
\node[note, below=0.15cm of fd] (fd_note) {incl.\ hard-coded size modifiers};

% =============================================
% OUTPUT
% =============================================
\node[box=outputpurple, below=2.2cm of fd] (C) {%
    \begin{tabular}{c}
    total SOC \\ $\mathbf{C}(t)$
    \end{tabular}};

% =============================================
% ENVIRONMENTAL DRIVERS (top row)
% =============================================
\node[data, above=5.2cm of fi, xshift=-0.5cm] (T) {$T_{\text{monthly}}(t)$};
\node[data, right=1.8cm of T] (M) {$M_{\text{monthly}}(t)$};
\node[data, right=1.8cm of M] (clay) {clay \%};

% =============================================
% TEMPERATURE BRANCH
% =============================================
% Temporal aggregation
\node[tinybox=bordergray, below=0.5cm of T] (T_agg) {%
    \begin{tabular}{c}
    $\overline{T}$ or $T_m$
    \end{tabular}};
\node[note, left=0.1cm of T_agg] {avg/pass};

% Temperature modifier
\node[smallbox=modifiergold, below=0.5cm of T_agg] (xiT) {$\xi_{T_k}\!\big(T\big)$};

% =============================================
% MOISTURE BRANCH
% =============================================
% Moisture compatibility layer
\node[tinybox=compatgreen, below=0.5cm of M] (fm) {%
    \begin{tabular}{c}
    $f_{m_k}$
    \end{tabular}};
\node[note, right=0.1cm of fm, text width=2cm, align=left] {$P$ or TSMD\\or $\emptyset$};

% Moisture modifier
\node[smallbox=modifiergold, below=0.5cm of fm] (xiM) {$\xi_{M_k}\!\big(f_{m_k}\big)$};

% =============================================
% EDAPHIC BRANCH
% =============================================
\node[smallbox=modifiergold, below=1.55cm of clay] (xiE) {$\xi_e$};
\node[note, right=0.1cm of xiE] {RothC only};

% =============================================
% COMPOSITE MODIFIER
% =============================================
\node[box=modifiergold, below=0.8cm of xiM] (xi) {%
    \begin{tabular}{c}
    Kinetic Modifier \\ $\boldsymbol{\xi}_k(t) = \xi_T \cdot \xi_M \cdot \xi_e$
    \end{tabular}};

% =============================================
% ARROWS — Inputs
% =============================================
\draw[arr=inputblue] (I_litter.east) -- ++(0.8,0) |- (fi.west);
\draw[arr=inputblue] (I_species.east) -- ++(0.8,0) |- (fi.west);

% fi to fd
\draw[arr=compatgreen] (fi.east) -- (fd.west)
    node[midway, below=2pt, font=\footnotesize\sffamily, compatgreen] {$\mathbf{J}(t)$};

% fd to output
\draw[arr=decompred] (fd.east) -- ++(1.2,0) |- (C.east);

% =============================================
% ARROWS — Temperature branch
% =============================================
\draw[arr=bordergray] (T.south) -- (T_agg.north);
\draw[arr=modifiergold] (T_agg.south) -- (xiT.north);

% =============================================
% ARROWS — Moisture branch
% =============================================
\draw[arr=bordergray] (M.south) -- (fm.north);
\draw[arr=modifiergold] (fm.south) -- (xiM.north);

% =============================================
% ARROWS — Edaphic branch
% =============================================
\draw[arr=modifiergold] (clay.south) -- (xiE.north);

% =============================================
% ARROWS — Sub-modifiers to composite
% =============================================
\draw[arr=modifiergold] (xiT.south) -- ++(0,-0.2) -| ([xshift=-6pt]xi.north);
\draw[arr=modifiergold] (xiM.south) -- (xi.north);
\draw[arr=modifiergold] (xiE.south) -- ++(0,-0.2) -| ([xshift=6pt]xi.north);

% =============================================
% ARROW — Composite modifier to fd
% =============================================
\draw[arr=modifiergold] (xi.south) -- ++(0,-0.35) -| (fd.north);

% =============================================
% BACKGROUND — Modifier group
% =============================================
\begin{scope}[on background layer]
    \node[fit=(T_agg)(xiT)(fm)(xiM)(xiE)(xi),
          fill=modifiergold!4, draw=modifiergold!25,
          rounded corners=6pt, inner sep=10pt] {};
\end{scope}

\end{tikzpicture}
\caption{Modular architecture of the SOC decomposition framework. Raw inputs $\mathbf{I}(t)$ are transformed by the input compatibility layer $f_i$ into model-specific inputs $\mathbf{J}(t)$. Environmental drivers are combined into the composite kinetic modifier $\boldsymbol{\xi}(t)$. Both feed into the model-specific decomposition function $f_d$.}
\label{fig:architecture}
\end{figure}

\newpage


# 3. Input requirements

## 3.1 Standardised input format — $\mathbf{I}(t)$
Dimensions $N_{plots} \times N_{time steps}$

The standard input data for all models $\mathbf{I}(t)$ contains litter mass inputs from six cohorts (foliage, fine roots, branches, coarse roots, stump, stem).
The species group (pine, spruce, birch) are partitioned based on species composition data, which are assumed constant per plot.

## 3.2 Standised environmental drivers

### 3.2.1 Time-variant drivers
Dimensions $N_{plots} \times N_{time steps}$

Temperature and moisture inputs are provided as monthly time series.
Temperature relies only on a compatibility layer to aggregate to annual summaries where needed. 
Moisture is more complicated since the models rely on different inputs (precipitation vs topsoil moisture deficit), so we need a compatibility layer to convert precipitation to soil moisture deficit for RothC, and to pass precipitation directly to Yasso. The Q-model does not explicitly include a moisture modifier, but implicitly via litter quality.

### 3.2.2 Static edaphic modifiers
Dimensions $N_{plots} \times 1$

Clay content is provided as a scalar per plot, as well as tree species composition.

# 3. Output structure
## 3.1 Model-specific output

Dimensions $N_{plots} \times N_{time steps} \times N_{pools}$, replicated for each model (different data frames for each model, with different pool names and numbers).

## 3.1 Total SOC output
Dimensions $N_{plots} \times N_{time steps} \times N_{models}$$



# 3. Components requirements

## 3.1 Input Compatibility Layer — $f_i\!\big(\mathbf{I}(t)\big)$


The input layer $f_i$ maps this standardised input into the model-specific input vector $\mathbf{J}(t)$:

$$
f_i: \mathbf{I}(t) \;\longrightarrow\; \mathbf{J}(t) \in \mathbb{R}^{n_{\text{input pools}}}
$$

The mapping differs across models:

| Model | Input pools $\mathbf{J}(t)$ | Key transformation |
|:------|:---------------------------|:-------------------|
| Yasso07/15/20 | AWEN fractions $(A, W, E, N)$ | Total mass to AWEN via component-specific coefficients |
| RothC | DPM, RPM, BIO, HUM, IOM | Total mass $\to$ DPM/RPM split via component-dependent coefficients |
| Q-model | Quality-distributed cohorts per tissue type | Separate cohorts retained; quality $q_0$ assigned per species and component |




## 3.2 Kinetic Modifier — $\boldsymbol{\xi}(t)$

The kinetic modifier captures all environmental effects on decomposition rates. It is composed multiplicatively:

$$
\boldsymbol{\xi}(t) = \xi_T\!\big(T(t)\big) \;\times\; \xi_M\!\big(f_{m_k}(M(t))\big) \;\times\;  \xi_k
$$

Each sub-modifier is a vector over time (or, for edaphic effects $\xi_k$, a scalar, constant over time) that rescales the baseline decomposition rates.

### 3.2.1 Temperature Modifier — $\xi_T$

All functions accept the same input, with the only compatibility layer being a montlhy to annual averaging.  

The functional form of the temperature response differs across models:

| Model | Form | Expression |
|:------|:-----|:-----------|
| Yasso07/15 | Gaussian-like, 4-point annual | $\xi_T = \frac{1}{4}\sum_{i=1}^{4}\exp\!\big(\beta_1 T_i + \beta_2 T_i^2\big)$, $T_i$ from annual mean + amplitude |
| Yasso20 | Gaussian-like, exact monthly | $\xi_T = \frac{1}{12}\sum_{m=1}^{12}\exp\!\big(\beta_1 T_m + \beta_2 T_m^2\big)$, $T_m$ are observed monthly values |
| RothC | Arrhenius-based | $\xi_T = \dfrac{47.9}{1 + \exp\!\big(\frac{106}{T+18.3}\big)}$ |
| Q-model | Linear | $\xi_T = \max\!\big(0,\; a_T + b_T \cdot T\big)$ |

### 3.2.2 Moisture/Precipitation Modifier — $\xi_M$

Moisture functions of different models accept different inputs, so we need another compatibility layer, $f_{m_k}$. Yasso uses precipitation, RothC uses topsoil moisture deficit, and Q-model does not explicitly include a moisture modifier (but implicitly via litter quality).

| Model | Driver | Notes |
|:------|:-------|:------|
| Yasso07/15 | Annual precipitation $P$ | $\xi_M = 1 - \exp\!\big(\gamma \cdot P/1000\big)$; pool-group-specific for Yasso15 |
| Yasso20 | Monthly precipitation (summed annually) | Same formula; $P$ derived internally as annual sum of monthly values |
| RothC | Topsoil moisture deficit | Rate-modifying factor from TSMD accumulation |
| Q-model | — | Not explicitly included (implicitly via litter quality) |

### 3.2.3 Additional Static Modifiers — $\xi_k$
| Modifier | Applicable models | Description |
|:---------|:-----------------|:------------|
| Clay ($\xi_{\text{clay}}$) | RothC | Modifies BIO/HUM partitioning: $x = 1.67 (1.85 + 1.60 \exp(-0.0786\, \%\text{clay}))$ |
| Soil cover | RothC | Binary modifier on decomposition rate for bare soil, WON'T BE IMPLEMENTED!|

### 3.2.4 Hard-coded Static Modifiers
Both Q and Yasso have static modifiers dependent on size that are specific to litter classes. These are be hard-coded in each model function, rather than passed as separate modifiers.  

For the Yasso family of models, all size modifiers are hard-coded in the R wrappers




# 4. Decomposition Equations

This section presents the complete decomposition dynamics for each model

## 4.1 Yasso07

### Pool structure

Five pools: $\mathbf{C} = [C_A,\, C_W,\, C_E,\, C_N,\, C_H]^T$, representing Acid-soluble, Water-soluble, Ethanol-soluble, Non-soluble (lignin-like), and Humus.

### Dynamics

$$
\frac{d\mathbf{C}}{dt} = \mathbf{A}\,\mathbf{C} + \mathbf{I}
$$

where $\mathbf{I} = [I_A,\, I_W,\, I_E,\, I_N,\, 0]^T$ (no direct input to Humus).

**Diagonal elements** (loss from pool $i$):
$$A_{ii} = -\alpha_i \cdot \xi_T \cdot \xi_M$$

where $\alpha_i$ is the base decomposition rate. For the N pool, an additional size modifier applies to woody litter:
$$A_{44} = -\alpha_N \cdot \xi_T \cdot \xi_M \cdot f_{\text{size}}$$

with $f_{\text{size}} = \min\!\big(1,\; \exp(-\delta_1 \cdot d^{\delta_2} \cdot |r|)\big)$, where $d$ is the wood diameter (cm).

**Off-diagonal elements** (transfer from pool $j$ to pool $i$):
$$A_{ij} = p_{ij} \cdot |\alpha_j \cdot \xi_T \cdot \xi_M|$$

**Humus formation** (row 5, columns 1--4):
$$A_{5j} = p_H \cdot |\alpha_j \cdot \xi_T \cdot \xi_M| \quad \text{for } j \in \{A, W, E, N\}$$

**Respiration** is the residual: $R_h = \sum_j |A_{jj}| C_j - \sum_{i \neq j} A_{ij} C_j$.

### Environmental modifiers

**Temperature** — four-point approximation of intra-annual variation:
$$\xi_T = \frac{1}{4} \sum_{i=1}^{4} \exp\!\big(\beta_1 T_i + \beta_2 T_i^2\big)$$
where the quarterly temperatures are derived from annual mean $\bar{T}$ and amplitude $A_T$:
\begin{align*}
T_1 &= \bar{T} + \tfrac{4A_T}{\pi}\!\left(\tfrac{1}{\sqrt{2}} - 1\right), &
T_2 &= \bar{T} - \tfrac{4A_T}{\sqrt{2}\,\pi} \\
T_3 &= \bar{T} + \tfrac{4A_T}{\pi}\!\left(1 - \tfrac{1}{\sqrt{2}}\right), &
T_4 &= \bar{T} + \tfrac{4A_T}{\sqrt{2}\,\pi}
\end{align*}

**Precipitation:**
$$\xi_M = 1 - \exp\!\left(\gamma \cdot \frac{P}{1000}\right)$$
where $P$ is annual precipitation (mm).

### Analytical solution

For annual timestep $\Delta t = 1$\,yr:
$$\mathbf{C}(t+1) = e^{\mathbf{A}} \cdot \mathbf{C}(t) + \left(e^{\mathbf{A}} - \mathbf{I}_5\right) \mathbf{A}^{-1} \mathbf{I}$$
where $e^{\mathbf{A}}$ is the matrix exponential.

**Steady state:**
$$\mathbf{C}^* = -\mathbf{A}^{-1}\,\mathbf{I}$$

### Default parameters

| Parameter | Value | Description |
|:----------|:------|:------------|
| $\alpha_A,\, \alpha_W,\, \alpha_E,\, \alpha_N$ | 0.66, 4.30, 0.35, 0.22 | Base decomposition rates (yr$^{-1}$) |
| $\alpha_H$ | 0.0033 | Humus decomposition rate (yr$^{-1}$) |
| $p_H$ | 0.04 | Fraction to Humus |
| $\beta_1,\, \beta_2$ | 0.096, $-$0.0014 | Temperature response parameters |
| $\gamma$ | $-$1.21 | Precipitation response parameter |
| $\delta_1,\, \delta_2,\, r$ | $-$1.7, 0.86, $-$0.306 | Size modifier parameters |

\newpage

## 4.2 Yasso15

Yasso15 retains the AWENH pool structure and transfer matrix of Yasso07, but introduces two structural changes.

**Pool-group-specific climate responses.** Where Yasso07 applies a single temperature and precipitation modifier to all pools, Yasso15 separates the climate response into three groups — AWE, N, and H — each with independent parameters:

$$\xi_{T}^{\text{AWE}} = \frac{1}{4}\sum_{i=1}^{4}\exp\!\big(\beta_1 T_i + \beta_2 T_i^2\big), \quad \xi_{T}^{N} = \frac{1}{4}\sum_{i=1}^{4}\exp\!\big(\beta_{N1} T_i + \beta_{N2} T_i^2\big), \quad \xi_{T}^{H} = \frac{1}{4}\sum_{i=1}^{4}\exp\!\big(\beta_{H1} T_i + \beta_{H2} T_i^2\big)$$

Similarly for precipitation:

$$\xi_{M}^{\text{AWE}} = 1 - \exp\!\big(\gamma \cdot P/1000\big), \qquad \xi_{M}^{N} = 1 - \exp\!\big(\gamma_N \cdot P/1000\big), \qquad \xi_{M}^{H} = 1 - \exp\!\big(\gamma_H \cdot P/1000\big)$$

The combined climate modifier is applied per pool group when constructing the diagonal of $\mathbf{A}$:

\begin{align*}
A_{ii} &= -\alpha_i \cdot \xi_T^{\text{AWE}} \cdot \xi_M^{\text{AWE}} \cdot f_{\text{size}} & \text{for } i \in \{A, W, E\} \\
A_{44} &= -\alpha_N \cdot \xi_T^{N} \cdot \xi_M^{N} \cdot f_{\text{size}} \\
A_{55} &= -\alpha_H \cdot \xi_T^{H} \cdot \xi_M^{H} & \text{(no size effect)}
\end{align*}

Off-diagonal elements and humus formation follow the same structure as Yasso07, with transfers proportional to the source pool's modified rate.

**Leaching.** Yasso15 adds a precipitation-dependent leaching term to the AWEN pools (not H):

$$A_{ii} \mathrel{+}= w \cdot P/1000 \quad \text{for } i \in \{A, W, E, N\}$$

The size modifier, transfer matrix structure, analytical solution (matrix exponential), and steady-state formula ($\mathbf{C}^* = -\mathbf{A}^{-1}\mathbf{b}$) are identical to Yasso07.

\newpage

## 4.3 Yasso20

Yasso20 shares the AWENH pool structure, pool-group-specific climate responses, and leaching of Yasso15. The structural equations are the same. The principal differences are in parameterisation and the size modifier formula.

**Modified size dependence.** Yasso07/15 use an exponential form:

$$f_{\text{size}}^{\text{07/15}} = \min\!\big(1,\; \exp(-\delta_1 \cdot d^{\delta_2} \cdot |r|)\big)$$

Yasso20 uses a quadratic formulation:

$$f_{\text{size}}^{\text{20}} = \big(1 + \theta_1\, d + \theta_2\, d^2\big)^{-|r|}$$

which produces a stronger size-dependent slowdown for coarse woody debris.

**Substantially different parameter values.** The most consequential changes from Yasso07/15 are:

| Parameter | Yasso07 | Yasso20 | Change | Effect |
|:----------|:--------|:--------|:-------|:-------|
| $p_H$ | 0.04 | 0.0042 | $-$90\% | Much less humus formation |
| $\alpha_H$ | 0.0033 | 0.0015 | $-$54\% | Slower humus decomposition |
| $\alpha_W$ | 4.30 | 5.19 | $+$21\% | Faster W-pool turnover |
| $\alpha_E$ | 0.35 | 0.13 | $-$63\% | Slower E-pool turnover |
| $\alpha_N$ | 0.22 | 0.10 | $-$55\% | Slower N-pool turnover |

The dramatically reduced $p_H$ (9.5$\times$ less humus formation) is the primary driver of Yasso20 predicting $\sim$20\% lower total SOC than Yasso07/15 for Finnish forests, despite the compensating effect of slower humus decomposition.


## 4.4 RothC

### Pool structure

Five pools: $\mathbf{C} = [C_{\text{DPM}},\, C_{\text{RPM}},\, C_{\text{BIO}},\, C_{\text{HUM}},\, C_{\text{IOM}}]^T$, representing Decomposable Plant Material, Resistant Plant Material, Microbial Biomass, Humified Organic Matter, and Inert Organic Matter.

### Dynamics

RothC operates on a **monthly timestep** $\Delta t = 1/12$\,yr. For each active pool $i \in \{\text{DPM}, \text{RPM}, \text{BIO}, \text{HUM}\}$:

$$C_i(t + \Delta t) = C_i(t) \cdot \exp\!\big(-k_i \cdot r(t) \cdot \Delta t\big) + I_i \cdot \Delta t + \text{transfers}_i$$

where $r(t) = \xi_T(t) \cdot \xi_M(t) \cdot \xi_V$ is the composite rate modifier (normalised so that $\xi_T / 47.9$ gives the fractional effect).

The IOM pool is inert: $C_{\text{IOM}}(t) = 0.049 \cdot (\%\text{clay})^{1.139}$.

**Decomposed material** from each pool:
$$D_i = C_i \cdot \big(1 - \exp(-k_i \cdot r \cdot \Delta t)\big)$$

This is partitioned into three fluxes governed by the clay-dependent ratio $x$:

$$x = 1.67\big(1.85 + 1.60\,\exp(-0.0786 \cdot \%\text{clay})\big)$$

\begin{align*}
f_{\text{CO}_2} &= x / (x + 1) \\
f_{\text{BIO}} &= 0.46 / (x + 1) \\
f_{\text{HUM}} &= 0.54 / (x + 1)
\end{align*}

So the transfers are:
\begin{align*}
\text{CO}_{2} &= \sum_i D_i \cdot f_{\text{CO}_2} \\
C_{\text{BIO}} &\mathrel{+}= \sum_i D_i \cdot f_{\text{BIO}} \\
C_{\text{HUM}} &\mathrel{+}= \sum_i D_i \cdot f_{\text{HUM}}
\end{align*}

Only DPM and RPM receive external inputs: $I_{\text{DPM}},\, I_{\text{RPM}}$ (partitioned from total litter via the DPM/RPM ratio).

### Environmental modifiers

**Temperature** (monthly):
$$\xi_T = \frac{47.9}{1 + \exp\!\big(106\,/\,(T + 18.3)\big)}$$

**Moisture** — based on accumulated topsoil moisture deficit (TSMD):
$$\text{TSMD}_m = \max\!\big(0,\; E_m - P_m\big)$$

The moisture modifier reduces decomposition when soil dries:
$$\xi_M = \max\!\big(0.2,\; 1 - \text{TSMD} / \text{TSMD}_{\max}\big)$$

where $\text{TSMD}_{\max}$ depends on clay content.

**Vegetation cover:**
$$\xi_V = \begin{cases} 0.6 & \text{vegetated (all forest plots)} \\ 1.0 & \text{bare soil} \end{cases}$$

### Analytical steady state

The system can be expressed as $\mathbf{A}\,\mathbf{C}^* = -\mathbf{b}$ for the four active pools, where $\mathbf{A}$ is the $4 \times 4$ matrix encoding losses and transfers, and $\mathbf{b} = [I_{\text{DPM}},\, I_{\text{RPM}},\, 0,\, 0]^T$.

$$\mathbf{C}^*_{\text{active}} = -\mathbf{A}^{-1}\,\mathbf{b}$$

The full steady state appends $C_{\text{IOM}}$ calculated from clay content.

### Default parameters

| Parameter | Value | Description |
|:----------|:------|:------------|
| $k_{\text{DPM}}$ | 10.0 | DPM decomposition rate (yr$^{-1}$) |
| $k_{\text{RPM}}$ | 0.3 | RPM decomposition rate (yr$^{-1}$) |
| $k_{\text{BIO}}$ | 0.66 | BIO decomposition rate (yr$^{-1}$) |
| $k_{\text{HUM}}$ | 0.02 | HUM decomposition rate (yr$^{-1}$) |

\newpage

## 4.5 Q-model

### Pool structure

Unlike the discrete-pool models above, the Q-model (Bosatta \& Ågren, 1985) describes decomposition along a **continuous quality axis** $q$. Organic matter is characterised by its current quality $q \in (0, q_0]$, where $q_0$ is the initial quality at litter input. As material decomposes, quality declines and decomposition slows — encoding the universal observation that fresh litter decomposes faster than aged organic matter.

In practice, the model is discretised into **cohort tracking**: each annual litter input forms a cohort whose remaining mass is tracked as a function of age and tissue type.

Five tissue classes are tracked separately, distinguished by initial quality and entry dynamics:

| Tissue class | Quality parameter | Entry dynamics |
|:-------------|:-----------------|:---------------|
| Foliage (needles) | $q_{0n}$ (species-specific) | Immediate |
| Fine roots | $q_{0f}$ (species-specific) | Immediate |
| Understorey | $q_{0n}$ (as foliage) | Immediate |
| Branches | $q_{0w}$ (species-specific) | Gradual fragmentation ($t_{\max,b}$ = 13 yr) |
| Stems | $q_{0w}$ (species-specific) | Gradual fragmentation ($t_{\max,s}$ = 60 yr) |

### Core equations

**Microbial uptake rate** as a function of quality:
$$u(q) = u_0 \cdot q^{\beta}$$

where $u_0$ is the temperature-dependent base rate and $\beta$ is the quality--rate exponent (default $\beta = 7$).

**Temperature dependence** (linear form):
$$u_0(T) = u_{00} + u_{01} \cdot T$$

where $T$ is the mean annual temperature (°C), $u_{00} = 0.0855$, and $u_{01} = 0.0157$.

An alternative thermodynamic (Arrhenius) formulation (not implemented here) exists (Bosatta \& Ågren 1999):
$$u(q, T) = u_0^{\text{ref}} \cdot \exp\!\left(\frac{|\Delta G^0|}{R\,q}\left(\frac{1}{T_{\text{ref}}} - \frac{1}{T}\right)\right)$$

where $T$ is in Kelvin, $R = 8.314$\,J\,mol$^{-1}$\,K$^{-1}$, and $\Delta G^0$ is the free energy change per enzymatic step. This formulation implies that low-quality substrates have stronger temperature sensitivity than high-quality substrates — a feature not captured by the linear form.

**Derived parameters.** The quality decline exponent $z$ controls how decomposition slows over time:
$$z = \frac{1 - e_0}{\beta \cdot \eta_{11} \cdot e_0}$$

where $e_0$ is the initial microbial growth efficiency (0.25) and $\eta_{11}$ is the carbon use efficiency parameter (0.36). With default parameters, $z \approx 1.19$.

The composite decomposition rate $\alpha$ for a given tissue type combines all factors:
$$\alpha = f_C \cdot \beta \cdot \eta_{11} \cdot u_0 \cdot q_0^{\beta}$$

where $f_C = 0.5$ is the carbon fraction of organic matter.

### Remaining fraction functions

**Simple tissues** (foliage, fine roots, understorey) enter the soil immediately and follow power-law decay:
$$g(t) = \frac{1}{(1 + \alpha\, t)^z}$$

**Branches** fragment gradually over $t_{\max,b}$ years. The remaining fraction integrates decomposition over the fragmenting material. Defining $i_b = \min(t,\, t_{\max,b})$:

$$g_b(t) = \frac{2\Big[(1 + \alpha t)^{1-z} - (1 + \alpha(t - i_b))^{1-z}(1 - i_b/t_{\max,b})\Big]}{\alpha\,(1-z)\,t_{\max,b}} + \frac{2\Big[(1 + \alpha(t-i_b))^{2-z} - (1+\alpha t)^{2-z}\Big]}{\alpha^2(1-z)(2-z)\,t_{\max,b}^2} + \left(1 - \frac{i_b}{t_{\max,b}}\right)^2$$

**Stems** follow the same structure with $t_{\max,s}$ replacing $t_{\max,b}$.

### Quality evolution

Quality declines with mass loss:
$$q(t) = q_0 \left(\frac{C(t)}{C(0)}\right)^{1/z}$$

This couples quality and mass: as carbon is lost, the remaining material becomes progressively more recalcitrant.

### Transient temperature implementation

The standard Q-model analytical solution assumes constant $\alpha$ (i.e.\ constant temperature). Under transient climate, we track **cumulative decomposition potential** for each cohort:

$$\tilde{\alpha}(t) = \sum_{\tau=1}^{t} \alpha\big(T(\tau)\big)$$

where $\alpha(T(\tau))$ is recalculated each year from the current temperature. The remaining fraction then becomes:

$$g(t) = \frac{1}{\big(1 + \tilde{\alpha}(t)\big)^z}$$

This ensures all carbon pools respond to current temperature conditions rather than being locked to the temperature at the time of input — a critical requirement for climate change projections. Without this correction, the model underestimates SOC loss under warming by 20--40\%.

### Initialisation: hybrid approach

Rather than running a 1500-year numerical spinup, we use an analytical-hybrid approach:

1. **Analytical steady state.** For simple tissues:
$$C_{\text{ss}} = L \cdot \sum_{t=1}^{T_{\max}} g(t)$$
where $L$ is the annual litter input. Branches and stems use numerical integration of their respective $g_b$, $g_s$ functions over 500--1000 years.

2. **Bulk pool initialisation.** The total $C_{\text{ss}}$ is treated as a single bulk pool that decomposes via:
$$C_{\text{init}}(t) = C_{\text{ss}} \cdot (1 + \alpha_0\, t)^{1-z}$$

3. **Short spinup** (10--20 years) builds up explicit cohort matrices while the bulk pool decays. Total SOC = bulk pool + new cohorts.

This reduces initialisation time from $\sim$1500 years to 10--20 years while maintaining accuracy.

### Default parameters

| Parameter | Value | Description |
|:----------|:------|:------------|
| $\beta$ | 7 | Quality--rate exponent |
| $e_0$ | 0.25 | Initial microbial growth efficiency |
| $\eta_{11}$ | 0.36 | Carbon use efficiency parameter |
| $f_C$ | 0.5 | Carbon fraction of organic matter |
| $u_{00}$ | 0.0855 | Base uptake rate at 0°C |
| $u_{01}$ | 0.0157 | Temperature sensitivity of uptake |
| $t_{\max,b}$ | 13 | Branch fragmentation time (yr) |
| $t_{\max,s}$ | 60 | Stem fragmentation time (yr) |
| $q_{0n}$ | 1.01 (spruce) | Initial quality, foliage |
| $q_{0f}$ | 1.036 | Initial quality, fine roots |
| $q_{0w}$ | 1.00 | Initial quality, woody tissues |





## 4.6 Common Interface and Implementation Architecture

All five decomposition models share a common calling interface, enabling the simulation loop and Bayesian calibration framework to treat them interchangeably. The implementation is split across two languages: performance-critical numerical code in **Fortran**, and data handling, calibration logic, and input preparation in **R**.

### Fortran--R boundary

The guiding principle is to minimise R$\leftrightarrow$Fortran call overhead during MCMC sampling. Since the calibration loop evaluates the likelihood millions of times across $\sim$3,700 plots, each evaluation should require a **single Fortran call per plot** that runs the full time-series simulation, rather than hundreds of per-timestep round-trips.

\begin{figure}[H]
\centering
\begin{tikzpicture}[
    >=Stealth,
    node distance=0.8cm and 1.6cm,
    box/.style={
        rectangle, rounded corners=4pt, draw=#1, fill=#1!8,
        line width=1pt, minimum height=1.0cm, minimum width=3.4cm,
        text centered, font=\small\sffamily
    },
    widebox/.style={
        rectangle, rounded corners=4pt, draw=#1, fill=#1!8,
        line width=1pt, minimum height=1.0cm, minimum width=5.0cm,
        text centered, font=\small\sffamily
    },
    data/.style={
        rectangle, rounded corners=2pt, draw=bordergray, fill=bglight,
        line width=0.6pt, minimum height=0.6cm, minimum width=2.0cm,
        text centered, font=\footnotesize\ttfamily
    },
    arr/.style={->, line width=1pt, color=#1},
    note/.style={font=\scriptsize\sffamily\itshape, text=black!50},
    layerlabel/.style={font=\small\sffamily\bfseries, text=#1, rotate=90, anchor=south}
]

% =============================================
% OUTSIDE-MCMC
% =============================================

\node[box=compatgreen] (fi) at (-4.0, 2.5) {%
    \begin{tabular}{c}
    $f_{i_k}(\mathbf{I}(t))$ \\ Input layer
    \end{tabular}};
\node[note, below=0.05cm of fi, text width=2.8cm, align=center]
    {once per plot};

\node[box=compatgreen] (SOC) at (3.0, 2.5) {%
    \begin{tabular}{c}
    SOC measurements
    \end{tabular}};
\node[note, below=0.05cm of SOC, text width=2.8cm, align=center]
    {once per plot};

\node[box=compatgreen] (envdata) at (6.5, 2.5) {%
    \begin{tabular}{c}
    $g_{i_k}(T(t),\,M(t))$ \\ Climate layer
    \end{tabular}};
\node[note, below=0.05cm of envdata, text width=2.8cm, align=center]
    {once per plot};

% =============================================
% INSIDE-MCMC: Likelihood
% =============================================

\node[widebox=outputpurple] (likelihood) at (3, 0.0) {%
    \begin{tabular}{c}
    Likelihood \\ $\mathcal{L}(\boldsymbol{\theta} \,|\, \text{SOC}_{\text{obs}})$
    \end{tabular}};

% =============================================
% INSIDE-MCMC: R wrapper components
% =============================================

% steady_state (Fortran, red) — leftmost
\node[box=decompred] (steadyblock) at (-4.0, -3.0) {%
    \begin{tabular}{c}
    \texttt{steady\_state()} \\
    $\mathbf{C}^* = -\mathbf{A}^{-1}\mathbf{b}$
    \end{tabular}};

% model_run (Fortran, red) — middle
\node[
    rectangle, rounded corners=4pt,
    draw=decompred, fill=decompred!8,
    line width=1pt,
    minimum height=3.0cm, minimum width=4.2cm,
    text centered, font=\small\sffamily
] (frun) at (1.0, -3.0) {};

\node[font=\small\sffamily, text=decompred, anchor=north]
    at (frun.north) [yshift=-6pt] {%
    \begin{tabular}{c}
    \texttt{model\_run()} \\ loop over $c = 1,\ldots,n_c$
    \end{tabular}};

\node[box=decompred, anchor=center] (fstep) at (1.0, -3.6) {%
    \begin{tabular}{c}
    \texttt{model\_step()} \\ single $\Delta t$
    \end{tabular}};

% compute_xi (R, gold) — right of model_run
\node[box=modifiergold] (fxi) at (6.5, -3.0) {%
    \begin{tabular}{c}
    \texttt{compute\_xi()} \\ $\boldsymbol{\xi}(T, M, \boldsymbol{\theta})$
    \end{tabular}};

% Output
\node[data] (output) at (-1.5, -6.5) {SOC$(t)$ vector};

% =============================================
% GROUPING BOXES (background layer)
% =============================================
\begin{scope}[on background layer]

\node[
    rectangle, rounded corners=8pt,
    draw=outputpurple, fill=outputpurple!4,
    line width=1.5pt, dashed,
    fit=(likelihood)(steadyblock)(frun)(fxi)(output),
    inner sep=24pt
] (mcmcbox) {};

\node[
    rectangle, rounded corners=6pt,
    draw=compatgreen, fill=compatgreen!4,
    line width=1.2pt, dashed,
    fit=(steadyblock)(frun)(fxi)(output),
    inner sep=14pt
] (rwrapbox) {};

\end{scope}

% Labels
\node[font=\small\sffamily\bfseries, text=compatgreen,
    anchor=south west] at (rwrapbox.north west)
    {R wrapper};
\node[font=\small\sffamily\bfseries, text=outputpurple,
    below=0.4cm of mcmcbox.south, anchor=south]
    {MCMC sampler};
\node[note, below=0.05cm of mcmcbox.north, anchor=north,
    text width=5cm, align=center]
    {re-evaluated each proposal $\boldsymbol{\theta}$};

% =============================================
% ARROWS — outside to inside
% =============================================

% SOC -> likelihood
\draw[arr=compatgreen] (SOC.south) -- (likelihood.north)
    node[midway, right=3pt, note] {$\text{SOC}_{\text{obs}}$};

% Input layer -> steady_state
\draw[arr=compatgreen] (fi.south) -- (steadyblock.north)
    node[midway, left=3pt, note] {$\mathbf{J}_c(t),\,\bar{\mathbf{J}}$};

% Input layer -> model_run
\draw[arr=compatgreen] (fi.south) -- ++(0,-1.0) -| (frun.north)
    node[near end, right=3pt, note] {$\mathbf{J}_c(t)$};

% Climate layer -> compute_xi only
\draw[arr=compatgreen] (envdata.south) -- (fxi.north)
    node[midway, right=3pt, note] {$T(t),\,M(t)$};

% =============================================
% ARROWS — inside MCMC
% =============================================

% Likelihood -> steady_state
\draw[arr=outputpurple] (likelihood.south) -- ++(0,-0.3) -| (steadyblock.110)
    node[near end, left=2pt, note] {$\boldsymbol{\theta}$};

% Likelihood -> model_run
\draw[arr=outputpurple] (likelihood.south) -- ++(0,-0.3) -| (frun.110)
    node[near end, left=2pt, note] {$\boldsymbol{\theta}$};

% Likelihood -> compute_xi
\draw[arr=outputpurple] (likelihood.east) -- ++(0.7,0) -- (fxi.119)
    node[near end, left=2pt, note] {$\boldsymbol{\theta}$};

% compute_xi -> model_run
\draw[arr=modifiergold] (fxi.west) -- (frun.east)
    node[midway, above=2pt, note] {$\boldsymbol{\xi}(t)$};

% steady_state -> model_run
\draw[arr=decompred] (steadyblock.east) -- (frun.west)
    node[midway, above=2pt, note] {$\mathbf{C}^*$};

% model_run -> output
\draw[arr=outputpurple] (frun.south) -- ++(0,-0.5) |- (output.east);

% output -> likelihood
\draw[arr=outputpurple, line width=1.5pt]
    (output.west) -- ++(-3.8,0) |- (likelihood.west)
    node[near start, above=-30pt, note] {SOC predictions};

\end{tikzpicture}
\caption{%
Revised implementation architecture.
Three blocks outside the MCMC loop are computed once per plot:
the input layer $f_{i_k}$ maps raw litter data to model-specific arrays $\mathbf{J}_c(t)$
and mean inputs $\bar{\mathbf{J}}$; the climate layer $g_{i_k}$ aggregates raw climate
to the resolution required by each model; and SOC observations feed the likelihood.
At each MCMC proposal $\boldsymbol{\theta}$, the likelihood dispatches parameters
to the \textbf{R wrapper}, which runs three components:
\texttt{steady\_state()} (Fortran, red) receives $\boldsymbol{\theta}$ and $\bar{\mathbf{J}}$
and initialises pools $\mathbf{C}^*$;
\texttt{compute\_xi()} (R, gold) receives $\boldsymbol{\theta}$ and the pre-computed
climate series to produce $\boldsymbol{\xi}(t)$;
and \texttt{model\_run()} (Fortran, red) receives $\mathbf{C}^*$, $\boldsymbol{\xi}(t)$,
$\boldsymbol{\theta}$, and $\mathbf{J}_c(t)$, iterates \texttt{model\_step()} over all
cohorts and timesteps, and returns the SOC trajectory for likelihood evaluation.}
\label{fig:implementation}
\end{figure}


### Fortran module structure

Each model is implemented as a Fortran module exposing three subroutines. All share the same calling conventions.

**\texttt{model\_run()}** — the top-level entry point called from R via \texttt{.Fortran()}. It receives the full pre-processed input arrays, parameter vector, and climate time series for a single plot, and returns the complete SOC trajectory.

```{r, eval=FALSE}
# R side: single call per plot
result <- .Fortran("yasso15_run",
  n_years    = as.integer(n_years),
  params     = as.double(param_vector),    # calibrated parameters
  J_t        = as.double(input_matrix),    # [n_years x n_input_pools]
  T_monthly  = as.double(temp_matrix),     # [n_years x 12]
  P_monthly  = as.double(precip_matrix),   # [n_years x 12]
  diameter   = as.double(wood_diameters),  # [n_litter_types]
  C_out      = double(n_years * n_pools),  # output: [n_years x n_pools]
  resp_out   = double(n_years)             # output: annual respiration
)
```

Internally, \texttt{model\_run()} performs:

1. Call \texttt{model\_steady\_state()} to compute $\mathbf{C}^*$ from mean climate and inputs.
2. Initialise pools from $\mathbf{C}^*$ (or from the hybrid approach for Q-model).
3. Loop over years (and months for RothC), calling \texttt{compute\_xi()} and \texttt{model\_step()} at each timestep.
4. Write annual pool states and respiration to output arrays.

**\texttt{model\_steady\_state()}** — computes the analytical equilibrium. For Yasso and RothC, this is a matrix inversion $\mathbf{C}^* = -\mathbf{A}^{-1}\mathbf{b}$. For the Q-model, it performs the numerical summation of remaining fractions over a truncated age series.

**\texttt{model\_step()}** — advances the pool state by one timestep $\Delta t$:

| Model | $\Delta t$ | Core operation |
|:------|:-----------|:---------------|
| Yasso07/15/20 | 1 yr | Build $\mathbf{A}$, compute matrix exponential $e^{\mathbf{A}}$, update $\mathbf{C}$ |
| RothC | 1/12 yr | Exponential decay per pool, partition decomposed C via clay ratio $x$ |
| Q-model | 1 yr | Update cumulative $\tilde{\alpha}$ per cohort, compute remaining fraction, prune old cohorts |

**\texttt{compute\_xi()}** — evaluates the environmental modifiers from climate data and static site properties. Called internally by \texttt{model\_step()} (or once per year for annual models, 12 times for RothC).


### R wrapper layer

The R wrapper for each model handles:

- **Climate mapping**: calls `map_climate_<model>()` once per plot to produce the climate data frame expected by that model. Yasso07/15 use `map_climate_yasso07()` (annual mean + amplitude); Yasso20 and RothC share `map_climate_monthly()` (full monthly series).
- **Input mapping**: calls `map_inputs_<model>()` once per plot to aggregate the six raw cohorts into model-specific litter arrays. Yasso07, Yasso15 and Yasso20 share the same input structure via `map_inputs_yasso07()`.
- **Parameter mapping**: translates the named BayesianTools parameter vector into the positional double-precision vector expected by Fortran.
- **Output formatting**: reshapes the flat Fortran output arrays into named R structures for likelihood evaluation and diagnostics.

All models are run plot-by-plot with a consistent `lapply` → `rbind` pattern. Climate and input mapping are performed once per plot, outside the MCMC loop:

```{r, eval=FALSE}
# --- Once, before calibration ---
plots          <- unique(input_raw$plot_id)
Yasso07_climate <- map_climate_yasso07(input_raw)   # annual: mean, amplitude, precip
monthly_climate <- map_climate_monthly(input_raw)   # monthly: shared by Yasso20 + RothC
Yasso07_inputs  <- map_inputs_yasso07(input_raw)    # shared by Yasso07, Yasso15, Yasso20
RothC_inputs    <- map_inputs_rothc(input_raw, site_raw)

# --- Per-plot simulation (inside likelihood, called each MCMC proposal) ---
result <- do.call(rbind, lapply(plots, function(pid) {

  clim   <- Yasso07_climate[Yasso07_climate$plot_id == pid, ]
  inputs <- Yasso07_inputs[Yasso07_inputs$plot_id   == pid, ]

  # Stage 1: Xi time series from plot-specific climate
  xi <- compute_xi_yasso07(clim, params)

  # Stage 2: Steady-state initialisation from plot-specific means
  C_init <- yasso07_steady_state(
    params   = params,
    nwl_mean = colMeans(inputs[, c("nwl_A","nwl_W","nwl_E","nwl_N")]),
    fwl_mean = colMeans(inputs[, c("fwl_A","fwl_W","fwl_E","fwl_N")]),
    cwl_mean = colMeans(inputs[, c("cwl_A","cwl_W","cwl_E","cwl_N")]),
    xi_mean  = mean(xi)
  )

  # Stage 3: Transient simulation — single Fortran call per plot
  out <- yasso07_run(inputs, params, C_init, xi)

  cbind(plot_id = pid, out)
}))
```


### What stays in R vs.\ Fortran

| Component | Language | Rationale |
|:----------|:---------|:----------|
| Climate mapping (`map_climate_*`) | R | Called once per plot; benefits from R data wrangling |
| Input mapping (`map_inputs_*`) | R | Called once per plot; benefits from R data wrangling |
| $\xi_k$ — modifier computation | Fortran | Called millions of times inside MCMC loop |
| $f_{d_k}$ — decomposition step | Fortran | Innermost loop; matrix operations |
| Steady-state solver | Fortran | Matrix inversion, numerical integration |
| Full simulation loop | Fortran | Avoids per-timestep R$\leftrightarrow$Fortran overhead |
| Parameter mapping | R | String/name handling; negligible cost |
| Likelihood evaluation | R | Interface with BayesianTools; simple arithmetic |
| I/O, aggregation, diagnostics | R | Flexibility; not performance-critical |

\newpage

## 4.7 Yasso Family: Three-Stage Call Pattern

All three Yasso versions (07, 15, 20) follow the same three-stage execution pattern in R. The differences are confined to Stage 1 (Xi computation) and the climate object passed to each stage.

### Climate and input mapping functions

| Model | Climate function | Input function |
|:------|:----------------|:---------------|
| Yasso07 | `map_climate_yasso07()` → annual mean, amplitude, precip | `map_inputs_yasso07()` |
| Yasso15 | `map_climate_yasso07()` (reused) | `map_inputs_yasso07()` (reused) |
| Yasso20 | `map_climate_monthly()` (shared with RothC) | `map_inputs_yasso07()` (reused) |

All three versions share the same litter input structure: six raw cohorts aggregated to three litter types (nwl, fwl, cwl) with AWEN fractions preserved.

### Stage 1 — Xi: climate modifier time series

The key structural difference among the three versions is in how $\xi$ is computed:

| Model | Function | Climate resolution | Pool-group specificity |
|:------|:---------|:-------------------|:-----------------------|
| Yasso07 | `compute_xi_yasso07()` | Annual (mean + amplitude → 4-point Gaussian) | Single $\xi$ for all pools |
| Yasso15 | `compute_xi_yasso15()` | Annual (mean + amplitude → 4-point Gaussian) | Three $\xi$: AWE, N, H |
| Yasso20 | `compute_xi_yasso20()` | Monthly (exact 12-month average) | Three $\xi$: AWE, N, H |

### Stage 2 — Steady state: initial pool conditions $\mathbf{C}^*$

Steady state is computed from plot-specific mean climate and mean litter inputs. For Yasso20, a synthetic 12-row `mean_clim` data frame is constructed from the plot's monthly climatology:

```{r, eval=FALSE}
# Yasso07 / Yasso15
C_init <- yassoXX_steady_state(
  params   = params,
  nwl_mean = colMeans(inputs[, c("nwl_A","nwl_W","nwl_E","nwl_N")]),
  fwl_mean = colMeans(inputs[, c("fwl_A","fwl_W","fwl_E","fwl_N")]),
  cwl_mean = colMeans(inputs[, c("cwl_A","cwl_W","cwl_E","cwl_N")]),
  xi_mean  = mean(xi)   # or list(xi_awe, xi_n, xi_h) for Yasso15
)

# Yasso20 — requires a synthetic single-year mean climate data frame
# precip must be monthly mean (NOT divided by 12); the function sums
# internally over 12 months to recover the annual total
mean_clim <- data.frame(
  year     = 9999L,
  month    = 1:12,
  temp_air = tapply(clim$temp_air, clim$month, mean, na.rm = TRUE),
  precip   = mean(clim$precip)   # monthly mean ~50 mm; sums to ~600 mm/yr internally
)
C_init <- yasso20_steady_state(mean_clim, params, nwl_mean, fwl_mean, cwl_mean)
```

### Stage 3 — Transient simulation

All three versions return a flat data frame with columns `year | A | W | E | N | H | total_soc | respiration`.

```{r, eval=FALSE}
out_07 <- yasso07_run(inputs_df,    params, C_init, xi_array)
out_15 <- yasso15_run(inputs_df,    params, C_init, xi_arrays, precip, leac)
out_20 <- yasso20_run(climate_df,   inputs_df, params, C_init, leac)
#                     ^^^^^^^^^^
#                     monthly_climate — xi_arrays and precip derived internally
```
