---
title: "Modular SOC Decomposition Framework"
subtitle: "Architecture Documentation for Multi-Model Intercomparison"
author: "Lorenzo — HIKET Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{tikz}
  - \usepackage{colortbl}
  - \usetikzlibrary{arrows.meta, positioning, calc, fit, backgrounds, shapes.geometric, decorations.pathreplacing}
  - \definecolor{inputblue}{HTML}{3B82F6}
  - \definecolor{compatgreen}{HTML}{10B981}
  - \definecolor{decompred}{HTML}{EF4444}
  - \definecolor{modifiergold}{HTML}{F59E0B}
  - \definecolor{outputpurple}{HTML}{8B5CF6}
  - \definecolor{bglight}{HTML}{F8FAFC}
  - \definecolor{bordergray}{HTML}{CBD5E1}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

\vspace{1em}

# 1. Overview

This document defines a modular architecture for soil organic carbon (SOC) decomposition modelling within the HIKET multi-model intercomparison framework. The goal is to define all models — Yasso07, Yasso15, Yasso20, RothC, and Q-model — through common functional interfaces that cleanly separates **litter input processing**, **environmental modification of kinetics**, and **decomposition dynamics**.

This separation enables modularity.

\newpage

# 2. General Framework

## 2.1 Core Equation

The simulated soil organic carbon over time is expressed as:

$$
\mathbf{C}(t, k) = f_{d_k}\!\Big(\, f_{i_k}\!\big(\mathbf{I}(t)\big),\;\boldsymbol{\xi_k}(t) \Big)
$$

where:

| Symbol | Description | Dimensions |
|:-------|:-----------|:-----------|
| $\mathbf{C}(t)$ | SOC pool vector at time $t$ | $[n_{\text{pools}} \times 1]$ |
| $\mathbf{I}(t)$ | Raw litter input data at time $t$ | model-agnostic format |
| $f_i(\cdot)$ | **Input compatibility layer** — transforms raw inputs into model-specific pool inputs | $\mathbf{I}(t) \to \mathbf{J}(t)$ |
| $f_d(\cdot)$ | **Decomposition function** — the structural core of each model | $(\mathbf{J}(t), \boldsymbol{\xi}(t)) \to \mathbf{C}(t)$ |
| $\boldsymbol{\xi}(t)$ | **Kinetic modifier** — composite environmental scalar(s) | $\in \mathbb{R}^{n_{\xi}}$ |

The $k$ denotes model-specific functions.

The kinetic modifier is itself composed of independent environmental response functions:

$$
{\xi}(t) = \xi_T\!\big(T(t)\big) \;\cdot\; \xi_M\!\big(M(t)\big) \;\cdot\;  \xi_e
$$

where $\xi_T$ is the temperature modifier, $\xi_M$ is the moisture/precipitation modifier, and $\xi_e$ represents any additional static or time-varying modifiers, likely edaphic (e.g., clay content, soil cover).

## 2.2 Architecture Schema

\begin{figure}[H]
\centering
\begin{tikzpicture}[
    >=Stealth,
    node distance=1.2cm and 2.0cm,
    box/.style={
        rectangle, rounded corners=4pt, draw=#1, fill=#1!8,
        line width=1pt, minimum height=1.1cm, minimum width=3.2cm,
        text centered, font=\small\sffamily
    },
    smallbox/.style={
        rectangle, rounded corners=3pt, draw=#1, fill=#1!8,
        line width=0.8pt, minimum height=0.8cm, minimum width=2.2cm,
        text centered, font=\footnotesize\sffamily
    },
    tinybox/.style={
        rectangle, rounded corners=2pt, draw=#1, fill=#1!8,
        line width=0.6pt, minimum height=0.6cm, minimum width=1.6cm,
        text centered, font=\scriptsize\sffamily
    },
    data/.style={
        rectangle, rounded corners=2pt, draw=bordergray, fill=bglight,
        line width=0.6pt, minimum height=0.7cm, minimum width=2.0cm,
        text centered, font=\footnotesize\ttfamily
    },
    arr/.style={->, line width=1pt, color=#1},
    darr/.style={->, line width=0.8pt, dashed, color=#1},
    brace/.style={decorate, decoration={brace, amplitude=6pt, raise=2pt}},
    note/.style={font=\scriptsize\sffamily\itshape, text=black!60}
]

% =============================================
% RAW INPUTS (left)
% =============================================
\node[data] (I_litter) {Litter masses by component};
\node[data, below=0.35cm of I_litter] (I_species) {Species composition};

% Brace for inputs
\draw[brace, bordergray, line width=0.8pt]
    (I_litter.north east) -- (I_species.south east)
    node[midway, right=12pt, font=\small\sffamily, inputblue] {$\mathbf{I}(t)$};

% =============================================
% INPUT COMPATIBILITY LAYER
% =============================================
\node[box=compatgreen, right=1.8cm of I_litter, yshift=-0.35cm] (fi) {%
    \begin{tabular}{c}
    I compatibility \\  Layer \\ $f_{i_k}\!\big(\mathbf{I}(t)\big)$
    \end{tabular}};

% =============================================
% DECOMPOSITION FUNCTION
% =============================================
\node[box=decompred, right=2.5cm of fi] (fd) {%
    \begin{tabular}{c}
    Decomposition \\ $f_{d_k}\!\big(\mathbf{J}(t),\, \boldsymbol{\xi}(t)\big)$
    \end{tabular}};

% Hard-coded note inside fd
\node[note, below=0.15cm of fd] (fd_note) {incl.\ hard-coded size modifiers};

% =============================================
% OUTPUT
% =============================================
\node[box=outputpurple, below=2.2cm of fd] (C) {%
    \begin{tabular}{c}
    total SOC \\ $\mathbf{C}(t)$
    \end{tabular}};

% =============================================
% ENVIRONMENTAL DRIVERS (top row)
% =============================================
\node[data, above=5.2cm of fi, xshift=-0.5cm] (T) {$T_{\text{monthly}}(t)$};
\node[data, right=1.8cm of T] (M) {$M_{\text{monthly}}(t)$};
\node[data, right=1.8cm of M] (clay) {clay \%};

% =============================================
% TEMPERATURE BRANCH
% =============================================
% Temporal aggregation
\node[tinybox=bordergray, below=0.5cm of T] (T_agg) {%
    \begin{tabular}{c}
    $\overline{T}$ or $T_m$
    \end{tabular}};
\node[note, left=0.1cm of T_agg] {avg/pass};

% Temperature modifier
\node[smallbox=modifiergold, below=0.5cm of T_agg] (xiT) {$\xi_{T_k}\!\big(T\big)$};

% =============================================
% MOISTURE BRANCH
% =============================================
% Moisture compatibility layer
\node[tinybox=compatgreen, below=0.5cm of M] (fm) {%
    \begin{tabular}{c}
    $f_{m_k}$
    \end{tabular}};
\node[note, right=0.1cm of fm, text width=2cm, align=left] {$P$ or TSMD\\or $\emptyset$};

% Moisture modifier
\node[smallbox=modifiergold, below=0.5cm of fm] (xiM) {$\xi_{M_k}\!\big(f_{m_k}\big)$};

% =============================================
% EDAPHIC BRANCH
% =============================================
\node[smallbox=modifiergold, below=1.55cm of clay] (xiE) {$\xi_e$};
\node[note, right=0.1cm of xiE] {RothC only};

% =============================================
% COMPOSITE MODIFIER
% =============================================
\node[box=modifiergold, below=0.8cm of xiM] (xi) {%
    \begin{tabular}{c}
    Kinetic Modifier \\ $\boldsymbol{\xi}_k(t) = \xi_T \cdot \xi_M \cdot \xi_e$
    \end{tabular}};

% =============================================
% ARROWS — Inputs
% =============================================
\draw[arr=inputblue] (I_litter.east) -- ++(0.8,0) |- (fi.west);
\draw[arr=inputblue] (I_species.east) -- ++(0.8,0) |- (fi.west);

% fi to fd
\draw[arr=compatgreen] (fi.east) -- (fd.west)
    node[midway, below=2pt, font=\footnotesize\sffamily, compatgreen] {$\mathbf{J}(t)$};

% fd to output
\draw[arr=decompred] (fd.east) -- ++(1.2,0) |- (C.east);

% =============================================
% ARROWS — Temperature branch
% =============================================
\draw[arr=bordergray] (T.south) -- (T_agg.north);
\draw[arr=modifiergold] (T_agg.south) -- (xiT.north);

% =============================================
% ARROWS — Moisture branch
% =============================================
\draw[arr=bordergray] (M.south) -- (fm.north);
\draw[arr=modifiergold] (fm.south) -- (xiM.north);

% =============================================
% ARROWS — Edaphic branch
% =============================================
\draw[arr=modifiergold] (clay.south) -- (xiE.north);

% =============================================
% ARROWS — Sub-modifiers to composite
% =============================================
\draw[arr=modifiergold] (xiT.south) -- ++(0,-0.2) -| ([xshift=-6pt]xi.north);
\draw[arr=modifiergold] (xiM.south) -- (xi.north);
\draw[arr=modifiergold] (xiE.south) -- ++(0,-0.2) -| ([xshift=6pt]xi.north);

% =============================================
% ARROW — Composite modifier to fd
% =============================================
\draw[arr=modifiergold] (xi.south) -- ++(0,-0.35) -| (fd.north);

% =============================================
% BACKGROUND — Modifier group
% =============================================
\begin{scope}[on background layer]
    \node[fit=(T_agg)(xiT)(fm)(xiM)(xiE)(xi),
          fill=modifiergold!4, draw=modifiergold!25,
          rounded corners=6pt, inner sep=10pt] {};
\end{scope}

\end{tikzpicture}
\caption{Modular architecture of the SOC decomposition framework. Raw inputs $\mathbf{I}(t)$ are transformed by the input compatibility layer $f_i$ into model-specific inputs $\mathbf{J}(t)$. Environmental drivers are combined into the composite kinetic modifier $\boldsymbol{\xi}(t)$. Both feed into the model-specific decomposition function $f_d$.}
\label{fig:architecture}
\end{figure}

\newpage


# 3. Input requirements

## 3.1 Standardised input format — $\mathbf{I}(t)$
Dimensions $N_{plots} \times N_{time steps}$

The standard input data for all models $\mathbf{I}(t)$ contains litter mass inputs from six cohorts (foliage, fine roots, branches, coarse roots, stump, stem).
The species group (pine, spruce, birch) are partitioned based on species composition data, which are assumed constant per plot.

## 3.2 Standised environmental drivers

### 3.2.1 Time-variant drivers
Dimensions $N_{plots} \times N_{time steps}$

Temperature and moisture inputs are provided as monthly time series.
Temperature relies only on a compatibility layer to aggregate to annual summaries where needed. 
Moisture is more complicated since the models rely on different inputs (precipitation vs topsoil moisture deficit), so we need a compatibility layer to convert precipitation to soil moisture deficit for RothC, and to pass precipitation directly to Yasso. The Q-model does not explicitly include a moisture modifier, but implicitly via litter quality.

### 3.2.2 Static edaphic modifiers
Dimensions $N_{plots} \times 1$

Clay content is provided as a scalar per plot, as well as tree species composition.

# 3. Output structure
## 3.1 Model-specific output

Dimensions $N_{plots} \times N_{time steps} \times N_{pools}$, replicated for each model (different data frames for each model, with different pool names and numbers).

## 3.1 Total SOC output
Dimensions $N_{plots} \times N_{time steps} \times N_{models}$$



# 3. Components requirements

## 3.1 Input Compatibility Layer — $f_i\!\big(\mathbf{I}(t)\big)$


The input layer $f_i$ maps this standardised input into the model-specific input vector $\mathbf{J}(t)$:

$$
f_i: \mathbf{I}(t) \;\longrightarrow\; \mathbf{J}(t) \in \mathbb{R}^{n_{\text{input pools}}}
$$

The mapping differs across models:

| Model | Input pools $\mathbf{J}(t)$ | Key transformation |
|:------|:---------------------------|:-------------------|
| Yasso07/15/20 | AWEN fractions $(A, W, E, N)$ | Total mass to AWEN via component-specific coefficients |
| RothC | DPM, RPM, BIO, HUM, IOM | Total mass $\to$ DPM/RPM split via component-dependent coefficients |
| Q-model | Quality-distributed cohorts per tissue type | Separate cohorts retained; quality $q_0$ assigned per species and component |




## 3.2 Kinetic Modifier — $\boldsymbol{\xi}(t)$

The kinetic modifier captures all environmental effects on decomposition rates. It is composed multiplicatively:

$$
\boldsymbol{\xi}(t) = \xi_T\!\big(T(t)\big) \;\times\; \xi_M\!\big(f_{m_k}(M(t))\big) \;\times\;  \xi_k
$$

Each sub-modifier is a vector over time (or, for edaphic effects $\xi_k$, a scalar, constant over time) that rescales the baseline decomposition rates.

### 3.2.1 Temperature Modifier — $\xi_T$

All functions accept the same input, with the only compatibility layer being a montlhy to annual averaging.  

The functional form of the temperature response differs across models:

| Model | Form | Expression |
|:------|:-----|:-----------|
| Yasso07/15 | Gaussian-like, 4-point annual | $\xi_T = \frac{1}{4}\sum_{i=1}^{4}\exp\!\big(\beta_1 T_i + \beta_2 T_i^2\big)$, $T_i$ from annual mean + amplitude |
| Yasso20 | Gaussian-like, exact monthly | $\xi_T = \frac{1}{12}\sum_{m=1}^{12}\exp\!\big(\beta_1 T_m + \beta_2 T_m^2\big)$, $T_m$ are observed monthly values |
| RothC | Arrhenius-based | $\xi_T = \dfrac{47.9}{1 + \exp\!\big(\frac{106}{T+18.3}\big)}$ |
| Q-model | Linear | $\xi_T = \max\!\big(0,\; a_T + b_T \cdot T\big)$ |

### 3.2.2 Moisture/Precipitation Modifier — $\xi_M$

Moisture functions of different models accept different inputs, so we need another compatibility layer, $f_{m_k}$. Yasso uses precipitation, RothC uses topsoil moisture deficit, and Q-model does not explicitly include a moisture modifier (but implicitly via litter quality).

| Model | Driver | Notes |
|:------|:-------|:------|
| Yasso07/15 | Annual precipitation $P$ | $\xi_M = 1 - \exp\!\big(\gamma \cdot P/1000\big)$; pool-group-specific for Yasso15 |
| Yasso20 | Monthly precipitation (summed annually) | Same formula; $P$ derived internally as annual sum of monthly values |
| RothC | Topsoil moisture deficit | Rate-modifying factor from TSMD accumulation |
| Q-model | — | Not explicitly included (implicitly via litter quality) |

### 3.2.3 Additional Static Modifiers — $\xi_k$
| Modifier | Applicable models | Description |
|:---------|:-----------------|:------------|
| Clay ($\xi_{\text{clay}}$) | RothC | Modifies BIO/HUM partitioning: $x = 1.67 (1.85 + 1.60 \exp(-0.0786\, \%\text{clay}))$ |
| Soil cover | RothC | Binary modifier on decomposition rate for bare soil, WON'T BE IMPLEMENTED!|

### 3.2.4 Hard-coded Static Modifiers
Both Q and Yasso have static modifiers dependent on size that are specific to litter classes. These are be hard-coded in each model function, rather than passed as separate modifiers.  

For the Yasso family of models, all size modifiers are hard-coded in the R wrappers




# 4. Decomposition Equations

This section presents the complete decomposition dynamics for each model

## 4.1 Yasso07

### Pool structure

Five pools: $\mathbf{C} = [C_A,\, C_W,\, C_E,\, C_N,\, C_H]^T$, representing Acid-soluble, Water-soluble, Ethanol-soluble, Non-soluble (lignin-like), and Humus.

### Dynamics

$$
\frac{d\mathbf{C}}{dt} = \mathbf{A}\,\mathbf{C} + \mathbf{I}
$$

where $\mathbf{I} = [I_A,\, I_W,\, I_E,\, I_N,\, 0]^T$ (no direct input to Humus).

**Diagonal elements** (loss from pool $i$):
$$A_{ii} = -\alpha_i \cdot \xi_T \cdot \xi_M$$

where $\alpha_i$ is the base decomposition rate. For the N pool, an additional size modifier applies to woody litter:
$$A_{44} = -\alpha_N \cdot \xi_T \cdot \xi_M \cdot f_{\text{size}}$$

with $f_{\text{size}} = \min\!\big(1,\; \exp(-\delta_1 \cdot d^{\delta_2} \cdot |r|)\big)$, where $d$ is the wood diameter (cm).

**Off-diagonal elements** (transfer from pool $j$ to pool $i$):
$$A_{ij} = p_{ij} \cdot |\alpha_j \cdot \xi_T \cdot \xi_M|$$

**Humus formation** (row 5, columns 1--4):
$$A_{5j} = p_H \cdot |\alpha_j \cdot \xi_T \cdot \xi_M| \quad \text{for } j \in \{A, W, E, N\}$$

**Respiration** is the residual: $R_h = \sum_j |A_{jj}| C_j - \sum_{i \neq j} A_{ij} C_j$.

### Environmental modifiers

**Temperature** — four-point approximation of intra-annual variation:
$$\xi_T = \frac{1}{4} \sum_{i=1}^{4} \exp\!\big(\beta_1 T_i + \beta_2 T_i^2\big)$$
where the quarterly temperatures are derived from annual mean $\bar{T}$ and amplitude $A_T$:
\begin{align*}
T_1 &= \bar{T} + \tfrac{4A_T}{\pi}\!\left(\tfrac{1}{\sqrt{2}} - 1\right), &
T_2 &= \bar{T} - \tfrac{4A_T}{\sqrt{2}\,\pi} \\
T_3 &= \bar{T} + \tfrac{4A_T}{\pi}\!\left(1 - \tfrac{1}{\sqrt{2}}\right), &
T_4 &= \bar{T} + \tfrac{4A_T}{\sqrt{2}\,\pi}
\end{align*}

**Precipitation:**
$$\xi_M = 1 - \exp\!\left(\gamma \cdot \frac{P}{1000}\right)$$
where $P$ is annual precipitation (mm).

### Analytical solution

For annual timestep $\Delta t = 1$\,yr:
$$\mathbf{C}(t+1) = e^{\mathbf{A}} \cdot \mathbf{C}(t) + \left(e^{\mathbf{A}} - \mathbf{I}_5\right) \mathbf{A}^{-1} \mathbf{I}$$
where $e^{\mathbf{A}}$ is the matrix exponential.

**Steady state:**
$$\mathbf{C}^* = -\mathbf{A}^{-1}\,\mathbf{I}$$

### Default parameters

| Parameter | Value | Description |
|:----------|:------|:------------|
| $\alpha_A,\, \alpha_W,\, \alpha_E,\, \alpha_N$ | 0.66, 4.30, 0.35, 0.22 | Base decomposition rates (yr$^{-1}$) |
| $\alpha_H$ | 0.0033 | Humus decomposition rate (yr$^{-1}$) |
| $p_H$ | 0.04 | Fraction to Humus |
| $\beta_1,\, \beta_2$ | 0.096, $-$0.0014 | Temperature response parameters |
| $\gamma$ | $-$1.21 | Precipitation response parameter |
| $\delta_1,\, \delta_2,\, r$ | $-$1.7, 0.86, $-$0.306 | Size modifier parameters |

\newpage

## 4.2 Yasso15

Yasso15 retains the AWENH pool structure and transfer matrix of Yasso07, but introduces two structural changes.

**Pool-group-specific climate responses.** Where Yasso07 applies a single temperature and precipitation modifier to all pools, Yasso15 separates the climate response into three groups — AWE, N, and H — each with independent parameters:

$$\xi_{T}^{\text{AWE}} = \frac{1}{4}\sum_{i=1}^{4}\exp\!\big(\beta_1 T_i + \beta_2 T_i^2\big), \quad \xi_{T}^{N} = \frac{1}{4}\sum_{i=1}^{4}\exp\!\big(\beta_{N1} T_i + \beta_{N2} T_i^2\big), \quad \xi_{T}^{H} = \frac{1}{4}\sum_{i=1}^{4}\exp\!\big(\beta_{H1} T_i + \beta_{H2} T_i^2\big)$$

Similarly for precipitation:

$$\xi_{M}^{\text{AWE}} = 1 - \exp\!\big(\gamma \cdot P/1000\big), \qquad \xi_{M}^{N} = 1 - \exp\!\big(\gamma_N \cdot P/1000\big), \qquad \xi_{M}^{H} = 1 - \exp\!\big(\gamma_H \cdot P/1000\big)$$

The combined climate modifier is applied per pool group when constructing the diagonal of $\mathbf{A}$:

\begin{align*}
A_{ii} &= -\alpha_i \cdot \xi_T^{\text{AWE}} \cdot \xi_M^{\text{AWE}} \cdot f_{\text{size}} & \text{for } i \in \{A, W, E\} \\
A_{44} &= -\alpha_N \cdot \xi_T^{N} \cdot \xi_M^{N} \cdot f_{\text{size}} \\
A_{55} &= -\alpha_H \cdot \xi_T^{H} \cdot \xi_M^{H} & \text{(no size effect)}
\end{align*}

Off-diagonal elements and humus formation follow the same structure as Yasso07, with transfers proportional to the source pool's modified rate.

**Leaching.** Yasso15 adds a precipitation-dependent leaching term to the AWEN pools (not H):

$$A_{ii} \mathrel{+}= w \cdot P/1000 \quad \text{for } i \in \{A, W, E, N\}$$

The size modifier, transfer matrix structure, analytical solution (matrix exponential), and steady-state formula ($\mathbf{C}^* = -\mathbf{A}^{-1}\mathbf{b}$) are identical to Yasso07.

\newpage

## 4.3 Yasso20

Yasso20 shares the AWENH pool structure, pool-group-specific climate responses, and leaching of Yasso15. The structural equations are the same. The principal differences are in parameterisation and the size modifier formula.

**Modified size dependence.** Yasso07/15 use an exponential form:

$$f_{\text{size}}^{\text{07/15}} = \min\!\big(1,\; \exp(-\delta_1 \cdot d^{\delta_2} \cdot |r|)\big)$$

Yasso20 uses a quadratic formulation:

$$f_{\text{size}}^{\text{20}} = \big(1 + \theta_1\, d + \theta_2\, d^2\big)^{-|r|}$$

which produces a stronger size-dependent slowdown for coarse woody debris.

**Substantially different parameter values.** The most consequential changes from Yasso07/15 are:

| Parameter | Yasso07 | Yasso20 | Change | Effect |
|:----------|:--------|:--------|:-------|:-------|
| $p_H$ | 0.04 | 0.0042 | $-$90\% | Much less humus formation |
| $\alpha_H$ | 0.0033 | 0.0015 | $-$54\% | Slower humus decomposition |
| $\alpha_W$ | 4.30 | 5.19 | $+$21\% | Faster W-pool turnover |
| $\alpha_E$ | 0.35 | 0.13 | $-$63\% | Slower E-pool turnover |
| $\alpha_N$ | 0.22 | 0.10 | $-$55\% | Slower N-pool turnover |

The dramatically reduced $p_H$ (9.5$\times$ less humus formation) is the primary driver of Yasso20 predicting $\sim$20\% lower total SOC than Yasso07/15 for Finnish forests, despite the compensating effect of slower humus decomposition.


## 4.4 RothC

### Pool structure

Five pools: $\mathbf{C} = [C_{\text{DPM}},\, C_{\text{RPM}},\, C_{\text{BIO}},\, C_{\text{HUM}},\, C_{\text{IOM}}]^T$, representing Decomposable Plant Material, Resistant Plant Material, Microbial Biomass, Humified Organic Matter, and Inert Organic Matter.

### Dynamics

RothC operates on a **monthly timestep** $\Delta t = 1/12$\,yr. For each active pool $i \in \{\text{DPM}, \text{RPM}, \text{BIO}, \text{HUM}\}$:

$$C_i(t + \Delta t) = C_i(t) \cdot \exp\!\big(-k_i \cdot r(t) \cdot \Delta t\big) + I_i \cdot \Delta t + \text{transfers}_i$$

where $r(t) = \xi_T(t) \cdot \xi_M(t) \cdot \xi_V$ is the composite rate modifier (normalised so that $\xi_T / 47.9$ gives the fractional effect).

The IOM pool is inert: $C_{\text{IOM}}(t) = 0.049 \cdot (\%\text{clay})^{1.139}$.

**Decomposed material** from each pool:
$$D_i = C_i \cdot \big(1 - \exp(-k_i \cdot r \cdot \Delta t)\big)$$

This is partitioned into three fluxes governed by the clay-dependent ratio $x$:

$$x = 1.67\big(1.85 + 1.60\,\exp(-0.0786 \cdot \%\text{clay})\big)$$

\begin{align*}
f_{\text{CO}_2} &= x / (x + 1) \\
f_{\text{BIO}} &= 0.46 / (x + 1) \\
f_{\text{HUM}} &= 0.54 / (x + 1)
\end{align*}

So the transfers are:
\begin{align*}
\text{CO}_{2} &= \sum_i D_i \cdot f_{\text{CO}_2} \\
C_{\text{BIO}} &\mathrel{+}= \sum_i D_i \cdot f_{\text{BIO}} \\
C_{\text{HUM}} &\mathrel{+}= \sum_i D_i \cdot f_{\text{HUM}}
\end{align*}

Only DPM and RPM receive external inputs: $I_{\text{DPM}},\, I_{\text{RPM}}$ (partitioned from total litter via the DPM/RPM ratio).

### Environmental modifiers

**Temperature** (monthly):
$$\xi_T = \frac{47.9}{1 + \exp\!\big(106\,/\,(T + 18.3)\big)}$$

**Moisture** — based on accumulated topsoil moisture deficit (TSMD):
$$\text{TSMD}_m = \max\!\big(0,\; E_m - P_m\big)$$

The moisture modifier reduces decomposition when soil dries:
$$\xi_M = \max\!\big(0.2,\; 1 - \text{TSMD} / \text{TSMD}_{\max}\big)$$

where $\text{TSMD}_{\max}$ depends on clay content.

**Vegetation cover:**
$$\xi_V = \begin{cases} 0.6 & \text{vegetated (all forest plots)} \\ 1.0 & \text{bare soil} \end{cases}$$

### Analytical steady state

The system can be expressed as $\mathbf{A}\,\mathbf{C}^* = -\mathbf{b}$ for the four active pools, where $\mathbf{A}$ is the $4 \times 4$ matrix encoding losses and transfers, and $\mathbf{b} = [I_{\text{DPM}},\, I_{\text{RPM}},\, 0,\, 0]^T$.

$$\mathbf{C}^*_{\text{active}} = -\mathbf{A}^{-1}\,\mathbf{b}$$

The full steady state appends $C_{\text{IOM}}$ calculated from clay content.

### Default parameters

| Parameter | Value | Description |
|:----------|:------|:------------|
| $k_{\text{DPM}}$ | 10.0 | DPM decomposition rate (yr$^{-1}$) |
| $k_{\text{RPM}}$ | 0.3 | RPM decomposition rate (yr$^{-1}$) |
| $k_{\text{BIO}}$ | 0.66 | BIO decomposition rate (yr$^{-1}$) |
| $k_{\text{HUM}}$ | 0.02 | HUM decomposition rate (yr$^{-1}$) |

\newpage

## 4.5 Q-model

### Pool structure

Unlike the discrete-pool models above, the Q-model (Bosatta \& Ågren, 1985) describes decomposition along a **continuous quality axis** $q$. Organic matter is characterised by its current quality $q \in (0, q_0]$, where $q_0$ is the initial quality at litter input. As material decomposes, quality declines and decomposition slows — encoding the universal observation that fresh litter decomposes faster than aged organic matter.

In practice, the model is discretised into **cohort tracking**: each annual litter input forms a cohort whose remaining mass is tracked as a function of age and tissue type.

Five tissue classes are tracked separately, distinguished by initial quality and entry dynamics:

| Tissue class | Quality parameter | Entry dynamics |
|:-------------|:-----------------|:---------------|
| Foliage (needles) | $q_{0n}$ (species-specific) | Immediate |
| Fine roots | $q_{0f}$ (species-specific) | Immediate |
| Understorey | $q_{0n}$ (as foliage) | Immediate |
| Branches | $q_{0w}$ (species-specific) | Gradual fragmentation ($t_{\max,b}$ = 13 yr) |
| Stems | $q_{0w}$ (species-specific) | Gradual fragmentation ($t_{\max,s}$ = 60 yr) |

### Core equations

**Microbial uptake rate** as a function of quality:
$$u(q) = u_0 \cdot q^{\beta}$$

where $u_0$ is the temperature-dependent base rate and $\beta$ is the quality--rate exponent (default $\beta = 7$).

**Temperature dependence** (linear form):
$$u_0(T) = u_{00} + u_{01} \cdot T$$

where $T$ is the mean annual temperature (°C), $u_{00} = 0.0855$, and $u_{01} = 0.0157$.

An alternative thermodynamic (Arrhenius) formulation (not implemented here) exists (Bosatta \& Ågren 1999):
$$u(q, T) = u_0^{\text{ref}} \cdot \exp\!\left(\frac{|\Delta G^0|}{R\,q}\left(\frac{1}{T_{\text{ref}}} - \frac{1}{T}\right)\right)$$

where $T$ is in Kelvin, $R = 8.314$\,J\,mol$^{-1}$\,K$^{-1}$, and $\Delta G^0$ is the free energy change per enzymatic step. This formulation implies that low-quality substrates have stronger temperature sensitivity than high-quality substrates — a feature not captured by the linear form.

**Derived parameters.** The quality decline exponent $z$ controls how decomposition slows over time:
$$z = \frac{1 - e_0}{\beta \cdot \eta_{11} \cdot e_0}$$

where $e_0$ is the initial microbial growth efficiency (0.25) and $\eta_{11}$ is the carbon use efficiency parameter (0.36). With default parameters, $z \approx 1.19$.

The composite decomposition rate $\alpha$ for a given tissue type combines all factors:
$$\alpha = f_C \cdot \beta \cdot \eta_{11} \cdot u_0 \cdot q_0^{\beta}$$

where $f_C = 0.5$ is the carbon fraction of organic matter.

### Remaining fraction functions

**Simple tissues** (foliage, fine roots, understorey) enter the soil immediately and follow power-law decay:
$$g(t) = \frac{1}{(1 + \alpha\, t)^z}$$

**Branches** fragment gradually over $t_{\max,b}$ years. The remaining fraction integrates decomposition over the fragmenting material. Defining $i_b = \min(t,\, t_{\max,b})$:

$$g_b(t) = \frac{2\Big[(1 + \alpha t)^{1-z} - (1 + \alpha(t - i_b))^{1-z}(1 - i_b/t_{\max,b})\Big]}{\alpha\,(1-z)\,t_{\max,b}} + \frac{2\Big[(1 + \alpha(t-i_b))^{2-z} - (1+\alpha t)^{2-z}\Big]}{\alpha^2(1-z)(2-z)\,t_{\max,b}^2} + \left(1 - \frac{i_b}{t_{\max,b}}\right)^2$$

**Stems** follow the same structure with $t_{\max,s}$ replacing $t_{\max,b}$.

### Quality evolution

Quality declines with mass loss:
$$q(t) = q_0 \left(\frac{C(t)}{C(0)}\right)^{1/z}$$

This couples quality and mass: as carbon is lost, the remaining material becomes progressively more recalcitrant.

### Transient temperature implementation

The standard Q-model analytical solution assumes constant $\alpha$ (i.e.\ constant temperature). Under transient climate, we track **cumulative decomposition potential** for each cohort:

$$\tilde{\alpha}(t) = \sum_{\tau=1}^{t} \alpha\big(T(\tau)\big)$$

where $\alpha(T(\tau))$ is recalculated each year from the current temperature. The remaining fraction then becomes:

$$g(t) = \frac{1}{\big(1 + \tilde{\alpha}(t)\big)^z}$$

This ensures all carbon pools respond to current temperature conditions rather than being locked to the temperature at the time of input — a critical requirement for climate change projections. Without this correction, the model underestimates SOC loss under warming by 20--40\%.

### Initialisation: hybrid approach

Rather than running a 1500-year numerical spinup, we use an analytical-hybrid approach:

1. **Analytical steady state.** For simple tissues:
$$C_{\text{ss}} = L \cdot \sum_{t=1}^{T_{\max}} g(t)$$
where $L$ is the annual litter input. Branches and stems use numerical integration of their respective $g_b$, $g_s$ functions over 500--1000 years.

2. **Bulk pool initialisation.** The total $C_{\text{ss}}$ is treated as a single bulk pool that decomposes via:
$$C_{\text{init}}(t) = C_{\text{ss}} \cdot (1 + \alpha_0\, t)^{1-z}$$

3. **Short spinup** (10--20 years) builds up explicit cohort matrices while the bulk pool decays. Total SOC = bulk pool + new cohorts.

This reduces initialisation time from $\sim$1500 years to 10--20 years while maintaining accuracy.

### Default parameters

| Parameter | Value | Description |
|:----------|:------|:------------|
| $\beta$ | 7 | Quality--rate exponent |
| $e_0$ | 0.25 | Initial microbial growth efficiency |
| $\eta_{11}$ | 0.36 | Carbon use efficiency parameter |
| $f_C$ | 0.5 | Carbon fraction of organic matter |
| $u_{00}$ | 0.0855 | Base uptake rate at 0°C |
| $u_{01}$ | 0.0157 | Temperature sensitivity of uptake |
| $t_{\max,b}$ | 13 | Branch fragmentation time (yr) |
| $t_{\max,s}$ | 60 | Stem fragmentation time (yr) |
| $q_{0n}$ | 1.01 (spruce) | Initial quality, foliage |
| $q_{0f}$ | 1.036 | Initial quality, fine roots |
| $q_{0w}$ | 1.00 | Initial quality, woody tissues |





## 4.6 Common Interface and Implementation Architecture

All five decomposition models share a common calling interface, enabling the simulation loop and Bayesian calibration framework to treat them interchangeably. The implementation is split across two languages: performance-critical numerical code in **Fortran**, and data handling, calibration logic, and input preparation in **R**.

### Fortran--R boundary

The guiding principle is to minimise R$\leftrightarrow$Fortran call overhead during MCMC sampling. Since the calibration loop evaluates the likelihood millions of times across $\sim$3,700 plots, each evaluation should require a **single Fortran call per plot** that runs the full time-series simulation, rather than hundreds of per-timestep round-trips.

\begin{figure}[H]
\centering
\begin{tikzpicture}[
    >=Stealth,
    node distance=0.8cm and 1.6cm,
    box/.style={
        rectangle, rounded corners=4pt, draw=#1, fill=#1!8,
        line width=1pt, minimum height=1.0cm, minimum width=3.4cm,
        text centered, font=\small\sffamily
    },
    widebox/.style={
        rectangle, rounded corners=4pt, draw=#1, fill=#1!8,
        line width=1pt, minimum height=1.0cm, minimum width=5.0cm,
        text centered, font=\small\sffamily
    },
    data/.style={
        rectangle, rounded corners=2pt, draw=bordergray, fill=bglight,
        line width=0.6pt, minimum height=0.6cm, minimum width=2.0cm,
        text centered, font=\footnotesize\ttfamily
    },
    arr/.style={->, line width=1pt, color=#1},
    note/.style={font=\scriptsize\sffamily\itshape, text=black!50},
    layerlabel/.style={font=\small\sffamily\bfseries, text=#1, rotate=90, anchor=south}
]

% =============================================
% OUTSIDE-MCMC
% =============================================

\node[box=compatgreen] (fi) at (-4.0, 2.5) {%
    \begin{tabular}{c}
    $f_{i_k}(\mathbf{I}(t))$ \\ Input layer
    \end{tabular}};
\node[note, below=0.05cm of fi, text width=2.8cm, align=center]
    {once per plot};

\node[box=compatgreen] (SOC) at (3.0, 2.5) {%
    \begin{tabular}{c}
    SOC measurements
    \end{tabular}};
\node[note, below=0.05cm of SOC, text width=2.8cm, align=center]
    {once per plot};

\node[box=compatgreen] (envdata) at (6.5, 2.5) {%
    \begin{tabular}{c}
    $g_{i_k}(T(t),\,M(t))$ \\ Climate layer
    \end{tabular}};
\node[note, below=0.05cm of envdata, text width=2.8cm, align=center]
    {once per plot};

% =============================================
% INSIDE-MCMC: Likelihood
% =============================================

\node[widebox=outputpurple] (likelihood) at (3, 0.0) {%
    \begin{tabular}{c}
    Likelihood \\ $\mathcal{L}(\boldsymbol{\theta} \,|\, \text{SOC}_{\text{obs}})$
    \end{tabular}};

% =============================================
% INSIDE-MCMC: R wrapper components
% =============================================

% steady_state (Fortran, red) — leftmost
\node[box=decompred] (steadyblock) at (-4.0, -3.0) {%
    \begin{tabular}{c}
    \texttt{steady\_state()} \\
    $\mathbf{C}^* = -\mathbf{A}^{-1}\mathbf{b}$
    \end{tabular}};

% model_run (Fortran, red) — middle
\node[
    rectangle, rounded corners=4pt,
    draw=decompred, fill=decompred!8,
    line width=1pt,
    minimum height=3.0cm, minimum width=4.2cm,
    text centered, font=\small\sffamily
] (frun) at (1.0, -3.0) {};

\node[font=\small\sffamily, text=decompred, anchor=north]
    at (frun.north) [yshift=-6pt] {%
    \begin{tabular}{c}
    \texttt{model\_run()} \\ loop over $c = 1,\ldots,n_c$
    \end{tabular}};

\node[box=decompred, anchor=center] (fstep) at (1.0, -3.6) {%
    \begin{tabular}{c}
    \texttt{model\_step()} \\ single $\Delta t$
    \end{tabular}};

% compute_xi (R, gold) — right of model_run
\node[box=modifiergold] (fxi) at (6.5, -3.0) {%
    \begin{tabular}{c}
    \texttt{compute\_xi()} \\ $\boldsymbol{\xi}(T, M, \boldsymbol{\theta})$
    \end{tabular}};

% Output
\node[data] (output) at (-1.5, -6.5) {SOC$(t)$ vector};

% =============================================
% GROUPING BOXES (background layer)
% =============================================
\begin{scope}[on background layer]

\node[
    rectangle, rounded corners=8pt,
    draw=outputpurple, fill=outputpurple!4,
    line width=1.5pt, dashed,
    fit=(likelihood)(steadyblock)(frun)(fxi)(output),
    inner sep=24pt
] (mcmcbox) {};

\node[
    rectangle, rounded corners=6pt,
    draw=compatgreen, fill=compatgreen!4,
    line width=1.2pt, dashed,
    fit=(steadyblock)(frun)(fxi)(output),
    inner sep=14pt
] (rwrapbox) {};

\end{scope}

% Labels
\node[font=\small\sffamily\bfseries, text=compatgreen,
    anchor=south west] at (rwrapbox.north west)
    {R wrapper};
\node[font=\small\sffamily\bfseries, text=outputpurple,
    below=0.4cm of mcmcbox.south, anchor=south]
    {MCMC sampler};
\node[note, below=0.05cm of mcmcbox.north, anchor=north,
    text width=5cm, align=center]
    {re-evaluated each proposal $\boldsymbol{\theta}$};

% =============================================
% ARROWS — outside to inside
% =============================================

% SOC -> likelihood
\draw[arr=compatgreen] (SOC.south) -- (likelihood.north)
    node[midway, right=3pt, note] {$\text{SOC}_{\text{obs}}$};

% Input layer -> steady_state
\draw[arr=compatgreen] (fi.south) -- (steadyblock.north)
    node[midway, left=3pt, note] {$\mathbf{J}_c(t),\,\bar{\mathbf{J}}$};

% Input layer -> model_run
\draw[arr=compatgreen] (fi.south) -- ++(0,-1.0) -| (frun.north)
    node[near end, right=3pt, note] {$\mathbf{J}_c(t)$};

% Climate layer -> compute_xi only
\draw[arr=compatgreen] (envdata.south) -- (fxi.north)
    node[midway, right=3pt, note] {$T(t),\,M(t)$};

% =============================================
% ARROWS — inside MCMC
% =============================================

% compute_xi -> Likelihood
\draw[arr=outputpurple] (likelihood.east) -- ++(0.7,0) -- (fxi.119)
    node[near end, left=2pt, note] {$\boldsymbol{\theta}$};

% compute_xi -> steady_state
\draw[arr=modifiergold] (fxi.south) -- ++(0,-1.8) -| (steadyblock.south)
    node[midway, below=2pt, note] {$\bar{\xi}$};
    
% Likelihood -> steady_state
\draw[arr=outputpurple] (likelihood.south) -- ++(0,-0.3) -| (steadyblock.110)
    node[near end, left=2pt, note] {$\boldsymbol{\theta}$};

% Likelihood -> model_run
\draw[arr=outputpurple] (likelihood.south) -- ++(0,-0.3) -| (frun.110)
    node[near end, left=2pt, note] {$\boldsymbol{\theta}$};


% compute_xi -> model_run
\draw[arr=modifiergold] (fxi.west) -- (frun.east)
    node[midway, above=2pt, note] {$\boldsymbol{\xi}(t)$};

% steady_state -> model_run
\draw[arr=decompred] (steadyblock.east) -- (frun.west)
    node[midway, above=2pt, note] {$\mathbf{C}^*$};

% model_run -> output
\draw[arr=outputpurple] (frun.south) -- ++(0,-0.5) |- (output.east);

% output -> likelihood
\draw[arr=outputpurple, line width=1.5pt]
    (output.west) -- ++(-3.8,0) |- (likelihood.west)
    node[near start, above=-30pt, note] {SOC predictions};

\end{tikzpicture}
\caption{%
Revised implementation architecture.
Three blocks outside the MCMC loop are computed once per plot:
the input layer $f_{i_k}$ maps raw litter data to model-specific arrays $\mathbf{J}_c(t)$
and mean inputs $\bar{\mathbf{J}}$; the climate layer $g_{i_k}$ aggregates raw climate
to the resolution required by each model; and SOC observations feed the likelihood.
At each MCMC proposal $\boldsymbol{\theta}$, the likelihood dispatches parameters
to the \textbf{R wrapper}, which runs three components:
\texttt{steady\_state()} (Fortran, red) receives $\boldsymbol{\theta}$ and $\bar{\mathbf{J}}$
and initialises pools $\mathbf{C}^*$;
\texttt{compute\_xi()} (R, gold) receives $\boldsymbol{\theta}$ and the pre-computed
climate series to produce $\boldsymbol{\xi}(t)$;
and \texttt{model\_run()} (Fortran, red) receives $\mathbf{C}^*$, $\boldsymbol{\xi}(t)$,
$\boldsymbol{\theta}$, and $\mathbf{J}_c(t)$, iterates \texttt{model\_step()} over all
cohorts and timesteps, and returns the SOC trajectory for likelihood evaluation.}
\label{fig:implementation}
\end{figure}


### Fortran module structure

Each model is implemented as a Fortran module exposing three subroutines. All share the same calling conventions.

**\texttt{model\_run()}** — the top-level entry point called from R via \texttt{.Fortran()}. It receives the full pre-processed input arrays, parameter vector, and climate time series for a single plot, and returns the complete SOC trajectory.

```{r, eval=FALSE}
# R side: single call per plot
result <- .Fortran("yasso15_run",
  n_years    = as.integer(n_years),
  params     = as.double(param_vector),    # calibrated parameters
  J_t        = as.double(input_matrix),    # [n_years x n_input_pools]
  T_monthly  = as.double(temp_matrix),     # [n_years x 12]
  P_monthly  = as.double(precip_matrix),   # [n_years x 12]
  diameter   = as.double(wood_diameters),  # [n_litter_types]
  C_out      = double(n_years * n_pools),  # output: [n_years x n_pools]
  resp_out   = double(n_years)             # output: annual respiration
)
```

Internally, \texttt{model\_run()} performs:

1. Call \texttt{model\_steady\_state()} to compute $\mathbf{C}^*$ from mean climate and inputs.
2. Initialise pools from $\mathbf{C}^*$ (or from the hybrid approach for Q-model).
3. Loop over years (and months for RothC), calling \texttt{compute\_xi()} and \texttt{model\_step()} at each timestep.
4. Write annual pool states and respiration to output arrays.

**\texttt{model\_steady\_state()}** — computes the analytical equilibrium. For Yasso and RothC, this is a matrix inversion $\mathbf{C}^* = -\mathbf{A}^{-1}\mathbf{b}$. For the Q-model, it performs the numerical summation of remaining fractions over a truncated age series.

**\texttt{model\_step()}** — advances the pool state by one timestep $\Delta t$:

| Model | $\Delta t$ | Core operation |
|:------|:-----------|:---------------|
| Yasso07/15/20 | 1 yr | Build $\mathbf{A}$, compute matrix exponential $e^{\mathbf{A}}$, update $\mathbf{C}$ |
| RothC | 1/12 yr | Exponential decay per pool, partition decomposed C via clay ratio $x$ |
| Q-model | 1 yr | Update cumulative $\tilde{\alpha}$ per cohort, compute remaining fraction, prune old cohorts |

**\texttt{compute\_xi()}** — evaluates the environmental modifiers from climate data and static site properties. Called internally by \texttt{model\_step()} (or once per year for annual models, 12 times for RothC).


### R wrapper layer

The R wrapper for each model handles:

- **Climate mapping**: calls `map_climate_<model>()` once per plot to produce the climate data frame expected by that model. Yasso07/15 use `map_climate_yasso07()` (annual mean + amplitude); Yasso20 and RothC share `map_climate_monthly()` (full monthly series).
- **Input mapping**: calls `map_inputs_<model>()` once per plot to aggregate the six raw cohorts into model-specific litter arrays. Yasso07, Yasso15 and Yasso20 share the same input structure via `map_inputs_yasso07()`.
- **Parameter mapping**: translates the named BayesianTools parameter vector into the positional double-precision vector expected by Fortran.
- **Output formatting**: reshapes the flat Fortran output arrays into named R structures for likelihood evaluation and diagnostics.

All models are run plot-by-plot with a consistent `lapply` → `rbind` pattern. Climate and input mapping are performed once per plot, outside the MCMC loop:

```{r, eval=FALSE}
# --- Once, before calibration ---
plots          <- unique(input_raw$plot_id)
Yasso07_climate <- map_climate_yasso07(input_raw)   # annual: mean, amplitude, precip
monthly_climate <- map_climate_monthly(input_raw)   # monthly: shared by Yasso20 + RothC
Yasso07_inputs  <- map_inputs_yasso07(input_raw)    # shared by Yasso07, Yasso15, Yasso20
RothC_inputs    <- map_inputs_rothc(input_raw, site_raw)

# --- Per-plot simulation (inside likelihood, called each MCMC proposal) ---
result <- do.call(rbind, lapply(plots, function(pid) {

  clim   <- Yasso07_climate[Yasso07_climate$plot_id == pid, ]
  inputs <- Yasso07_inputs[Yasso07_inputs$plot_id   == pid, ]

  # Stage 1: Xi time series from plot-specific climate
  xi <- compute_xi_yasso07(clim, params)

  # Stage 2: Steady-state initialisation from plot-specific means
  C_init <- yasso07_steady_state(
    params   = params,
    nwl_mean = colMeans(inputs[, c("nwl_A","nwl_W","nwl_E","nwl_N")]),
    fwl_mean = colMeans(inputs[, c("fwl_A","fwl_W","fwl_E","fwl_N")]),
    cwl_mean = colMeans(inputs[, c("cwl_A","cwl_W","cwl_E","cwl_N")]),
    xi_mean  = mean(xi)
  )

  # Stage 3: Transient simulation — single Fortran call per plot
  out <- yasso07_run(inputs, params, C_init, xi)

  cbind(plot_id = pid, out)
}))
```


### What stays in R vs.\ Fortran

| Component | Language | Rationale |
|:----------|:---------|:----------|
| Climate mapping (`map_climate_*`) | R | Called once per plot; benefits from R data wrangling |
| Input mapping (`map_inputs_*`) | R | Called once per plot; benefits from R data wrangling |
| $\xi_k$ — modifier computation | Fortran | Called millions of times inside MCMC loop |
| $f_{d_k}$ — decomposition step | Fortran | Innermost loop; matrix operations |
| Steady-state solver | Fortran | Matrix inversion, numerical integration |
| Full simulation loop | Fortran | Avoids per-timestep R$\leftrightarrow$Fortran overhead |
| Parameter mapping | R | String/name handling; negligible cost |
| Likelihood evaluation | R | Interface with BayesianTools; simple arithmetic |
| I/O, aggregation, diagnostics | R | Flexibility; not performance-critical |

\newpage

## 4.7 Yasso Family: Three-Stage Call Pattern

All three Yasso versions (07, 15, 20) follow the same three-stage execution pattern in R. The differences are confined to Stage 1 (Xi computation) and the climate object passed to each stage.

### Climate and input mapping functions

| Model | Climate function | Input function |
|:------|:----------------|:---------------|
| Yasso07 | `map_climate_yasso07()` → annual mean, amplitude, precip | `map_inputs_yasso07()` |
| Yasso15 | `map_climate_yasso07()` (reused) | `map_inputs_yasso07()` (reused) |
| Yasso20 | `map_climate_monthly()` (shared with RothC) | `map_inputs_yasso07()` (reused) |

All three versions share the same litter input structure: six raw cohorts aggregated to three litter types (nwl, fwl, cwl) with AWEN fractions preserved.

### Stage 1 — Xi: climate modifier time series

The key structural difference among the three versions is in how $\xi$ is computed:

| Model | Function | Climate resolution | Pool-group specificity |
|:------|:---------|:-------------------|:-----------------------|
| Yasso07 | `compute_xi_yasso07()` | Annual (mean + amplitude → 4-point Gaussian) | Single $\xi$ for all pools |
| Yasso15 | `compute_xi_yasso15()` | Annual (mean + amplitude → 4-point Gaussian) | Three $\xi$: AWE, N, H |
| Yasso20 | `compute_xi_yasso20()` | Monthly (exact 12-month average) | Three $\xi$: AWE, N, H |

### Stage 2 — Steady state: initial pool conditions $\mathbf{C}^*$

Steady state is computed from plot-specific mean climate and mean litter inputs. For Yasso20, a synthetic 12-row `mean_clim` data frame is constructed from the plot's monthly climatology:

```{r, eval=FALSE}
# Yasso07 / Yasso15
C_init <- yassoXX_steady_state(
  params   = params,
  nwl_mean = colMeans(inputs[, c("nwl_A","nwl_W","nwl_E","nwl_N")]),
  fwl_mean = colMeans(inputs[, c("fwl_A","fwl_W","fwl_E","fwl_N")]),
  cwl_mean = colMeans(inputs[, c("cwl_A","cwl_W","cwl_E","cwl_N")]),
  xi_mean  = mean(xi)   # or list(xi_awe, xi_n, xi_h) for Yasso15
)

# Yasso20 — requires a synthetic single-year mean climate data frame
# precip must be monthly mean (NOT divided by 12); the function sums
# internally over 12 months to recover the annual total
mean_clim <- data.frame(
  year     = 9999L,
  month    = 1:12,
  temp_air = tapply(clim$temp_air, clim$month, mean, na.rm = TRUE),
  precip   = mean(clim$precip)   # monthly mean ~50 mm; sums to ~600 mm/yr internally
)
C_init <- yasso20_steady_state(mean_clim, params, nwl_mean, fwl_mean, cwl_mean)
```

### Stage 3 — Transient simulation

All three versions return a flat data frame with columns `year | A | W | E | N | H | total_soc | respiration`.

```{r, eval=FALSE}
out_07 <- yasso07_run(inputs_df,    params, C_init, xi_array)
out_15 <- yasso15_run(inputs_df,    params, C_init, xi_arrays, precip, leac)
out_20 <- yasso20_run(climate_df,   inputs_df, params, C_init, leac)
#                     ^^^^^^^^^^
#                     monthly_climate — xi_arrays and precip derived internally
```



\newpage

# 5. Calibration Parameter Comparison

This section documents the calibration parameter vectors for all four models,
their grouping by functional role, and the structural equivalences and differences
across models.

## 5.1 Cross-model parameter map

The table below classifies calibration parameters by functional group. Each cell
lists the parameter symbol(s) and count. Dashes indicate that the corresponding
process is absent or hardcoded in that model.

\definecolor{y07col}{HTML}{DBEAFE}
\definecolor{y15col}{HTML}{D1FAE5}
\definecolor{y20col}{HTML}{FEF3C7}
\definecolor{allcol}{HTML}{F1F5F9}

\begingroup
\small
\setlength{\tabcolsep}{5pt}
\begin{table}[H]
\centering
\caption{Calibration parameters by functional group across all four models.
Numbers in parentheses give the count of free parameters in that group.}
\label{tab:param_map}
\begin{tabular}{lp{2.6cm}p{2.6cm}p{2.6cm}p{2.6cm}}
\toprule
\textbf{Group} & \textbf{Yasso07} & \textbf{Yasso15} & \textbf{Yasso20} & \textbf{RothC} \\
\midrule
Decomposition rates
  & $\alpha_{A,W,E,N,H}$ (5)
  & $\alpha_{A,W,E,N,H}$ (5)
  & $\alpha_{A,W,E,N,H}$ (5)
  & $k_{\text{DPM,RPM,Bio,Hum}}$ (4) \\
\addlinespace
Temperature response
  & $\beta_1, \beta_2$ (2)
  & $\beta_1,\beta_2$;\newline $\beta_{N1},\beta_{N2}$;\newline $\beta_{H1},\beta_{H2}$ (6)
  & same as Y15 (6)
  & $a_T, b_T, c_T$ (3) \\
\addlinespace
Moisture / precipitation response
  & $\gamma$ (1)
  & $\gamma,\,\gamma_N,\,\gamma_H$ (3)
  & $\gamma,\,\gamma_N,\,\gamma_H$ (3)
  & $r_{\text{SMD}}$ (1) \\
\addlinespace
Inter-pool transfers
  & $p_{ij}$ (12)
  & $p_{ij}$ (12)
  & $p_{ij}$ (12)
  & $f_{\text{bio}},\, f_{\text{hum}}$ (2) \\
\addlinespace
Humification
  & $p_H$ (1)
  & $p_H$ (1)
  & $p_H$ (1)
  & via $f_{\text{bio}}, f_{\text{hum}}$ \\
\addlinespace
Size / physical modifier
  & $\delta_1, \delta_2, r$ (3)
  & $\delta_1, \delta_2, r$ (3)
  & $\theta_1, \theta_2, r$ (3)
  & --- (absent) \\
\addlinespace
Leaching
  & --- (absent)
  & $w$ (1)
  & $w$ (1)
  & --- (absent) \\
\midrule
\textbf{Total}
  & \textbf{24}
  & \textbf{31}
  & \textbf{31}
  & \textbf{10} \\
\bottomrule
\end{tabular}
\end{table}
\endgroup

\medskip
\noindent\textbf{Structural notes.}
The Yasso family has no edaphic modifiers (clay, depth) as free parameters;
these enter only through fixed litter-size correction terms.

Regarding inter-pool transfers and humification, RothC and Yasso follow quite
different philosophies. In RothC, all four active pools (DPM, RPM, Bio, Hum)
decompose and their products are routed to the same two destinations ---
microbial biomass (Bio) and humus (Hum) --- in fixed proportions governed by
$f_{\text{bio}}$ and $f_{\text{hum}}$. The source pool identity is irrelevant:
a unit of carbon leaving DPM and a unit leaving Hum are partitioned identically.
This makes the transfer structure simple and symmetric.

In Yasso, the four litter-chemistry pools (A, W, E, N) exchange carbon
\emph{with each other} before anything reaches the humus pool H. Each pool $j$
has its own set of transfer fractions $p_{ij}$ controlling what fraction of its
decomposed material flows into each other pool $i$. For example, decomposed
W-pool carbon might partially re-enter the A pool, partially enter E, and so
on. These 12 fractions ($p_{WA}, p_{EA}, \ldots, p_{EN}$) together define the
full AWEN transfer matrix, and they are all free calibration parameters. On top
of this lateral exchange, a separate parameter $p_H$ controls the fraction of
decomposed carbon from \emph{any} AWEN pool that is additionally routed to the
humus pool H --- analogous in spirit to RothC's $f_{\text{hum}}$, but applied
uniformly across all source pools. The net effect is that Yasso's pool dynamics
are governed by a full $5 \times 5$ matrix, whereas RothC's are governed by a
simpler one-directional cascade with a shared partitioning step at the end.
RothC's $f_{\text{bio}}$ and $f_{\text{hum}}$ therefore simultaneously play the
role of both the inter-pool transfer coefficients \emph{and} the humification
fraction, compressed into just two parameters.


## 5.2 Yasso family parameter table

The three Yasso versions share the same pool structure and transfer matrix.
Differences are confined to the climate response parameterisation (Yasso15/20
add pool-group-specific modifiers) and the size modifier functional form
(Yasso20 uses a quadratic rather than exponential formulation).

\begingroup
\small
\setlength{\tabcolsep}{5pt}
\begin{table}[H]
\centering
\caption{Yasso family calibration parameters.
\colorbox{allcol}{\strut Gray}: shared by Y07, Y15, Y20.
\colorbox{y15col}{\strut Green}: introduced in Y15, retained in Y20.
\colorbox{y20col}{\strut Amber}: Y20-specific (replaces Y07/Y15 formulation).
All $\alpha$ rates are positive; the Fortran negates them internally on the diagonal.}
\label{tab:yasso_params}
\begin{tabular}{llllp{5.5cm}}
\toprule
\textbf{Group} & \textbf{Symbol} & \textbf{Default} & \textbf{Units} & \textbf{Description} \\
\midrule
\multicolumn{5}{l}{\textit{Decomposition rates — shared by Y07, Y15, Y20}} \\
\rowcolor{allcol}
& $\alpha_A$ & 0.66 & yr$^{-1}$ & Base decomp.\ rate, acid-soluble (A) pool \\
\rowcolor{allcol}
& $\alpha_W$ & 4.30 & yr$^{-1}$ & Base decomp.\ rate, water-soluble (W) pool \\
\rowcolor{allcol}
& $\alpha_E$ & 0.35 & yr$^{-1}$ & Base decomp.\ rate, ethanol-soluble (E) pool \\
\rowcolor{allcol}
& $\alpha_N$ & 0.22 & yr$^{-1}$ & Base decomp.\ rate, non-soluble/lignin (N) pool \\
\rowcolor{allcol}
& $\alpha_H$ & 0.0033 & yr$^{-1}$ & Base decomp.\ rate, humus (H) pool \\
\midrule
\multicolumn{5}{l}{\textit{Inter-pool transfers — shared by Y07, Y15, Y20}} \\
\rowcolor{allcol}
& $p_{WA},\, p_{EA},\, p_{NA}$ & & — & Fractions of W, E, N decomposition entering A \\
\rowcolor{allcol}
& $p_{AW},\, p_{EW},\, p_{NW}$ & & — & Fractions of A, E, N decomposition entering W \\
\rowcolor{allcol}
& $p_{AE},\, p_{WE},\, p_{NE}$ & & — & Fractions of A, W, N decomposition entering E \\
\rowcolor{allcol}
& $p_{AN},\, p_{WN},\, p_{EN}$ & & — & Fractions of A, W, E decomposition entering N \\
\rowcolor{allcol}
& $p_H$ & 0.04 / 0.0042$^*$ & — & Fraction of AWEN decomposition entering H \\
\midrule
\multicolumn{5}{l}{\textit{Temperature response (AWE pools) — shared by Y07, Y15, Y20}} \\
\rowcolor{allcol}
& $\beta_1$ & 0.096 & °C$^{-1}$ & Linear temperature coefficient \\
\rowcolor{allcol}
& $\beta_2$ & $-$0.0014 & °C$^{-2}$ & Quadratic temperature coefficient \\
\multicolumn{5}{l}{\textit{Temperature response (N pool) — Y15 and Y20 only}} \\
\rowcolor{y15col}
& $\beta_{N1}$ & & °C$^{-1}$ & Linear temperature coefficient, N pool \\
\rowcolor{y15col}
& $\beta_{N2}$ & & °C$^{-2}$ & Quadratic temperature coefficient, N pool \\
\multicolumn{5}{l}{\textit{Temperature response (H pool) — Y15 and Y20 only}} \\
\rowcolor{y15col}
& $\beta_{H1}$ & & °C$^{-1}$ & Linear temperature coefficient, H pool \\
\rowcolor{y15col}
& $\beta_{H2}$ & & °C$^{-2}$ & Quadratic temperature coefficient, H pool \\
\midrule
\multicolumn{5}{l}{\textit{Precipitation response (AWE pools) — shared by Y07, Y15, Y20}} \\
\rowcolor{allcol}
& $\gamma$ & $-$1.21 & m$^{-1}$ & Shape of precipitation response, $1-e^{\gamma P/1000}$ \\
\multicolumn{5}{l}{\textit{Precipitation response (N, H pools) — Y15 and Y20 only}} \\
\rowcolor{y15col}
& $\gamma_N$ & & m$^{-1}$ & Precipitation response shape, N pool \\
\rowcolor{y15col}
& $\gamma_H$ & & m$^{-1}$ & Precipitation response shape, H pool \\
\midrule
\multicolumn{5}{l}{\textit{Leaching — Y15 and Y20 only}} \\
\rowcolor{y15col}
& $w$ & & mm$^{-1}$ & Precipitation-driven leaching rate for AWEN pools \\
\midrule
\multicolumn{5}{l}{\textit{Size modifier — Y07 and Y15}} \\
\rowcolor{allcol}
& $\delta_1,\,\delta_2$ & $-$1.7, 0.86 & — & Exponent parameters, $\exp(-\delta_1 d^{\delta_2}|r|)$ \\
\rowcolor{allcol}
& $r$ & $-$0.306 & — & Scaling parameter (shared with Y20) \\
\multicolumn{5}{l}{\textit{Size modifier — Y20 only (replaces $\delta_1, \delta_2$)}} \\
\rowcolor{y20col}
& $\theta_1,\,\theta_2$ & & cm$^{-1}$, cm$^{-2}$ & Quadratic size modifier, $(1+\theta_1 d+\theta_2 d^2)^{-|r|}$ \\
\bottomrule
\multicolumn{5}{l}{$^*$ $p_H = 0.04$ for Y07/Y15; $p_H = 0.0042$ for Y20 — a 10-fold reduction.} \\
\end{tabular}
\end{table}
\endgroup


## 5.3 RothC parameter table

RothC's parameter vector is deliberately leaner than the Yasso family, reflecting
its simpler pool structure (no lateral AWEN transfers, no woody litter size effect).
The 10 proposed calibration parameters are listed below, with their functional
analog in the Yasso family where one exists.

\begin{table}[H]
\centering
\caption{RothC calibration parameters. The rightmost column identifies the
closest functional analog in the Yasso family where applicable.}
\label{tab:rothc_params}
\begin{tabular}{lllp{4.2cm}p{3.5cm}}
\toprule
\textbf{Group} & \textbf{Symbol} & \textbf{Default} & \textbf{Description} & \textbf{Yasso analog} \\
\midrule
\multicolumn{5}{l}{\textit{Decomposition rates}} \\
& $k_{\text{DPM}}$ & 10.0 & DPM decomp.\ rate (yr$^{-1}$) & $\alpha_A$ or $\alpha_W$ \\
& $k_{\text{RPM}}$ & 0.30 & RPM decomp.\ rate (yr$^{-1}$) & $\alpha_N$ \\
& $k_{\text{Bio}}$ & 0.66 & BIO decomp.\ rate (yr$^{-1}$) & $\alpha_E$ \\
& $k_{\text{Hum}}$ & 0.02 & HUM decomp.\ rate (yr$^{-1}$) & $\alpha_H$ \\
\midrule
\multicolumn{5}{l}{\textit{Temperature response}} \\
& $a_T$ & 47.91 & Numerator of Arrhenius modifier & $\beta_1, \beta_2$ (different form) \\
& $b_T$ & 106.06 & Activation energy proxy & $\beta_1, \beta_2$ (different form) \\
& $c_T$ & 18.27 & Temperature offset (°C) & $\beta_1, \beta_2$ (different form) \\
\midrule
\multicolumn{5}{l}{\textit{Moisture response}} \\
& $r_{\text{SMD}}$ & 0.444 & Breakpoint of piecewise linear TSMD response, as fraction of TSMD$_{\max}$ & $\gamma$ (different driver) \\
\midrule
\multicolumn{5}{l}{\textit{Inter-pool transfers and humification}} \\
& $f_{\text{bio}}$ & 0.46 & Fraction of decomposed C entering BIO & $p_H$, $p_{ij}$ (combined role) \\
& $f_{\text{hum}}$ & 0.54 & Fraction of decomposed C entering HUM & $p_H$, $p_{ij}$ (combined role) \\
\midrule
\multicolumn{5}{l}{\textit{Hardcoded — not calibrated}} \\
& evap fraction & 0.75 & Coefficient on open-pan evaporation in TSMD accumulation & --- \\
& $r_{\text{bare}}$ & 0.556 & Bare soil TSMD limit as fraction of TSMD$_{\max}$ & --- \\
& $\xi_{\text{min}}$ & 0.20 & Floor on moisture modifier under dry conditions & --- \\
& clay coefficients & & Pedotransfer function for TSMD capacity & --- \\
\bottomrule
\end{tabular}
\end{table}

\noindent\textbf{Note on $f_{\text{bio}}$ and $f_{\text{hum}}$.}
In the standard RothC formulation these fractions are derived from clay content
via $x = 1.67(1.85 + 1.60\,e^{-0.0786\,\%\text{clay}})$, with
$f_{\text{bio}} = 0.46/(x+1)$ and $f_{\text{hum}} = 0.54/(x+1)$.
By freeing the numerators $0.46$ and $0.54$ as direct calibration parameters
(while retaining the clay-dependent denominator), we preserve the soil-texture
sensitivity of the original formulation while giving the Bayesian sampler
a handle on the bio--hum split. This is the closest structural analog to Yasso's
$p_H$ combined with the inter-pool transfer fractions.













\newpage

# 6. Calibration Engine Architecture

## 6.1 Design Principle

The Bayesian calibration framework is split into two layers: a **model-agnostic engine** shared by all models, and a thin **model-specific interface** that each model script implements. The engine handles everything that does not depend on soil carbon dynamics; the interface provides only the three functions that do.

This design means that changes to convergence diagnostics, output management, logging, or MCMC settings are made once and propagate to all five model calibrations automatically. It also guarantees that all models are evaluated under identical statistical assumptions, which is a prerequisite for valid structural uncertainty quantification.

## 6.2 Engine Components

The following components are fully model-agnostic and implemented once in \texttt{calibration\_engine.R}:

\medskip
\noindent\textbf{Transform builder.} Constructs \texttt{to\_original()}, \texttt{to\_unconstrained()}, and the log-Jacobian function automatically from the model's \texttt{param\_spec} declaration. See Section 6.4 for details.

\medskip
\noindent\textbf{Prior predictive matching.} The procedure for choosing the prior width $\sigma_{\text{ppm}}$ is identical across models: draw samples from a Gaussian centred on \texttt{best\_x}, transform to physical space, run one plot forward, compute the coefficient of variation of predicted SOC, and find the smallest $\sigma$ whose prior predictive CV meets or exceeds the observed CV. The engine requires only a single-plot forward run function from the model interface. On synthetic data with few unique plots the matching may not converge; the engine warns and defaults to $\sigma_{\text{ppm}} = 1.0$.

\medskip
\noindent\textbf{Likelihood factory (\texttt{make\_likelihood()}).} Returns a byte-compiled likelihood function with all model callables and pre-computed data structures captured in its closure. Receives the three model interface functions, the pre-split data structures, and the transform functions. The returned function is passed to \texttt{run\_mcmc\_chains()}.

\medskip
\noindent\textbf{MCMC orchestration (\texttt{run\_mcmc\_chains()}).} Chain initialisation, the DEzs sampler loop, per-chain log files, the evaluation counter, and elapsed-time tracking. Receives the likelihood function, prior object, and both \texttt{mcmc\_settings} and \texttt{run\_config}. The outer chain loop (\texttt{lapply} on Mac, \texttt{mclapply} on Roihu) and the inner parallel plot loop (\texttt{mclapply} with \texttt{CORES\_PER\_CHAIN}) are both managed here.

\medskip
\noindent\textbf{Convergence diagnostics (\texttt{run\_diagnostics()}).} Gelman--Rubin $\hat{R}$ via \texttt{gelman.diag}, effective sample size via \texttt{effectiveSize}, posterior quantile tables, trace PDFs, and prior-vs-posterior marginal PDFs. Receives \texttt{mcmc\_settings} as a separate argument (not via \texttt{run\_config}) to access iteration counts. All diagnostics operate on raw MCMC samples and parameter names; they have no knowledge of what model produced them.

\medskip
\noindent\textbf{Output management (\texttt{save\_results()}).} Serialises three files to \texttt{DIR\_RUNS}: the raw BayesianTools chain objects (\texttt{MODEL\_chains\_RUNID.rds}), the physical-space posterior samples matrix \texttt{all\_phys} received from \texttt{run\_diagnostics()} (\texttt{MODEL\_posterior\_RUNID.rds}), and run metadata including convergence summaries (\texttt{MODEL\_metadata\_RUNID.rds}). Receives \texttt{mcmc\_settings} for iteration counts.

\medskip
\noindent\textbf{Input serialisation (\texttt{save\_inputs()}).} Serialises all pre-processed data structures that entered the likelihood to \texttt{DIR\_INPUTS} as \texttt{MODEL\_inputs\_RUNID.rds}. Called after prior predictive matching so that $\sigma_{\text{ppm}}$ is included in the saved object alongside \texttt{climate\_by\_plot}, \texttt{inputs\_by\_plot}, \texttt{litter\_means}, \texttt{obs\_meta}, \texttt{plots}, \texttt{plots\_real}, \texttt{free\_defaults}, and \texttt{best\_x}. Serves two purposes: enabling post-hoc analysis (posterior predictive checks, residual diagnostics) without re-sourcing the calibration script; and providing a validation record that the input pipeline produced the expected structures, independently verifiable against the raw data.

\medskip
\noindent\textbf{Startup checks.} Mandatory round-trip validation (physical $\to$ unconstrained $\to$ physical to machine precision) and likelihood finiteness check at default parameters, executed before sampling begins.

\medskip
\noindent\textbf{Configuration structure.} \texttt{run\_config} (output directories, model name, run ID, cores) and \texttt{mcmc\_settings} (chain count, iteration counts, burn-in, log interval) are both assembled in Section~0 of each model script at startup, before any data loading or computation. The only field added later is \texttt{run\_config\$n\_plots}, appended at the end of Section~3 once the plots vector is known. Both objects are passed explicitly to every engine function that needs them.

## 6.3 Model Interface

Each model script passes three functions to the engine as arguments to \texttt{make\_likelihood()}. These are not new functions written for calibration --- they are the same R wrapper functions already implemented for forward model runs, validated against the reference Fortran implementations, and used independently for testing and simulation.

\begin{table}[H]
\centering
\caption{The three functions constituting the model--engine interface.}
\label{tab:interface}
\begin{tabular}{p{3.5cm}p{5cm}p{5cm}}
\toprule
\textbf{Function} & \textbf{Signature} & \textbf{Role} \\
\midrule
\texttt{compute\_xi(clim, params)} &
  Climate data frame + parameter vector &
  Returns environmental modifier time series $\boldsymbol{\xi}(t)$; called once per plot per likelihood evaluation. Must live inside the sampler because climate sensitivity parameters ($\beta_1, \beta_2, \gamma$) are free calibration parameters. \\
\addlinespace
\texttt{steady\_state(params, litter\_means, xi\_mean)} &
  Parameter vector + mean litter inputs + mean $\xi$ &
  Returns initial pool vector $\mathbf{C}^*$; called once per plot per likelihood evaluation. \\
\addlinespace
\texttt{run\_model(inputs, params, C\_init, xi)} &
  Input data frame + parameters + $\mathbf{C}^*$ + $\boldsymbol{\xi}(t)$ &
  Runs full transient simulation; returns SOC trajectory; single Fortran call per plot. \\
\bottomrule
\end{tabular}
\end{table}

For example, the Yasso07 model script passes \texttt{compute\_xi\_yasso07},
\texttt{yasso07\_steady\_state}, and \texttt{yasso07\_run} to \texttt{make\_likelihood()},
which captures them in the likelihood closure. The same functions are used for
standalone forward runs and validation against the reference implementation.

```r
# Forward run -- direct function call, fixed parameters
out <- yasso07_run(inputs_df,
                   params   = YASSO07_DEFAULT_PARAMS,
                   C_init   = C_init,
                   xi_array = xi_array)

# Calibration -- same functions injected into likelihood factory,
# which is then passed to the MCMC engine
ll_fn <- make_likelihood(
  compute_xi      = compute_xi_yasso07,   # same function as forward run
  steady_state    = yasso07_steady_state, # same function as forward run
  run_model       = yasso07_run,          # same function as forward run
  to_original     = to_original,
  log_jacobian    = log_jacobian,
  assemble_params = assemble_model_params,
  plots           = plots,
  climate_by_plot = climate_by_plot,
  inputs_by_plot  = inputs_by_plot,
  litter_means    = litter_means,
  obs_meta        = obs_meta,
  n_cores         = CORES_PER_CHAIN
)

chain_results <- run_mcmc_chains(
  ll_fn         = ll_fn,
  prior         = prior,
  best_x        = best_x,
  param_names   = FREE_NAMES,
  mcmc_settings = mcmc_settings,
  run_config    = run_config
)
```

The only requirement the engine imposes on the three interface functions is
that they respect the calling conventions listed in the table above. Any
function that runs correctly in forward mode already satisfies this requirement.
There is no translation layer, no reformatting of inputs or outputs, and no
risk of the calibration exercising a different code path than the validated
forward run.

## 6.4 Parameter Transformation Framework

Transformations between constrained physical space and unconstrained sampling
space are generated automatically from a \textbf{parameter specification table}
defined in each model script. Rather than writing model-specific
\texttt{to\_original} and \texttt{to\_unconstrained} functions, the engine
constructs these from a list of typed parameter groups.

Three primitive transform types cover all parameters across all four models:

\begin{table}[H]
\centering
\caption{Primitive transformation types and their application across models.}
\label{tab:transforms}
\begin{tabular}{llp{4cm}p{5cm}}
\toprule
\textbf{Type} & \textbf{Constraint} & \textbf{Transform} & \textbf{Applied to} \\
\midrule
\texttt{stick\_break} & $\sum p_i \leq B$, $p_i \geq 0$ & Sequential logit fractions of remaining budget & Yasso AWEN inter-pool transfer fractions ($p_{ij}$, 12 per model) \\
\addlinespace
\texttt{log} & $\theta > 0$ & $x = \log\theta$ & Variance and scaling terms ($\sigma_{\text{obs}}$, $\sigma_{\text{init}}$, $\sigma_{\text{input}}$) \\
\addlinespace
\texttt{unconstrained} & $\theta \in \mathbb{R}$ & Identity & Temperature and precipitation response parameters ($\beta_i$, $\gamma$); size modifier parameters ($\delta_i$, $r$) \\
\addlinespace
\texttt{logit} & $\theta \in (0,1)$ & $x = \log\frac{\theta}{1-\theta}$ & Single proportions; RothC $f_{\text{bio}}$, $f_{\text{hum}}$ \\
\bottomrule
\end{tabular}
\end{table}

\noindent\textbf{Fixed vs.\ free parameters.}
Decomposition rate constants ($\alpha_{A,W,E,N,H}$ for Yasso; $k_{\text{DPM,RPM,Bio,Hum}}$ for RothC) and the humification fraction $p_H$ are held fixed at published values throughout MCMC. These encode each model's pool chemical identity and are independently constrained by experimental data; calibrating them jointly with transfer fractions is not identifiable and would blur the structural differences the intercomparison is designed to measure. The free parameters --- transfer fractions, climate sensitivity, size modifiers, and variance and scaling terms --- govern climate response, partitioning, and input uncertainty, the quantities most relevant to Finnish climate projections.

\noindent The Yasso07 parameter specification takes the form:

```r
# Budget for stick-breaking: fraction not going to humus
B <- 1.0 - fixed_rates["p_H"]

param_spec <- list(
  # Inter-pool transfer fractions -- one stick-breaking group per source pool
  list(names = c("p_AW","p_AE","p_AN"), type = "stick_break", budget = B),
  list(names = c("p_WA","p_WE","p_WN"), type = "stick_break", budget = B),
  list(names = c("p_EA","p_EW","p_EN"), type = "stick_break", budget = B),
  list(names = c("p_NA","p_NW","p_NE"), type = "stick_break", budget = B),
  # Climate sensitivity -- unconstrained (can be negative)
  list(names = c("beta1","beta2","gamma"),              type = "unconstrained"),
  # Size modifiers -- unconstrained
  list(names = c("delta1","delta2","r"),                type = "unconstrained"),
  # Variance and scaling terms -- strictly positive
  list(names = c("sigma_obs","sigma_init","sigma_input"), type = "log")
)
```

The engine builds \texttt{to\_original}, \texttt{to\_unconstrained}, and the Jacobian
from this spec. The Jacobian is assembled by summing contributions per group:
stick-breaking groups contribute the sequential logit volume correction;
log-transformed parameters contribute $\log\theta$ each; unconstrained parameters
contribute zero.

\noindent\textbf{Test mode replication.}
The \texttt{N\_PLOTS\_TEST} parameter replicates plot IDs in the \texttt{plots} vector
to simulate production wallclock cost without requiring the full dataset:

```r
# plots is a vector of plot IDs passed to the likelihood -- either the full
# production set or N_PLOTS_TEST replications of plots_real for timing tests.
plots <- if (!is.na(N_PLOTS_TEST)) {
  rep(plots_real, length.out = N_PLOTS_TEST)
} else {
  plots_real
}
```

The replication is a lightweight index trick: \texttt{climate\_by\_plot} and
\texttt{inputs\_by\_plot} retain only the unique real plots; the likelihood
evaluates the same plots repeatedly. This gives a realistic wallclock estimate
for production scaling but provides no new statistical information.
\textbf{Convergence diagnostics from replicated test runs are meaningless and should be ignored} --- the sampler sees repeated identical data and cannot distinguish structural from sampling uncertainty.

## 6.5 Known Validation Items

The following issues were identified during implementation and should be verified before the production run on real data:

\medskip
\noindent\textbf{$\xi_{\text{mean}}$ timing.}
The analytical steady-state uses \texttt{mean(xi\_array)} computed over the full simulation time series, including years after the last observation. For a plot measured in 1997--2006, this includes post-2006 climate. Strictly, the steady-state should use pre-measurement climate only. This inconsistency is absorbed into \texttt{sigma\_init} and may not be detectable, but introduces a systematic bias in plots with strong climate trends.

\medskip
\noindent\textbf{Jacobian passthrough in BayesianTools.}
The log Jacobian is added inside the likelihood rather than the prior, which keeps the BayesianTools prior object clean. This assumes \texttt{createPrior()} passes the custom density function through unchanged and does not apply additional normalisation. This assumption should be verified against the BayesianTools source for the version in use.

\medskip
\noindent\textbf{Year alignment between Fortran output and \texttt{obs\_meta\$idx}.}
The observation year indices are computed as \texttt{match(obs\_plot\$year, clim\$year)}, which assumes the Fortran model returns one row per year in the same order as the climate input data frame. If the Fortran wrapper reorders or drops years, \texttt{meta\$idx} points to wrong rows without triggering any guard. This should be validated by running a forward model call and checking that \texttt{run\_out\$total\_soc[meta\$idx]} corresponds to the correct years.

## 6.6 Prior Specification

All free parameters share a common Gaussian prior in unconstrained sampling space:

$$x_i \sim \mathcal{N}\!\big(\texttt{best\_x}_i,\; \sigma_{\text{ppm}}^2\big) \qquad \forall\, i$$

where $\texttt{best\_x}_i$ is the published default value of parameter $i$ transformed to unconstrained space, and $\sigma_{\text{ppm}}$ is a single scalar width shared across all parameters. The value of $\sigma_{\text{ppm}}$ is chosen separately for each model by prior predictive matching: it is set to the smallest value such that the coefficient of variation of SOC predictions drawn from the prior matches the observed CV across plots. This normalises prior coverage in prediction space rather than parameter space, ensuring that structural differences in posterior predictions reflect model structure rather than differences in prior informativeness. A model requiring a wider $\sigma_{\text{ppm}}$ to reach the observed CV is collectively less sensitive to its free parameters in SOC space --- itself a structural finding. Fixed parameters --- decomposition rate constants and humification fraction $p_H$ --- are held at published values and do not enter the prior.

\begin{table}[H]
\centering
\caption{Prior specification for all calibrated parameters. All parameters share
the same Gaussian prior form in unconstrained space; the transform determines the
implied prior shape in physical space. $\sigma_{\text{ppm}}$ is set by prior
predictive matching and differs across models.}
\label{tab:priors}
\begin{tabular}{p{3.8cm}p{4.2cm}lp{3.8cm}}
\toprule
\textbf{Parameter class} & \textbf{Parameters} & \textbf{Transform} & \textbf{Prior (unconstrained)} \\
\midrule
Inter-pool transfer fractions
  & $p_{ij}$ (12, Yasso); $f_{\text{bio}},\,f_{\text{hum}}$ (RothC)
  & stick-breaking / logit
  & $\mathcal{N}(\texttt{best\_x},\,\sigma_{\text{ppm}}^2)$ \\
\addlinespace
Climate sensitivity
  & $\beta_1,\,\beta_2,\,\gamma$ (Yasso); $a_T,\,b_T,\,c_T,\,r_{\text{SMD}}$ (RothC)
  & unconstrained
  & $\mathcal{N}(\texttt{best\_x},\,\sigma_{\text{ppm}}^2)$ \\
\addlinespace
Size modifiers
  & $\delta_1,\,\delta_2,\,r$ (Yasso only)
  & unconstrained
  & $\mathcal{N}(\texttt{best\_x},\,\sigma_{\text{ppm}}^2)$ \\
\addlinespace
Variance \& scaling terms
  & $\sigma_{\text{obs}},\,\sigma_{\text{init}},\,\sigma_{\text{input}}$
  & log
  & $\mathcal{N}(\texttt{best\_x},\,\sigma_{\text{ppm}}^2)$ \\
\midrule
\textit{Fixed — not calibrated}
  & $\alpha_{A,W,E,N,H}$;\; $k_{\text{DPM,RPM,Bio,Hum}}$;\; $p_H$
  & —
  & fixed at published values \\
\bottomrule
\end{tabular}
\end{table}

\noindent\textbf{Variance and scaling terms.}
Three nuisance parameters are jointly calibrated with the structural parameters.
$\sigma_{\text{obs}}$ is the standard deviation of residuals between predicted and
observed SOC at measurement time, capturing measurement error, within-plot spatial
heterogeneity, and any model structural error that presents as noise.
$\sigma_{\text{init}}$ captures uncertainty in the analytical steady-state
initialisation: real plots are not at equilibrium, and the mean climate used in
$\mathbf{C}^* = -\mathbf{A}^{-1}\mathbf{b}$ is itself an estimate.
$\sigma_{\text{input}}$ is a global multiplicative scaling factor on all litter inputs,
$\mathbf{J}_{\text{scaled}}(t) = \sigma_{\text{input}} \cdot \mathbf{J}(t)$, which
accounts for systematic bias in the allometric-equation-derived litter estimates.
A posterior value below 1 indicates the allometric equations overestimate inputs for
Finnish forests; above 1 indicates underestimation. Because decomposition rates are
fixed, $\sigma_{\text{input}}$ is not strongly confounded with rates and is expected
to be identifiable from the SOC data.

## 6.7 Architecture Overview

\begin{figure}[H]
\centering
\begin{tikzpicture}[
    >=Stealth,
    node distance=0.9cm and 1.8cm,
    engine/.style={
        rectangle, rounded corners=4pt, draw=outputpurple, fill=outputpurple!8,
        line width=1pt, minimum height=0.9cm, minimum width=4.2cm,
        text centered, align=center, font=\small\sffamily
    },
    model/.style={
        rectangle, rounded corners=4pt, draw=decompred, fill=decompred!8,
        line width=1pt, minimum height=0.9cm, minimum width=3.6cm,
        text centered, align=center, font=\small\sffamily
    },
    data/.style={
        rectangle, rounded corners=2pt, draw=bordergray, fill=bglight,
        line width=0.6pt, minimum height=0.7cm, minimum width=2.8cm,
        text centered, font=\footnotesize\ttfamily
    },
    arr/.style={->, line width=1pt, color=#1},
    note/.style={font=\scriptsize\sffamily\itshape, text=black!55, align=center},
    hdr/.style={font=\small\sffamily\bfseries, text=#1}
]

% =============================================
% MODEL SCRIPT (left column)
% =============================================
\node[hdr=decompred] (ms_hdr) at (-4.5, 5.2)
    {Model script (one per model)};

\node[model] (pspec)    at (-4.5,  4.2)  {\texttt{param\_spec}};
\node[model] (defaults) at (-4.5,  3.0)  {Default parameters};
\node[model] (cxi)      at (-4.5,  1.8)  {\texttt{compute\_xi()}};
\node[model] (css)      at (-4.5,  0.6)  {\texttt{steady\_state()}};
\node[model] (crun)     at (-4.5, -0.6)  {\texttt{run\_model()}};
\node[model] (dataprep) at (-4.5, -2.0)
    {Data preparation \\
     \texttt{map\_climate\_*()}, \texttt{map\_inputs\_*()} \\
     \texttt{climate\_by\_plot}, \texttt{obs\_meta}};

% =============================================
% SECTION 0 CONFIG (inside model script region)
% =============================================
\node[data] (runcfg) at (-2.0,  4.6)  {\texttt{run\_config}};
\node[data] (mset)   at (-2.0,  3.6)  {\texttt{mcmc\_settings}};

% =============================================
% ENGINE (right column)
% =============================================
\node[hdr=outputpurple] (eng_hdr) at (3.0, 5.2)
    {calibration\_engine.R};

\node[engine] (trans) at (3.0, 4.2)
    {Transform builder \\
     \texttt{to\_original()}, \texttt{to\_unconstrained()} \\ Jacobian};

\node[engine] (ppm) at (3.0, 2.4)
    {Prior predictive \\ matching};

\node[engine] (ll) at (3.0, 1.0)
    {Likelihood factory \\ \texttt{make\_likelihood()}};

\node[engine] (mcmc) at (3.0, -0.4)
    {MCMC orchestration \\ \texttt{run\_mcmc\_chains()}};

\node[engine] (diag) at (3.0, -1.7)
    {Diagnostics \\ \texttt{run\_diagnostics()}};

\node[engine] (save) at (3.0, -3.0)
    {Output \\ \texttt{save\_results()}, \texttt{save\_inputs()}};

% =============================================
% ARROWS — param_spec and defaults to engine
% =============================================
\draw[arr=decompred] (pspec.east) -- (trans.west)
    node[midway, above=2pt, note] {type, budget};

\draw[arr=decompred] (defaults.east) -| (0.2, 2.4) -- (ppm.west)
    node[near end, above=2pt, note] {\texttt{best\_x}};

% =============================================
% ARROWS — three interface functions to likelihood
% =============================================
\draw[arr=decompred] (cxi.east)  -- (ll.west);
\draw[arr=decompred] (css.east)  -- (ll.west);
\draw[arr=decompred] (crun.east) -- (ll.west);
\node[note] at (-1.2, 0.9) {interface\\functions};

% =============================================
% ARROWS — data prep to likelihood, ppm, save
% =============================================
\draw[arr=decompred] (dataprep.east) -- ++(1.8, 0) |- (ll.south west)
    node[near start, above=2pt, note] {pre-split data};
\draw[arr=decompred] (dataprep.east) -- ++(1.8, 0) |- (ppm.south west);
\draw[arr=decompred] (dataprep.east) -- ++(1.8, 0) |- (save.south west)
    node[near start, below=2pt, note] {inputs snapshot};

% =============================================
% ARROWS — Section 0 config to engine functions
% =============================================
\draw[arr=bordergray] (runcfg.east) -- ++(0.4, 0) |- (mcmc.west);
\draw[arr=bordergray] (runcfg.east) -- ++(0.4, 0) |- (diag.west);
\draw[arr=bordergray] (runcfg.east) -- ++(0.4, 0) |- (save.west);
\draw[arr=bordergray] (mset.east)   -- ++(0.4, 0) |- (mcmc.west);
\draw[arr=bordergray] (mset.east)   -- ++(0.4, 0) |- (diag.west);

% =============================================
% ARROWS — within engine (top to bottom)
% =============================================
\draw[arr=outputpurple] (trans.south) -- (ppm.north);
\draw[arr=outputpurple] (ppm.south) -- (ll.north)
    node[midway, right=3pt, note] {$\sigma_{\text{ppm}}$};
\draw[arr=outputpurple] (ll.south) -- (mcmc.north)
    node[midway, right=3pt, note] {likelihood fn};
\draw[arr=outputpurple] (mcmc.south) -- (diag.north)
    node[midway, right=3pt, note] {chain results};
\draw[arr=outputpurple] (diag.south) -- (save.north)
    node[midway, right=3pt, note] {\texttt{all\_phys}};

% =============================================
% BACKGROUND BOXES — two regions only
% =============================================
\begin{scope}[on background layer]
    \node[
        rectangle, rounded corners=8pt,
        draw=decompred!40, fill=decompred!3,
        line width=1.2pt, dashed,
        fit=(pspec)(defaults)(cxi)(css)(crun)(dataprep)(runcfg)(mset),
        inner sep=12pt] {};
    \node[
        rectangle, rounded corners=8pt,
        draw=outputpurple!40, fill=outputpurple!3,
        line width=1.2pt, dashed,
        fit=(trans)(ppm)(ll)(mcmc)(diag)(save),
        inner sep=12pt] {};
\end{scope}

\end{tikzpicture}
\caption{%
High-level structure of the calibration framework.
The \textbf{model script} (red) contains everything model-specific:
a typed parameter specification, published default parameters,
the three interface functions (\texttt{compute\_xi}, \texttt{steady\_state},
\texttt{run\_model}), model-specific data preparation
(input mapping functions, pre-splitting into per-plot lookup structures),
and Section~0 configuration (\texttt{run\_config}, \texttt{mcmc\_settings}).
The \textbf{engine} (\texttt{calibration\_engine.R}, purple) is fully
model-agnostic: it constructs transformation functions and the Jacobian
from the parameter specification, runs prior predictive matching to set
$\sigma_{\text{ppm}}$, assembles the generic likelihood, orchestrates
MCMC sampling, computes diagnostics, and serialises outputs.
All data structures produced by the model script's preparation step
(\texttt{climate\_by\_plot}, \texttt{obs\_meta}, etc.) are passed into
the engine via \texttt{make\_likelihood()}'s closure arguments ---
they are not a separate architectural layer.}
\label{fig:engine}
\end{figure}